---
title: "Using Code Examples"
---

## How to read the examples

Most chapters include runnable R code interleaved with the explanation. Code
blocks show the call, and the output shown below each block is the actual result
of running it against the book's example data. Where a block builds on an
earlier one, the objects it depends on were created earlier in the same chapter.

The book favors clarity over brevity: intermediate objects are usually named and
inspected rather than chained into a single unbroken pipeline, so that each step
in an analysis is visible and can be checked.

## What the examples assume

The examples use only two families of packages:

- [`tidycreel`](https://chrischizinski.github.io/tidycreel/) — the companion
  package that provides the creel-specific design objects, estimators,
  diagnostics, and plotting helpers.
- The [tidyverse](https://www.tidyverse.org/) — `dplyr`, `tidyr`, `ggplot2`,
  `purrr`, `stringr`, `forcats`, `lubridate`, and `here` for general data
  manipulation and graphics.

A typical chapter begins by loading these:

```{r}
#| eval: false
library(tidycreel)
library(tidyverse)
```

No other packages are required to reproduce the analyses.

## Where the data comes from

The running examples use a small set of example datasets — most prominently the
Harlan Reservoir survey — distributed with the book in `data/` or bundled with
`tidycreel`. The datasets, their structure, and the assumptions behind them are
described in the [Data Notes](appendices/02-data-notes.qmd) appendix. Notation
used across chapters is collected in [Notation and Symbols](appendices/01-notation.qmd).

## Which versions produced these numbers

Every estimate printed in this book was produced by the versions below, at the
time the book was last rendered:

```{r}
#| label: book-versions
#| echo: false
cat("tidycreel:", as.character(packageVersion("tidycreel")), "\n")
cat("R:        ", R.version$version.string, "\n")
```

These are read from the rendering session rather than written down, so they
cannot fall out of step with the numbers on the page.

::: {.callout-important appearance="simple" title="Estimates are version-sensitive"}
Reproducing a chapter under a different version of `tidycreel` may give
different **standard errors** for the same data and the same point estimates.
This is expected, and it is not a sign that something has gone wrong.

The package is under active development in exactly the area the book reports:
how uncertainty is propagated from the quantities a survey measures into the
quantities it reports. Two recent examples, both of which moved published
numbers in this book:

- **3.2.0** began carrying the party-size estimate's own standard error into
  the effort standard error. Harlan's effort SE moved from 3,057 to 3,182
  angler-hours. The effort estimate itself did not change.
- **3.3.0** stopped combining a party-size estimate *shared across strata* as
  though the strata were independent, which raised the standard errors on
  total catch, harvest, and release. The totals themselves did not change.

In both cases the newer number is the correct one and the older number was
understated. If your standard errors differ from the book's, check your package
version before looking for an error in your data.

**3.4.0** made the same correction for a party-size estimate shared across
*sections*, and changed the mixed case — an estimate that straddles the
sections unevenly — from a quietly understated number to `NA`. No number
previously published in this book moved, because until @sec-sections-purpose
was added the book used no sectioned designs. The behaviour is worth knowing
about anyway: it is the one place where asking for a lake-wide standard error
can correctly return nothing at all.

**4.0.0** is the first release to move a *point* estimate in this book, and it
moved one a long way. An aerial count is a raw observer count, and it needs two
corrections that had no home in the package: a visibility correction for the
anglers the observer missed, and an angler-to-people ratio for the people who
were seen but were not fishing. Neither was applied, and neither could be
stated. @sec-aerial-design now supplies both from @smucker2010, and the aerial
effort estimate moved from 379 to 673 angler-hours while its 95% confidence
interval widened from 314–443 to 367–979.

Both parts of that are the point. The estimate moved because two corrections
that belong in the arithmetic were missing from it. The interval widened
because those corrections are themselves *estimated* — from paired air-ground
counts, with standard errors the source paper reports — and a correction
treated as a known constant contributes no uncertainty at all. The old interval
was not more precise; it was measuring less than it claimed to.

Point estimates are otherwise far more stable than standard errors across
versions. When a release does change a point estimate, it is called out as a
breaking change in [the package's NEWS file](https://github.com/chrischizinski/tidycreel/blob/main/NEWS.md).
:::

## Running the code yourself

To run the examples locally, install `tidycreel` and the tidyverse, then work
through a chapter top to bottom. Because later blocks depend on earlier ones,
run a chapter's code in order rather than in isolation. Details on how the book
stays reproducible are in the [Reproducibility](appendices/03-reproducibility.qmd)
appendix.

To install the exact version this book was rendered against rather than the
current development state:

```{r}
#| label: install-pin
#| echo: false
#| results: asis
pinned <- as.character(packageVersion("tidycreel"))
cat(
  "``` r\n",
  "remotes::install_github(\"chrischizinski/tidycreel@v", pinned, "\")\n",
  "```\n",
  sep = ""
)
```

That version is read from the rendering session for the same reason the table
above is, so it always names whatever produced the numbers on this page.

One caveat follows from that. Between releases the book may be built against a
development version whose tag does not exist yet — the version number is
claimed in `DESCRIPTION` as soon as work on it starts, while the tag is only
cut when it ships. If the command above fails because the tag is not found,
that is what has happened, and the fix is to install from the commit the book
was built from rather than from the tag:

```{r}
#| eval: false
# Substitute the commit recorded in the Reproducibility appendix.
remotes::install_github("chrischizinski/tidycreel@<commit>")
```
