---
title: "Aerial, Camera-Assisted, and Remote Count Designs"
---

```{r}
#| label: ch06-setup
#| include: false
library(tidycreel)
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
```

## Why remote counts? {#sec-remote-why}

Standard creel survey designs — access-point (@sec-access-point) and roving
(@sec-roving) — assume a field crew can stand at an access point or travel a
route and count anglers directly. That assumption fails when the waterbody is
too large for crew coverage, when weather prevents reliable manual counts, when
budget limits staffing to only a few days per week, or when the survey requires
higher temporal resolution than a crew can provide.

Remote and technology-assisted count methods solve the coverage problem by
replacing or supplementing the human observer with a sensor — a camera at a
boat ramp, an aircraft flying a systematic route, a vehicle counter at a road
crossing. Coverage goes up, field labor goes down, and the record becomes
persistent: a camera image can be reviewed; a flight log can be audited.

The estimation logic does not change. Effort is still the product of a count
of anglers and the hours those anglers were available to fish. What changes is
how the count is collected and how timing uncertainty is handled. An aerial
flight covers the entire waterbody in thirty minutes but captures only a
snapshot at the flight time. A boat-ramp camera records every ingress event
but produces a vehicle count, not a direct angler-hour measurement. Both
require a link between the count and total effort — through a model, through
a calibration ratio, or through mean trip length from interviews.

`tidycreel` provides dedicated design types and estimators for each major
remote-count approach: `design_type = "aerial"` with
`estimate_effort_aerial_glmm()` for time-of-day-adjusted aerial surveys, and
`design_type = "camera"` with `est_effort_camera()` for ingress/egress counter
designs. The same `creel_design()` → `add_counts()` → `add_interviews()` →
estimate pipeline applies to both, with a few additional required arguments
that capture the characteristics of the observation system.

The common thread is simple: the remote sensor changes how the count is
observed, but not what the analyst ultimately needs. The target is still
angler-hours, so each design has to translate the sensor output back into
that unit.

## Aerial surveys {#sec-aerial}

### The observation problem {#sec-aerial-obs}

An aerial survey produces one count per flight, and the count reflects the
number of anglers visible at a single moment in time. Angler density on a lake
changes continuously across the fishing day: low at first light, building
through midmorning as anglers arrive, peaking around midday or early afternoon,
then declining as anglers leave and daylight wanes. A single flight at 10:00
captures a different density than a flight at 14:00 on the same day.

The naive expansion — count × hours_open — assumes the observed density is
representative of the full day, which it is only if the flight happens exactly
at the time when mean daily density occurs. If flights are consistently timed
early or late relative to peak use, the effort estimate is biased in a
predictable direction.

The standard fix is a **time-of-day model** that predicts how angler density
varies across the fishing day. Multiple flights on the same days, spaced at
different times, provide the data to fit that model. The fitted curve is then
integrated across the full open-water period to recover total daily effort.
`estimate_effort_aerial_glmm()` fits this model as a generalized linear mixed
model (GLMM) with a quadratic time-of-day effect and a day-level random
intercept:

$$
\text{count}_{d,t} = \exp\bigl(\beta_0 + \beta_1 t + \beta_2 t^2 + u_d\bigr)
$$

where $t$ is flight time, $u_d \sim N(0, \sigma^2)$ is a day-level random
effect that absorbs day-to-day variation in total use, and the exponential
link keeps predictions non-negative. Integrating the fitted curve over the
$h_{\text{open}}$ open hours of the fishing day recovers a day-specific total
angler-hours estimate, which is then expanded to the full season using the
stratified ratio estimator.

### Data structure {#sec-aerial-data}

An aerial survey dataset contains one row per flight observation — not one row
per day. Multiple flights on the same day appear as separate rows, linked by
date and distinguished by `time_of_flight`. The `tidycreel` bundled dataset
`example_aerial_glmm_counts` illustrates this structure:

```{r}
#| label: aerial-data-structure
#| echo: true
example_aerial_glmm_counts |>
  arrange(date, time_of_flight) |>
  head(12)
```

Four flights per day at hours 7, 10, 13, and 16. The counts rise from early
morning to a midday peak, then fall — exactly the pattern the GLMM is designed
to capture. With four time points per day, the quadratic curve is well
constrained.

### Building the aerial design {#sec-aerial-design}

An aerial design requires three additions to the standard `creel_design()` call:

- `design_type = "aerial"` — routes count data to the GLMM estimator
- `h_open` — total fishing hours per day (daylight hours accessible to anglers)
- Raw, unaggregated counts passed to `add_counts()` without `count_time_col`,
  so `time_of_flight` is preserved in `design$counts` for the model

```{r}
#| label: aerial-design-build
#| echo: true
aerial_calendar <- tibble(
  date     = seq(min(example_aerial_glmm_counts$date),
                 max(example_aerial_glmm_counts$date), by = "day"),
  day_type = if_else(wday(date) %in% c(1L, 7L), "weekend", "weekday")
)

design_aerial <- creel_design(
  calendar    = aerial_calendar,
  date        = "date",
  strata      = "day_type",
  design_type = "aerial",
  h_open      = 14
) |>
  add_counts(example_aerial_glmm_counts) |>
  add_interviews(
    interviews    = example_aerial_interviews,
    catch         = "walleye_catch",
    harvest       = "walleye_kept",
    effort        = "hours_fished",
    trip_status   = "trip_status",
    trip_duration = "hours_fished"
  )
```

The contrast with a standard instantaneous design is in `design_type` and
`h_open`. Everything else — `add_interviews()`, estimation calls — follows
the same pattern. The GLMM handles the time-of-day adjustment internally;
the analyst provides the data and the open-water window.

### Time-of-day pattern {#sec-aerial-tod}

Before estimating effort, it is worth visualizing what the GLMM is fitting.
@fig-aerial-tod shows mean count by flight time across both day types, with
the quadratic pattern that drives the model.

```{r}
#| label: fig-aerial-tod
#| echo: true
#| fig-cap: "Mean aerial count by time of flight and day type. Points show stratum means across sampled days; the curve is a loess smoother that approximates the quadratic GLMM fit. Counts peak in early afternoon and are consistently higher on weekends, reflecting higher recreational use."
#| fig-alt: "Scatter plot with time of flight (hours) on the x axis from 7 to 16, mean angler count on the y axis. Two smooth curves, one per day type, both arching upward then downward with weekends higher than weekdays."
example_aerial_glmm_counts |>
  group_by(day_type, time_of_flight) |>
  summarise(mean_count = mean(n_anglers), .groups = "drop") |>
  ggplot(aes(time_of_flight, mean_count, colour = day_type, group = day_type)) +
  geom_point(size = 3) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
  scale_colour_manual(
    values = c(weekday = "#2171b5", weekend = "#e6550d"),
    labels = c(weekday = "Weekday", weekend = "Weekend"),
    name   = NULL
  ) +
  labs(
    x = "Time of flight (hour of day)",
    y = "Mean angler count"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

The inverted-U shape confirms that a single flight at 07:00 would
underestimate total daily effort, while a flight at 13:00 would overestimate
it. The GLMM integrates the full curve rather than multiplying one snapshot
count by hours_open.

That is the reason the model is doing real work here. It is not just smoothing
noise; it is replacing a one-time observation with a full-day representation
of use.

### Estimating effort {#sec-aerial-estimate}

`estimate_effort_aerial_glmm()` fits the negative-binomial GLMM, integrates
the fitted curve across `h_open` hours, and expands to the full season using
the stratified frame. The result is a total angler-hours estimate for the
survey period with a delta-method standard error.

```{r}
#| label: aerial-effort-estimate
#| echo: true
effort_aerial <- estimate_effort_aerial_glmm(design_aerial, time_col = time_of_flight)
effort_aerial
```

The GLMM estimates
`r round(effort_aerial$estimates$estimate)` total angler-hours
(95% CI: `r round(effort_aerial$estimates$ci_lower)`–`r round(effort_aerial$estimates$ci_upper)`).
The total and its standard error reflect two sources of variance: between-day
variation in total use (absorbed by the GLMM day-level random effect) and
between-stratum variation in sampling intensity (handled by the stratified
expansion). The ratio of `se_between` to total `se` indicates how much of the
uncertainty comes from which-days-were-sampled versus which-time-flights-were-
conducted.

The GLMM requires `lme4`. If the package is not installed, `tidycreel` will
prompt installation before the call proceeds.

### Naive estimate vs GLMM {#sec-aerial-naive}

To see what the time-of-day correction buys, it helps to compare the GLMM
result against the naive single-flight expansion. `example_aerial_counts`
contains one count per day — a midmorning flight — that can be treated as a
standard instantaneous-count design using `estimate_effort()`:

```{r}
#| label: aerial-naive-compare
#| echo: true
#| warning: false
# Standard instantaneous design with one flight per day
# h_open goes into creel_design(), not estimate_effort()
design_naive <- creel_design(
  calendar = aerial_calendar,
  date     = "date",
  strata   = "day_type",
  h_open   = 14
) |>
  add_counts(example_aerial_counts) |>
  add_interviews(
    interviews    = example_aerial_interviews,
    catch         = "walleye_catch",
    harvest       = "walleye_kept",
    effort        = "hours_fished",
    trip_status   = "trip_status",
    trip_duration = "hours_fished"
  )

effort_naive <- estimate_effort(design_naive)
effort_naive$estimates
```

```{r}
#| label: fig-aerial-compare
#| echo: true
#| fig-cap: "Aerial effort estimates from a naive single-flight expansion versus the GLMM time-of-day model. The naive estimate treats a midmorning count as representative of the full day; the GLMM integrates a fitted density curve. The difference illustrates the timing bias that motivates multi-flight designs."
#| fig-alt: "Point-and-range plot with two rows, one per method. Both show total angler-hours with 95% confidence intervals. The naive estimate is higher than the GLMM estimate, indicating upward bias from the midmorning flight timing."
bind_rows(
  effort_naive$estimates  |> mutate(method = "Naive (single flight × h_open)"),
  effort_aerial$estimates |> mutate(method = "GLMM (multi-flight, time-of-day model)")
) |>
  ggplot(aes(estimate, method)) +
  geom_pointrange(aes(xmin = ci_lower, xmax = ci_upper), size = 0.8) +
  labs(
    x = "Total angler-hours (season estimate)",
    y = NULL
  ) +
  theme_bw(base_size = 12)
```

The midmorning flight catches anglers near the daily peak, so the naive
expansion inflates the estimate relative to the GLMM, which integrates the
full day including early-morning and late-afternoon periods of lower activity.
The confidence intervals also differ: the naive estimate has wider uncertainty
because it does not leverage the within-day temporal structure to constrain
the daily total.

### Catch rate from an aerial design {#sec-aerial-cpue}

Remote count designs change how effort is estimated but leave the catch-rate
pipeline unchanged. `estimate_catch_rate()` on an aerial design object behaves
identically to Ch04 and Ch13:

```{r}
#| label: aerial-cpue
#| echo: true
estimate_catch_rate(design_aerial)
```

The CPUE estimate uses completed-trip interviews only, exactly as in a
standard access-point design. The `design_type = "aerial"` flag affects only
effort estimation; catch rate, harvest, and release estimation are design-type
agnostic.

That asymmetry is deliberate: aerial data revise the effort side of the
estimate, while completed-trip interviews still supply the catch-rate side.
Only the effort side is sensor-specific.

## Camera-assisted designs {#sec-camera}

### Access-point counters {#sec-camera-counter}

Boat-ramp cameras and vehicle counters take a different approach to remote
counting: instead of a snapshot of who is on the water, they record every
arrival (ingress) and, when configured to capture both directions, every
departure (egress). The count is not a density at a moment in time — it is a
cumulative total of use-events for the day.

The challenge this introduces is **downtime**: cameras fail. Batteries die,
memory cards fill, lenses ice over, and power outages interrupt recording.
A day with camera downtime is a day where the count is missing or
underrepresented. If those days are disproportionately high-use days —
weekends, good-weather days — excluding them biases the effort estimate
downward.

In other words, the camera can fail precisely when use is highest. That makes
missingness part of the design problem, not just a maintenance issue.

`impute_camera_counts()` handles missing counts using a stratum-specific
generalized linear model fitted to the operational days. The imputed values
carry a flag (`.imputed = TRUE`) that allows downstream diagnostics to
distinguish observed from modeled counts.

### Camera data and downtime {#sec-camera-data}

The bundled `example_camera_counts` dataset has one day with `NA` ingress
count due to battery failure:

```{r}
#| label: camera-data
#| echo: true
example_camera_counts
```

`impute_camera_counts()` fills the missing value using a within-stratum
GLM fitted to the operational rows. The imputed count is flagged in the
`.imputed` column:

```{r}
#| label: camera-impute
#| echo: true
camera_counts_clean <- impute_camera_counts(
  example_camera_counts,
  count_col  = "ingress_count",
  strata_col = "day_type",
  status_col = "camera_status"
)
camera_counts_clean
```

Weekday mean ingress on operational days is 51 vehicles; the imputed value
for June 11 (weekday, battery failure) is 51 — the stratum mean. With more
operational weekdays in the training set, the GLM would use day-of-week or
environmental covariates if they were present in the data.

@fig-camera-counts shows the ingress counts across the survey period, with
the imputed observation visually distinguished.

```{r}
#| label: fig-camera-counts
#| echo: true
#| fig-cap: "Daily boat-ramp camera ingress counts over the survey period. Observed counts (solid points) are distinguished from the imputed value on June 11 (open triangle, battery failure). Weekend counts are consistently higher than weekday counts, reflecting the typical recreational fishing pattern."
#| fig-alt: "Scatter plot with date on the x axis and ingress count on the y axis. Points are coloured by day type (weekday blue, weekend orange) and shaped by imputed status (solid circles for observed, open triangle for imputed). Weekend points cluster around 80-100; weekday points cluster around 45-60."
camera_counts_clean |>
  mutate(date = as.Date(date)) |>
  ggplot(aes(date, ingress_count, colour = day_type, shape = .imputed)) +
  geom_point(size = 3.5) +
  scale_colour_manual(
    values = c(weekday = "#2171b5", weekend = "#e6550d"),
    labels = c(weekday = "Weekday", weekend = "Weekend"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c(`FALSE` = 16, `TRUE` = 2),
    labels = c(`FALSE` = "Observed", `TRUE` = "Imputed (battery failure)"),
    name   = NULL
  ) +
  labs(
    x = NULL,
    y = "Ingress count (vehicles)"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

### Building the camera design {#sec-camera-design}

Camera designs use `design_type = "camera"` and require `camera_mode` to
specify what the counter records. The current supported modes are:

- `"counter"` — a simple ingress counter (vehicles or anglers entering)
- `"ingress_egress"` — paired ingress and egress timestamps, from which
  `preprocess_camera_timestamps()` computes trip-level effort

The `counter` mode is the most common and requires only the imputed count
table passed to `add_counts()`. This is the simplest camera case: the sensor
gives a daily traffic total, and the analyst reconstructs angler-hours from
counts and mean trip length.

```{r}
#| label: camera-design-build
#| echo: true
camera_calendar <- tibble(
  date     = seq(min(example_camera_counts$date),
                 max(example_camera_counts$date), by = "day"),
  day_type = if_else(wday(date) %in% c(1L, 7L), "weekend", "weekday")
)

design_camera <- creel_design(
  calendar    = camera_calendar,
  date        = "date",
  strata      = "day_type",
  design_type = "camera",
  camera_mode = "counter",
  h_open      = 14
) |>
  add_counts(camera_counts_clean) |>
  add_interviews(
    interviews    = example_camera_interviews,
    catch         = "walleye",
    harvest       = "walleye_kept",
    effort        = "hours_fished",
    trip_status   = "trip_status",
    trip_duration = "hours_fished"
  )
```

### Ratio-calibrated effort estimation {#sec-camera-estimate}

`est_effort_camera()` uses the ratio of mean interview trip duration to the
camera count — a ratio estimator that converts ingress events to angler-hours.
The intuition: if the camera records 60 vehicles entering and the mean trip
length from interviews is 3.5 hours, the day-level effort is approximately
60 × 3.5 = 210 angler-hours. The ratio is estimated within strata and
expanded to the full survey period:

```{r}
#| label: camera-effort-estimate
#| echo: true
effort_camera <- est_effort_camera(
  design_camera,
  interviews = example_camera_interviews,
  effort_col = "hours_fished"
)
effort_camera
```

The result is a total angler-hours estimate with a Taylor-linearization
standard error. For the example data,
`r round(effort_camera$estimates$estimate, 1)` angler-hours
(95% CI: `r round(effort_camera$estimates$ci_lower, 1)`–`r round(effort_camera$estimates$ci_upper, 1)`)
across the 10-day window. The `se_between` component reflects between-day
variation in the ingress × trip-length product; `se_within` reflects within-day
interview variance in trip duration.

That split tells you where the next precision gain is likely to come from:
more camera-days if between-day variation dominates, more interviews if
trip-length variation dominates.

As with the aerial design, catch rate estimation is unchanged:

```{r}
#| label: camera-cpue
#| echo: true
estimate_catch_rate(design_camera)
```

The `design_type` affects only how effort is estimated. From
`estimate_catch_rate()` onward, camera and access-point designs use identical
code paths.

### Ingress-egress timestamps {#sec-camera-timestamps}

When cameras record both arrival and departure times for individual vehicles,
`preprocess_camera_timestamps()` converts the raw timestamp pairs into
daily effort totals. This requires a data frame with one row per vehicle
event and `ingress_time` and `egress_time` columns as `POSIXct` values:

```{r}
#| label: camera-timestamps
#| echo: true
example_camera_timestamps

daily_effort <- preprocess_camera_timestamps(
  example_camera_timestamps,
  date_col    = "date",
  ingress_col = "ingress_time",
  egress_col  = "egress_time"
)
daily_effort
```

The output is a daily summary of total recorded angler-hours from timestamp
differences. This can substitute for or supplement the interview-based mean
trip duration estimate — the camera itself measures effort directly, rather
than inferring it from a count and a mean duration.

This is the cleanest camera case because effort is observed directly from the
start and end times rather than inferred from counts.

## Sample size for camera designs {#sec-camera-n}

Once a pilot camera season establishes the stratum-level mean and variance of
the ingress × trip-length product, `creel_n_camera()` returns the number of
sampled days needed to achieve a target coefficient of variation. The
required inputs mirror those of `creel_n_effort()`: target CV, total days in
the stratum, and stratum-level mean and variance from the pilot:

```{r}
#| label: camera-sample-size
#| echo: true
#| warning: false
# Compute stratum summary from operational days only
cam_summary <- camera_counts_clean |>
  filter(!.imputed) |>
  group_by(day_type) |>
  summarise(
    N_h    = n(),
    ybar_h = mean(ingress_count),
    s2_h   = var(ingress_count),
    .groups = "drop"
  )
cam_summary

# Days needed per stratum to achieve CV = 0.15
# creel_n_camera() requires named numeric vectors
creel_n_camera(
  cv_target = 0.15,
  N_h    = setNames(cam_summary$N_h,    cam_summary$day_type),
  ybar_h = setNames(cam_summary$ybar_h, cam_summary$day_type),
  s2_h   = setNames(cam_summary$s2_h,   cam_summary$day_type)
)
```

With only five operational weekday observations and four weekend observations,
the variance estimate is too thin to say much, so both strata bottom out at
n = 1. That is a toy-data artifact: real pilot seasons with 20–30 operational
days per stratum usually provide enough spread in ingress counts to make the
calculation informative. Treat the output as a minimum sample-size target for
each stratum, not as a precise optimum. `creel_n_camera()` also emits a warning
when the pilot falls below the empirical minimum recommended by Feltz–
Middaugh (2025); that warning is suppressed here because the example data
intentionally fall below it.

## Proxy counts and validation {#sec-proxy}

### Vehicle counters and gate sensors {#sec-proxy-types}

Aerial counts and camera-monitored ramps are the most thoroughly supported
remote-count methods in `tidycreel`, but other proxy systems appear in
practice: pneumatic road tubes at boat launch entrances, infrared gate sensors
at trailheads, and acoustic hydrophone arrays that detect boat motors in
restricted channels. Each records a correlated signal of fishing activity
rather than a direct observation of anglers.

The key question for any proxy is **how strongly it tracks what the survey
domain requires**. A pneumatic tube at a single-access boat ramp on a lake
with no bank fishing likely correlates nearly 1:1 with launched boats. A
road-tube counter at a multi-access reservoir where some anglers walk in from
informal paths captures only the trailered-boat fraction. The correlation
structure — and therefore the calibration factor — depends on access geometry,
angler behavior, and whether the proxy integrates the full reporting domain.

### Calibration against ground truth {#sec-proxy-calibration}

Paired sampling is the standard validation approach: on a subset of days,
both the proxy count and a direct ground count are collected on the same time
window. The ratio of ground count to proxy count on paired days becomes the
calibration factor applied to proxy-only days.

A calibration regression extends this: `ground = a + b × proxy` with the
slope $b$ estimated from paired days. The intercept accounts for baseline
angler activity not captured by the proxy (walk-in anglers, for example),
and the slope converts proxy units to ground-count equivalents.

```{r}
#| label: proxy-calibration
#| echo: true
# Paired calibration: road-tube counter vs direct creel-crew count on the same days
# Simulated paired data: 15 days where both methods were deployed simultaneously
set.seed(42)
paired_days <- tibble(
  date        = seq.Date(as.Date("2024-06-03"), by = "2 days", length.out = 15),
  tube_count  = as.integer(rpois(15, lambda = 55)),
  crew_count  = as.integer(round(0.82 * tube_count + rnorm(15, 0, 4)))
)

cal_fit <- lm(crew_count ~ tube_count, data = paired_days)
coef(cal_fit)
```

The intercept (`r round(coef(cal_fit)[1], 1)`) captures angler activity not
counted by the road tube — walk-in anglers, kayak launches from an unmonitored
site. The slope (`r round(coef(cal_fit)[2], 3)`) converts tube-count units to
crew-count equivalents. On proxy-only days, multiply the tube count by this
slope and add the intercept to recover the calibrated angler count before
applying mean trip length.

A slope near 1.0 indicates the proxy and direct count are on the same scale;
a slope below 1 indicates systematic undercounting — the tube misses some
fraction of the angler population — that the calibration corrects. The
regression diagnostic ($R^2$, residual plot) should be checked before trusting
the calibration: if the relationship is noisy or nonlinear, a simple ratio
calibration will not transfer reliably to proxy-only days.

### Documenting downtime and coverage gaps {#sec-proxy-downtime}

No remote monitoring system has perfect uptime. Effective survey documentation
requires a record of:

- Days when the system was fully operational
- Days with partial downtime (camera down for X of Y hours)
- Days with complete failure (no data)
- Known failure modes (battery, connectivity, vandalism, weather)

The `camera_status` column in `example_camera_counts` illustrates the minimum
required record. In practice, richer downtime records allow the imputation
model to condition on failure type — battery failures cluster by temperature
and season; connectivity failures cluster by storm events.

```{r}
#| label: downtime-summary
#| echo: true
example_camera_counts |>
  count(camera_status, day_type) |>
  arrange(desc(n))
```

A camera with 90% uptime over a 30-day survey loses roughly 3 operational
days. If those failures are random relative to angler activity, imputation adds
noise but not much bias. If failures correlate with weather — cameras fail in
cold rain, which also suppresses fishing — the imputed values reflect low-use
conditions, and the estimate is biased downward. Understanding the failure
mechanism matters as much as the imputation method.

## `tidycreel` support summary {#sec-remote-tidycreel}

The three major remote-count workflows and their `tidycreel` entry points:

| Design | `design_type` | Count function | Estimate function |
|--------|---------------|---------------|-------------------|
| Aerial GLMM | `"aerial"` | `add_counts()` (raw flights) | `estimate_effort_aerial_glmm()` |
| Camera counter | `"camera"` | `add_counts()` (imputed) | `est_effort_camera()` |
| Camera ingress/egress | `"camera"` + timestamps | `preprocess_camera_timestamps()` | `est_effort_camera()` |
| Proxy count (manual calibration) | any | calibration regression | ratio applied to standard estimate |

All four share the same downstream pipeline: design object → `add_interviews()`
→ `estimate_catch_rate()` / `estimate_total_harvest()`. The remote-count
method changes how effort is estimated; catch rate and harvest estimation
proceed identically to access-point and roving designs.

## Key considerations {#sec-remote-considerations}

**Frame alignment.** The camera or aircraft must cover the same spatial
domain as the reporting unit. A ramp camera that misses an informal carry-in
launch is not counting the full angler population. Aerial coverage that
excludes a productive back bay leaves effort in that bay unrecorded. Frame
gaps in remote designs are harder to detect than in creel-crew designs
because there is no field crew to notice unexpected activity outside the
monitored area.

**Timing alignment.** For aerial designs, `h_open` must match what the GLMM
integrates over. If the model integrates from 06:00 to 20:00 (14 hours) but
fishing in the reporting domain occurs only between 07:00 and 18:00, the
integration window overestimates effort. Setting `h_open` conservatively from
local sunrise-to-sunset tables avoids this bias.

**Sensor resolution.** A camera that counts vehicles counts launched boats
(typically one to four anglers per boat). Converting vehicle counts to
angler-hours requires an estimate of mean party size from interviews — an
additional ratio estimator step with its own variance. The standard error
from `est_effort_camera()` propagates this uncertainty automatically when
party-size information is included in `add_interviews()`.

**Uptime documentation.** Effort estimates from camera designs are only as
reliable as the imputation model for failed days. More operational days in
the calibration sample → better imputation. Effort estimates should report
the fraction of days that were imputed alongside the estimate and standard
error, so users can assess how much of the result rests on modeled versus
observed counts.

Remote count designs extend the creel survey toolkit substantially: larger
waterbodies become tractable, high-frequency temporal monitoring becomes
affordable, and record-keeping becomes auditable. What they do not change is
the need for a defined frame, a defensible sampling design, and explicit
assumptions about what the sensor measures and what it misses.
