---
title: "Creel Data as Linked Tables"
---

```{r}
#| label: setup-ch09
#| include: false
library(tidycreel)
library(tidyverse)

harlan_schedule   <- readRDS(here::here("data/harlan_schedule.rds"))
harlan_counts     <- readRDS(here::here("data/harlan_counts.rds"))
harlan_interviews <- readRDS(here::here("data/harlan_interviews.rds"))
harlan_catch      <- readRDS(here::here("data/harlan_catch.rds"))
```

## Why creel data resists a single spreadsheet {#sec-linked-why}

A creel survey produces at least four distinct kinds of records. The sampling
schedule records which days were eligible for sampling and which were actually
sampled. Count rounds record how many anglers were observed at each site and time
during those sampled days. Interviews record what individual angler parties
reported about their trips. Catch records break each party's report into
species-level outcomes.

Each record type answers a different question and belongs to a different
observation unit. A count record belongs to a site-period combination. An
interview record belongs to an angler party. A catch record belongs to a single
species-outcome combination within a single interview. These are different units
and should not share rows in the same table.

The practical consequence is statistical. Effort estimators divide an observed
angler count by an inclusion probability derived from the schedule. CPUE
estimators divide total catch by total angler-hours, where angler-hours come from
individual trip records. When the data are stored in a single flat spreadsheet, a
join that combines catch with trip records creates duplicate rows — and the
denominator for the trip-hours calculation expands with them. The estimate changes
not because the underlying data changed but because the data structure changed.

Keeping records in separate tables and joining them only when an analysis
requires it is the safest guard against this class of error. This chapter
introduces the four Harlan Reservoir survey tables, explains their keys and
relationships, and shows how `tidycreel` validates the structure before any
estimation begins.

## The four Harlan tables {#sec-four-tables}

The Harlan 2022 creel survey data are distributed across four files, one per
record type. Together they cover `r nrow(harlan_schedule)` scheduled days,
`r nrow(harlan_counts)` count observations, `r nrow(harlan_interviews)` angler
party interviews, and `r nrow(harlan_catch)` catch records.

### Sampling schedule

The schedule is the sampling frame: the complete list of days that were available
to be sampled and a flag for whether each was actually sampled. It is the
denominator layer — every inclusion probability calculation for the season traces
back to this table.

```{r}
#| label: schedule-glimpse
harlan_schedule
```

The four columns carry distinct roles. `date` is the calendar day. `reservoir`
identifies the waterbody — a single schedule can cover multiple reservoirs if they
share a survey design. `day_type` records whether the day was a weekend or weekday,
which is the primary stratum in the Harlan design. `sampled` is the binary flag
that separates the realized sample from the non-sampled days in the frame.

```{r}
#| label: schedule-summary
harlan_schedule |>
  count(sampled, day_type)
```

Of `r nrow(harlan_schedule)` days in the frame,
`r sum(harlan_schedule$sampled)` were sampled — roughly
`r round(100 * mean(harlan_schedule$sampled))`% of the available days. Weekend and
weekday days were sampled at similar rates, reflecting a stratified design where
both strata are explicitly covered.

A day that is absent from the schedule cannot be assigned an inclusion probability.
Days outside the frame are outside the survey population by definition, so any
count or interview record that falls on an unscheduled date is a data quality
problem rather than a legitimate observation.

### Count records

Count rounds record the instantaneous number of anglers observed at each site
during a sampled day. This is the raw observation that gets converted to an
estimate of angler effort (@sec-effort-why). Each row is
one count at one site at one moment in time.

```{r}
#| label: counts-glimpse
harlan_counts
```

Four structural columns require explanation. `period` identifies the count round
within the day — in the Harlan design, period 1 is the morning count and period 2
is the afternoon count. `section` identifies the spatial sub-area of the reservoir
that was surveyed; section "0" is a whole-lake count, and other values identify
geographic sub-sections. `count_time` records the exact clock time of the
observation, which is used to estimate the fraction of the day each count
represents. Together, `period`, `section`, and `count_time` partition the effort
counts into the strata and time windows the estimator will sum over.

The three count columns — `bank_anglers`, `angler_boats`, and `boat_anglers` —
record observed counts of individual anglers on the bank, boats containing anglers,
and the total anglers estimated to be in those boats respectively. `boat_anglers`
is the count used in effort estimation; `angler_boats` supports boat-party
interception rate calculations.

```{r}
#| label: counts-summary
harlan_counts |>
  group_by(period) |>
  summarise(
    n_count_events = n(),
    total_bank     = sum(bank_anglers),
    total_boats    = sum(angler_boats),
    total_boat_ang = sum(boat_anglers),
    .groups = "drop"
  )
```

The counts table records observed anglers, not interviewed anglers. A high count
on a given day does not imply a high number of interviews on that day; a site
could be counted without any parties accepting an interview.

### Interview records

Each row in the interviews table represents one angler party intercept. The table
holds trip-level information: who was sampled, when, how they fished, and how far
into their trip they were when the interviewer found them.

```{r}
#| label: interviews-glimpse
harlan_interviews
```

Two identifier columns coexist in this table and serve different purposes.
`party_id` is a UUID assigned in the field data system — it is stable across
potential data system reloads or imports. `interview_id` is a sequential integer
assigned at analysis time and is the key used to link this table to the catch
table. In practice `interview_id` is simpler to work with; `party_id` is the audit
trail back to the original record in the field system.

`trip_status` records whether the interview captured a complete trip (the angler
had finished fishing) or an incomplete trip (the angler was still fishing when
intercepted). This distinction matters for catch estimation: incomplete trip catch
cannot simply be summed with complete trip catch without applying a correction for
the unobserved portion of the trip.

```{r}
#| label: interviews-summary
harlan_interviews |>
  summarise(
    n_interviews     = n(),
    n_sampled_dates  = n_distinct(date),
    complete_trips   = sum(trip_status == "complete"),
    incomplete_trips = sum(trip_status == "incomplete"),
    boat_parties     = sum(angler_type == "boat"),
    bank_parties     = sum(angler_type == "bank"),
    mean_hours       = round(mean(hours_fished), 2)
  )
```

The `r nrow(harlan_interviews)` interviews span `r n_distinct(harlan_interviews$date)`
sampled dates — not all sampled days yielded interviews, and not all interviews
occurred on every sampled day. The mean trip duration across the season was
`r round(mean(harlan_interviews$hours_fished), 1)` hours, though this mixes complete and
incomplete trips.

### Catch records

The catch table holds one row per species per catch outcome (harvested or
released) per interview. There is no trip-level information in this table: `interview_id`
is the only link back to the party, and everything else is species-level.

```{r}
#| label: catch-glimpse
harlan_catch
```

The structure means a single interview with three species can appear as six rows
in the catch table — one harvested and one released row per species. This is
intentional. Storing species and outcomes in separate rows keeps the table tidy
as species lists expand and makes it straightforward to filter, aggregate, or
join by species or outcome.

```{r}
#| label: catch-summary
harlan_catch |>
  group_by(species) |>
  summarise(
    n_records  = n(),
    harvested  = sum(n_fish[catch_type == "harvested"]),
    released   = sum(n_fish[catch_type == "released"]),
    .groups = "drop"
  ) |>
  arrange(desc(harvested + released))
```

Eleven species appear in the catch records for the 2022 Harlan season. White Bass
and Walleye dominate the harvest numbers — a pattern consistent with Harlan
Reservoir's well-known walleye fishery and its abundant forage-fish populations.

## Keys and join relationships {#sec-keys-joins}

The four tables connect through two types of keys. Understanding which key
connects which pair of tables is required before any join can be written
correctly.

**Composite keys** link the schedule to the counts and interviews tables.
No single column uniquely identifies a sampling event; instead, the combination
of `reservoir`, `date`, `period`, and `section` identifies one count round or one
interview occasion. When a count round and an interview occurred at the same
site-period, those records carry the same composite key values and can be joined.

**Surrogate keys** link interviews to catch. The `interview_id` column is a single
integer that uniquely identifies one party intercept and appears in `harlan_catch`
as a foreign key. Every catch row belongs to exactly one interview row.

The join path follows a strict hierarchy:

| From | To | Key columns |
|------|----|-------------|
| `harlan_schedule` | `harlan_counts` | `date`, `reservoir` |
| `harlan_schedule` | `harlan_interviews` | `date`, `reservoir` |
| `harlan_counts` | `harlan_interviews` | `date`, `period`, `section`, `reservoir` |
| `harlan_interviews` | `harlan_catch` | `interview_id` |

The schedule is the root. Counts and interviews are both children of the schedule,
connected through shared date and context columns. Catch is a child of interviews,
connected through the surrogate key.

Joining in the wrong direction or joining without all required key columns are
the two most common sources of phantom rows and silent data loss.

The counts-to-interviews join is worth demonstrating explicitly. It produces one
output row per interview, augmented with the count context from the same
site-period:

```{r}
#| label: counts-interviews-join
counts_context <- harlan_counts |>
  select(reservoir, date, period, section, bank_anglers, boat_anglers)

interviews_with_context <- left_join(
  harlan_interviews,
  counts_context,
  by = c("reservoir", "date", "period", "section")
)

# Row count should match interviews exactly
nrow(interviews_with_context) == nrow(harlan_interviews)

interviews_with_context |>
  select(interview_id, date, period, section, hours_fished, bank_anglers, boat_anglers) |>
  head(5)
```

The row count is preserved — one output row per interview — and each interview
row now carries the angler count from the concurrent count round. A mismatch in
row count would indicate either an interview with no corresponding count or
multiple count rounds matching the same interview key.

## Validating joins {#sec-validate-joins}

Because the tables are linked by keys, each join can be tested before it is used
in estimation. The standard tests are anti-joins (finding unmatched records on
either side) and row-count comparisons (confirming that joins expand or contract
the row count as expected).

### Catch records without interviews

Every catch row should trace back to a real interview. An orphan catch record —
one whose `interview_id` does not exist in the interviews table — indicates either
a data entry error or a deleted interview that left catch records behind.

```{r}
#| label: catch-orphan-check
orphan_catch <- anti_join(harlan_catch, harlan_interviews, by = "interview_id")
nrow(orphan_catch)
```

Zero orphans. Every catch record in the 2022 Harlan data has a matching interview.
This check is not guaranteed to pass on data from other projects or other years, so
it should be part of any standard data preparation script.

### Interviews without catch records

The reverse anti-join finds interviews that produced no catch records. This is
expected — anglers who caught nothing or refused to report catch produce an
interview row but no corresponding catch rows. The number should be interpretable
given the fishery.

```{r}
#| label: interviews-no-catch
no_catch_interviews <- anti_join(
  harlan_interviews,
  harlan_catch,
  by = "interview_id"
)

no_catch_interviews |>
  count(trip_status, name = "n_interviews")
```

Of the `r nrow(harlan_interviews)` interviews,
`r nrow(no_catch_interviews)` produced no catch records. The breakdown by
`trip_status` shows that most of these are incomplete trips. Parties still fishing
when intercepted often had not yet landed fish, so the absence of catch records is
consistent with the data — not an error. @sec-trip-status covers how to handle
incomplete-trip catch in the estimation step.

### Schedule versus counts

Count records should come from days that the schedule marks as sampled. A count
date that is absent from the schedule as a sampled day suggests either a data
entry error in the schedule or an unsanctioned extra sampling day.

```{r}
#| label: schedule-counts-check
sampled_dates <- harlan_schedule |>
  filter(sampled) |>
  pull(date)

harlan_counts |>
  filter(!date %in% sampled_dates) |>
  distinct(date, reservoir)
```

Two count dates fall outside the set of days marked as sampled in the Harlan
schedule. In a flat file these two dates would be indistinguishable from the rest
of the count records. In the linked-table structure, the discrepancy surfaces in
two lines of code. Whether the correct response is to add those dates to the
schedule or to remove the counts is a domain decision; the structure makes the
problem findable.

## Validating tables with `validate_creel_data()` {#sec-validate-data}

`tidycreel` provides `validate_creel_data()` to run a systematic battery of checks
on count and interview tables. It verifies column types, flags impossible values
(negative counts, implausible dates, NA values in identifier fields), and reports
the result as a structured pass/warn/fail summary.

```{r}
#| label: validate-creel-data
validate_creel_data(
  counts     = harlan_counts,
  interviews = harlan_interviews
)
```

The Harlan data passes all 63 checks with no warnings and no failures. This
outcome is worth recording at the start of any analysis session: a passing run
confirms the tables have not been modified in a way that would silently break
downstream estimation. A failure identifies specific columns and rows to
investigate before estimation proceeds.

The checks `validate_creel_data()` runs are not redundant with the anti-joins in
@sec-validate-joins. The anti-joins test relational integrity between tables. The
validation function tests internal integrity within each table: whether a date
column contains actual dates, whether angler counts contain negative numbers,
whether identifier columns have unexpected missing values. Both sets of checks
should pass before estimation begins.

## Formalizing the schema with `creel_schema()` {#sec-schema}

`creel_schema()` records the mapping from your column names to the conceptual
roles that `tidycreel` estimation functions expect. Survey data rarely arrives with
column names that match package conventions. Rather than forcing users to rename
columns in every function call, the schema holds the mapping once and passes it
forward.

```{r}
#| label: creel-schema-build
harlan_schema <- creel_schema(
  survey_type        = "instantaneous",
  interviews_table   = "harlan_interviews",
  counts_table       = "harlan_counts",
  catch_table        = "harlan_catch",
  date_col           = "date",
  effort_col         = "hours_fished",
  trip_status_col    = "trip_status",
  interview_uid_col  = "interview_id",
  bank_anglers_col   = "bank_anglers",
  angler_boats_col   = "angler_boats",
  n_anglers_col      = "n_anglers",
  angler_type_col    = "angler_type",
  angler_method_col  = "angler_method",
  species_col        = "species",
  catch_count_col    = "n_fish",
  catch_type_col     = "catch_type"
)

harlan_schema
```

The printed schema shows which column in each table maps to which conceptual role.
`effort -> hours_fished` means that the `hours_fished` column in the interviews
table carries the trip-duration information the estimator uses. `catch_count -> n_fish`
means that `n_fish` in the catch table is the count column.

Schemas are useful because the data are in separate tables. If all columns were
in one flat file, distinguishing between `hours_fished` (trip-level) and
`n_fish` (catch-level) would require column-name conventions rather than table
structure.

In the next chapter, the schema is consumed by `creel_design()` to produce the
design object that estimation functions take as their primary argument.

## Standardizing columns with `prep_interview_catch()` {#sec-prep}

`tidycreel` includes a family of `prep_*` functions that standardize column names
within individual tables before passing them to the package internals.
`prep_interview_catch()` handles the catch table.

```{r}
#| label: prep-interview-catch
prepped_catch <- prep_interview_catch(
  data          = harlan_catch,
  interview_uid = "interview_id",
  species       = "species",
  count         = "n_fish",
  catch_type    = "catch_type"
)

prepped_catch
```

The output carries standardized names — `interview_uid`, `species`, `count`,
`catch_type` — that the internal estimation pipeline expects. Running
`prep_interview_catch()` at the start of an analysis session means that column
name differences between projects do not propagate into estimation errors.

The prep functions are thin wrappers: they rename columns and return the data
with the same number of rows. They do not validate values, impute missing data,
or recode factors. Validation (via `validate_creel_data()`) and join checking
(via anti-joins) should run before the prep step, not after it.

## The danger of premature flattening {#sec-flattening}

The most common data management mistake in creel analysis is joining tables
before the analysis requires it. The typical failure mode is joining interviews
to catch before aggregating catch to the party level. The interview table has one
row per party. The catch table has one row per species per outcome per party.
The join expands the interview table, and any trip-level statistic computed on
the result uses the wrong denominator.

```{r}
#| label: premature-flatten-demo
interviews_flat <- left_join(
  harlan_interviews,
  harlan_catch,
  by = "interview_id"
)

tibble(
  source   = c("harlan_interviews", "harlan_catch", "flat join"),
  n_rows   = c(nrow(harlan_interviews), nrow(harlan_catch), nrow(interviews_flat))
)
```

The flat join produces `r nrow(interviews_flat)` rows from an interview table with
`r nrow(harlan_interviews)` rows. Interviews with multiple species-outcome
combinations appear multiple times. To see how many times each interview appears:

```{r}
#| label: flatten-duplication-check
interviews_flat |>
  count(interview_id, name = "n_rows_in_join") |>
  count(n_rows_in_join, name = "n_interviews")
```

Most interviews appear more than once in the flat join. An interview that
reported three species with two outcomes each would appear six times. If
trip-level effort hours are then averaged on the flat file, the denominator is
inflated and the mean is wrong:

```{r}
#| label: premature-flatten-wrong-mean
tibble(
  calculation = c("interview table (correct)", "flat join (wrong)"),
  mean_hours  = c(
    mean(harlan_interviews$hours_fished),
    mean(interviews_flat$hours_fished)
  )
)
```

The correct pattern is to aggregate catch to the party level first, then join. The
resulting joined table has the same row count as the interview table.

```{r}
#| label: correct-join-pattern
catch_by_interview <- harlan_catch |>
  group_by(interview_id) |>
  summarise(
    total_fish     = sum(n_fish),
    species_caught = n_distinct(species[n_fish > 0]),
    .groups = "drop"
  )

interviews_with_catch <- left_join(
  harlan_interviews,
  catch_by_interview,
  by = "interview_id"
)

nrow(interviews_with_catch)
```

Aggregating before joining preserves the one-row-per-party structure:

```{r}
#| label: correct-join-result
interviews_with_catch |>
  select(interview_id, date, angler_type, hours_fished, total_fish, species_caught) |>
  head(5)
```

The row count matches the interview table and each row carries trip-level effort
alongside party-level catch totals. This is the correct shape for CPUE
calculation: one row per trip, with hours and catch from the same observation
unit.

## Visualizing the data hierarchy {#sec-viz-hierarchy}

### Record counts across tables

The observation hierarchy is clearest when the row counts are shown together. Each
table is a different resolution of the same survey: one sampled day produces
multiple count events, those count events coincide with multiple interviews, and
each interview yields multiple catch records.

```{r}
#| label: fig-record-counts
#| fig-cap: "Row counts for the four Harlan survey tables. Each table represents a different observation unit. The increasing counts from schedule to catch reflect the one-to-many relationships in the join hierarchy."
#| fig-alt: "Horizontal bar chart with four bars showing row counts for each Harlan survey table, ordered from fewest rows (schedule) to most rows (catch)."

record_counts <- tibble(
  table = c("harlan_schedule", "harlan_counts", "harlan_interviews", "harlan_catch"),
  n     = c(nrow(harlan_schedule), nrow(harlan_counts), nrow(harlan_interviews), nrow(harlan_catch)),
  role  = c("sampling frame", "effort counts", "party interviews", "catch records")
) |>
  mutate(table = fct_inorder(table))

ggplot(record_counts, aes(x = n, y = fct_rev(table), fill = role)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = paste0(n, "  (", role, ")")),
    hjust = -0.05, size = 3.2
  ) +
  scale_fill_manual(values = creel_palette(4), guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.25))) +
  labs(
    x     = "Number of rows",
    y     = NULL,
    title = "Harlan Reservoir 2022 — observation hierarchy"
  ) +
  theme_creel()
```

The schedule and counts rows are similar in number —
`r nrow(harlan_schedule)` and `r nrow(harlan_counts)` respectively — because a
typical sampled day produces only a few count events (one per period per section).
Interviews are more numerous at `r nrow(harlan_interviews)` because multiple
parties are intercepted per count round. Catch records are the most numerous at
`r nrow(harlan_catch)` because each interview can produce records for multiple
species and outcomes.

### Coverage: which sampled days appear in each table?

Not every sampled day produces both counts and interviews. Plotting date coverage
reveals where data are dense and where they are sparse — and makes it immediately
visible that the three tables do not share identical date ranges.

```{r}
#| label: fig-date-coverage
#| fig-cap: "Date coverage across the schedule (sampled days), counts, and interview tables for Harlan Reservoir 2022. Points indicate days with records in each table. Days that appear in the schedule but not in counts or interviews indicate sampled days with no recorded activity."
#| fig-alt: "Three-row dot plot showing date coverage across the 2022 season for each of three tables: schedule, counts, and interviews."

coverage_data <- bind_rows(
  harlan_schedule |>
    filter(sampled) |>
    transmute(date, source = "schedule"),
  harlan_counts |>
    distinct(date) |>
    mutate(source = "counts"),
  harlan_interviews |>
    distinct(date) |>
    mutate(source = "interviews")
) |>
  mutate(source = factor(source, levels = c("schedule", "counts", "interviews")))

ggplot(coverage_data, aes(x = date, y = source, color = source)) +
  geom_point(size = 1.8, alpha = 0.7) +
  scale_color_manual(values = creel_palette(3), guide = "none") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(
    x     = "Date",
    y     = NULL,
    title = "Harlan Reservoir 2022 — sampled date coverage by table"
  ) +
  theme_creel() +
  theme(panel.grid.major.x = element_line(linewidth = 0.3))
```

Several dates appear in the schedule and counts rows but not in the interviews
row. A day can be counted without any parties agreeing to be interviewed. The
figure makes this gap visible. Treating the interview date range as equivalent
to the sampled date range would overstate the survey frame and understate the
non-response rate.

## Table ownership rules {#sec-ownership}

A relational model fails when the same information lives in multiple tables and
the copies diverge. Each table should own its fields exclusively — any column that
appears in two tables is a candidate for inconsistency over time.

The ownership assignments for a creel survey are driven by observation unit:

**Schedule** owns date-level information: which dates were in the frame, their
`day_type`, and whether they were sampled. No other table should store `day_type`
as an original column; if counts or interviews need `day_type`, they should derive
it by joining to the schedule rather than by maintaining their own copy. In the
Harlan data, `day_type` appears in both `harlan_counts` and `harlan_interviews` as
a convenience column. Verifying that these copies are consistent with the schedule
is a mandatory check:

```{r}
#| label: day-type-consistency
# Check that day_type in counts matches schedule
left_join(
  harlan_counts |> select(date, day_type_counts = day_type),
  harlan_schedule |> select(date, day_type_schedule = day_type),
  by = "date"
) |>
  filter(day_type_counts != day_type_schedule) |>
  nrow()
```

Zero mismatches — the `day_type` copies in Harlan counts are consistent with the
schedule. This check costs one join and should be part of any data preparation
script. If mismatches exist, the schedule is the authoritative source and the
copies in counts and interviews should be corrected by joining from the schedule,
not by hand-editing individual rows.

**Counts** own site-period-level observations: the number of anglers of each type
seen at a given section during a given count round. They do not own interview
counts or any trip-level information.

**Interviews** own party-level trip information: interception time, trip duration,
method, angler type, and trip completion status. They do not own species-level
catch. If an interview table contains a column for total catch by species, that
total will diverge from the sum of the catch table whenever catch records are
edited after the fact.

**Catch** owns species-level outcomes for each interview. It holds `interview_id`
as a foreign key but not any other trip-level columns. If catch records carry
`hours_fished` or `date` copied from the interview, edits to the interview no
longer propagate to the catch copies, and a join will produce two different values
for the same attribute.

Violations of these rules are common in creel programs that began as spreadsheets
and were later imported into a data system. Columns get copied as a convenience
and then diverge. The linked-table approach makes these copies unnecessary: the
information is always available through a join from the table that owns it.

## Common mistakes {#sec-common-mistakes}

Two mistakes recur in creel data management that are not fully covered by the
examples above.

**Silent loss from inner joins.** An inner join between interviews and catch drops
the `r nrow(no_catch_interviews)` Harlan interviews that have no catch records.
Those parties fished — their effort hours belong in the CPUE denominator. Dropping
them biases CPUE upward. The correct join is always a left join from interviews to
catch; the `NA` values that appear for no-catch parties are informative and should
be replaced with zero fish, not removed.

**Confusion between counts and interview denominators.** After the counts-to-
interviews join demonstrated in @sec-keys-joins, the joined table carries both
`bank_anglers` (from counts — all anglers visible at the site) and `n_anglers`
(from interviews — anglers in the intercepted party). These are different
quantities. `bank_anglers` is the instantaneous census; `n_anglers` is the party
size. Using the wrong column in an effort estimator produces a result that appears
numerically plausible but is conceptually wrong. Keeping the columns in separate
tables before joining makes the distinction visible; a flat file obscures it.

## Takeaway {#sec-linked-takeaway}

Creel data is naturally relational. The sampling frame, effort counts, party
interviews, and species-level catch records each describe a different
observation unit and belong in separate tables with stable keys connecting them.

Keeping these tables separate until the analysis requires them protects the row
counts and denominators that estimators depend on. It also makes data quality
problems discoverable: orphan catch records, count dates outside the schedule
frame, and interview dates that do not match sampled days all surface through
explicit join checks rather than manifesting silently as wrong estimates.

`tidycreel` supports this structure through `validate_creel_data()`, which checks
internal integrity within each table, and `creel_schema()`, which formalizes the
column mapping that estimation functions consume. The next chapter uses both to
construct the `creel_design` object.
