---
title: "Building a Creel Design Object"
---

```{r}
#| label: setup-ch10
#| include: false
library(tidycreel)
library(tidyverse)

harlan_schedule   <- readRDS(here::here("data/harlan_schedule.rds"))
harlan_counts     <- readRDS(here::here("data/harlan_counts.rds"))
harlan_interviews <- readRDS(here::here("data/harlan_interviews.rds"))
harlan_catch      <- readRDS(here::here("data/harlan_catch.rds"))
```

## What the design object is for {#sec-design-purpose}

The linked tables built in @sec-linked-why contain all the raw survey data, but they do
not encode the sampling design. The schedule table records which days were sampled,
but it does not specify the sampling units, the strata, or the inclusion
probabilities that estimation requires. The counts table records observed anglers,
but it does not say whether those counts came from a stratified sample or a simple
random sample of days.

Effort and catch estimators require this information [@pollock1994]. A stratified
estimator weights each stratum by its population size. A design-effect calculation
depends on knowing the primary sampling units. Without a single object that holds
all of this, the design logic has to be re-stated — and potentially re-stated
incorrectly — at every estimation step.

`tidycreel` addresses this with the `creel_design` object. The design object
holds the calendar, the strata, the counts, the interviews, and the species-level
catch in a single structure. Estimation functions from @sec-effort-why through @sec-uncertainty-anatomy take a
`creel_design` as their first argument, so the design object is the handoff point
between data preparation and estimation.

This chapter builds a `creel_design` for the Harlan 2022 survey, one step at a
time, and shows what each step computes and what checks the package runs along the
way.

## Step 1: the base design {#sec-design-base}

`creel_design()` initializes the object from the sampling schedule. It needs three
things: the schedule table, the name of the date column, and the name of the
strata column.

At this step, `creel_design()` constructs a placeholder `survey.design2` object
from the `survey` package using equal-probability assumptions. The message
`No weights or probabilities supplied, assuming equal probability` is expected here:
the schedule alone does not yet carry count observations, so the package has no
realized sampling fractions to use. The equal-probability assumption is replaced
when counts are added in Step 2.

```{r}
#| label: design-base
#| message: false
harlan_design <- creel_design(
  calendar = harlan_schedule,
  date     = "date",
  strata   = "day_type"
)

harlan_design
```

The printed output shows the core design metadata: survey type, date column,
strata variable, calendar span, and the number of stratum levels. Nothing about
counts, interviews, or catch appears yet.

The schedule itself tells the base design how many days were in each stratum and
how many were sampled:

```{r}
#| label: design-strata-sizes
harlan_schedule |>
  group_by(day_type) |>
  summarise(
    N_h      = n(),
    n_h      = sum(sampled),
    f_h      = round(mean(sampled), 3),
    .groups  = "drop"
  )
```

The Harlan design sampled approximately `r round(100 * mean(harlan_schedule$sampled))`% of
available days in each stratum — roughly equal sampling fractions in weekday and
weekend strata. Equal-fraction stratified sampling of this kind is common when
management interest extends to both stratum types equally.

## Step 2: adding count data {#sec-design-counts}

`add_counts()` attaches the count observations and tells the design object which
column records the exact time of each count. The count time matters because
instantaneous count estimators weight each observation by its position in the
counting day: a count taken at 10:00 represents a longer portion of the day than
one taken at 14:30 if counts are not uniformly spaced.

```{r}
#| label: design-counts
harlan_design <- add_counts(
  design         = harlan_design,
  counts         = harlan_counts,
  count_time_col = "count_time"
)

harlan_design
```

The design now records `r nrow(harlan_counts)` count observations, with `date` as
the primary sampling unit and `count_time` as the within-day variable. It also
shows that a `survey.design2` object has been constructed — this is the moment at
which the design moves from equal-probability assumptions to the actual realized
sampling fractions implied by the schedule and counts combined.

`add_counts()` returns a new design object rather than modifying the existing one.
This immutability means that `harlan_design` at Step 2 is a complete, independent
object. If the analysis later needs to compare different count configurations, the
Step 1 base design is still available and can be passed to a different `add_counts()`
call without conflict.

## Step 3: adding interview data {#sec-design-interviews}

`add_interviews()` attaches the party intercept records. The function expects a
column in the interviews table that records total catch per interview — not the
species-level table, but a single number per party. This column feeds the catch
estimator's numerator; the species-level breakdown is added separately in Step 4.

The Harlan interviews table does not include a precomputed catch total, so it is
derived from the catch table first. The join must start from `harlan_interviews`
(the complete set of 600 parties) and bring in harvest totals where they exist —
preserving the `r nrow(anti_join(harlan_interviews, harlan_catch, by = "interview_id"))`
parties that had no harvested catch as zero-row contributions:

```{r}
#| label: harvest-totals
harvest_by_interview <- harlan_catch |>
  filter(catch_type == "harvested") |>
  group_by(interview_id) |>
  summarise(total_harvest = sum(n_fish), .groups = "drop")

harlan_interviews_with_catch <- harlan_interviews |>
  left_join(harvest_by_interview, by = "interview_id") |>
  mutate(total_harvest = if_else(is.na(total_harvest), 0L, as.integer(total_harvest)))

# Verify: row count must match the original interviews table
nrow(harlan_interviews_with_catch) == nrow(harlan_interviews)

harlan_interviews_with_catch |>
  select(interview_id, date, hours_fished, trip_status, total_harvest) |>
  head(5)
```

Starting from `harlan_interviews` with `left_join` is the safest pattern: all
original rows survive, and `NA` in `total_harvest` after the join can only mean
the party had no harvested catch.

```{r}
#| label: design-interviews
harlan_design <- add_interviews(
  design         = harlan_design,
  interviews     = harlan_interviews_with_catch,
  catch          = "total_harvest",
  effort         = "hours_fished",
  trip_duration  = "hours_fished",
  trip_status    = "trip_status",
  n_anglers      = "n_anglers",
  angler_type    = "angler_type",
  angler_method  = "angler_method",
  interview_type = "access"
)

harlan_design
```

The output now shows interview counts broken down by trip completion status:
`r sum(harlan_interviews$trip_status == "complete")` complete and
`r sum(harlan_interviews$trip_status == "incomplete")` incomplete.
The 318 interviews flagged for zero catch are anglers who had not yet landed
anything when intercepted.

The `interview_type = "access"` argument tells the design that these are
access-point interviews: anglers intercepted at fixed locations as they exit the
fishery. This affects how the estimator handles incomplete trips. An access
interview of an incomplete trip can be projected to a full-trip equivalent; the
design needs to know the interview type to apply the correct correction.

## Step 4: adding species-level catch {#sec-design-catch}

The final builder step attaches the species-level catch records. These are the
rows from `harlan_catch` that link individual species and outcomes back to
interviews.

```{r}
#| label: design-catch
harlan_design <- add_catch(
  design        = harlan_design,
  data          = harlan_catch,
  catch_uid     = "interview_id",
  interview_uid = "interview_id",
  species       = "species",
  count         = "n_fish",
  catch_type    = "catch_type"
)

harlan_design
```

The design now reports `r nrow(harlan_catch)` catch rows covering
`r n_distinct(harlan_catch$species)` species, split between harvested and released
records. An advisory message notes that harvest totals in the catch table diverge
from `total_harvest` in the interviews table for some parties. That is expected:
`total_harvest` counts only harvested fish, while the catch table records both
harvested and released.

The full design object now holds every layer of the survey:

```{r}
#| label: design-slots
tibble(
  slot    = c("calendar", "counts", "interviews", "catch"),
  n_rows  = c(
    nrow(harlan_design$calendar),
    nrow(harlan_design$counts),
    nrow(harlan_design$interviews),
    nrow(harlan_design$catch)
  )
)
```

## Inspecting the design {#sec-design-inspect}

### Stratum diagnostics

`audit_strata()` computes the key stratum-level statistics that summarize how well
the sampling design is performing. For each stratum it returns the population size
(`N_h`), the realized sample size (`n_h`), the mean angler count per sampled day
aggregated across count rounds (`ybar_h`), the between-day variance of those
totals (`s2_h`), the relative standard error of the stratum mean (`RSE`), and the
design effect (`DEFF`).

```{r}
#| label: audit-strata
strata_audit <- audit_strata(harlan_design)
strata_audit$strata
```

Both strata meet the default RSE target of
`r strata_audit$rse_target`. The RSE for the weekday stratum
(`r round(strata_audit$strata$RSE[strata_audit$strata$stratum == "weekday"], 3)`)
is lower than for the weekend stratum
(`r round(strata_audit$strata$RSE[strata_audit$strata$stratum == "weekend"], 3)`),
primarily because the weekday stratum is larger and therefore has more power to
detect systematic variation.

The design effects deserve attention. The weekday DEFF of
`r round(strata_audit$strata$DEFF[strata_audit$strata$stratum == "weekday"], 2)`
indicates moderate clustering — sampled days within the weekday stratum are more
similar to one another than a simple random sample of individual counts would be.
The weekend DEFF of
`r round(strata_audit$strata$DEFF[strata_audit$strata$stratum == "weekend"], 2)`
is substantially higher. Weekend fishing activity is more variable: a sunny summer
weekend draws far more anglers than a cold autumn weekend, and this day-to-day
clustering inflates variance relative to a simple random sample of the same size.
When design effects are large, increasing the number of sampled days is more
effective than increasing the number of count rounds within already-sampled days
[@mccormick2013].

The overall design effect for the full survey is `r round(strata_audit$deff, 2)` —
close to one because the two strata have similar sizes and the weekday stratum
dominates the average.

### Coverage gaps

`check_completeness()` reports which parts of the design lack data.

```{r}
#| label: check-completeness
check_completeness(harlan_design)
```

The report shows a ✖ and flags 95 calendar days with no count data. The ✖ symbol
means the check found something that requires attention, not necessarily that
something is broken. In this case, 95 is the number of non-sampled days in the
schedule — days that were in the frame but not selected for sampling. That is
expected behavior. If the flag instead identified days that the schedule marks as
sampled but that have no count records, it would indicate a genuine data gap
requiring investigation.

The distinction matters when reading `check_completeness()` output: look at what
the flag says, not just whether a ✖ appears. A flag that reports non-sampled days
is informational. A flag that reports sampled days with missing data requires action
before estimation proceeds.

All strata exceed the `n_min = 10` interview threshold, meaning every reporting
domain has enough interviews to produce a stable estimate. When a stratum falls
below this threshold, estimation functions will still run, but the variance
estimate for that stratum will be unreliable.

### Survey calendar

`plot_design()` produces a calendar visualization of which days were sampled, how
they fall across strata, and where count data exist.

```{r}
#| label: fig-plot-design
#| fig-cap: "Survey calendar for the Harlan Reservoir 2022 creel survey. Each point represents a day in the sampling frame. Color distinguishes weekend and weekday strata. Filled points indicate sampled days."
#| fig-alt: "Calendar dot plot showing sampled and unsampled days across the 2022 season, colored by day type stratum."
plot_design(harlan_design, title = "Harlan Reservoir 2022 — survey calendar")
```

The calendar makes sampling density visible. April and early May show high sampling
coverage; late summer shows gaps where consecutive unsampled days cluster. These
patterns matter: if the unsampled gaps are not random — if they fell on weeks with
unusually poor or unusually good conditions — the stratified estimator cannot
correct for them and the estimates carry unquantified bias.

## Stratum-level sampling summary {#sec-strata-summary}

A compact view of what each stratum contributes makes it easier to catch
imbalances before moving to estimation.

```{r}
#| label: fig-strata-summary
#| fig-cap: "Sample sizes and mean angler counts per stratum for the Harlan 2022 design. Error bars show one standard error of the within-stratum mean count. Numbers above bars show the population (N_h) and sample (n_h) day counts."
#| fig-alt: "Two-panel bar chart comparing weekday and weekend strata on sample size and mean count per sampled day."

strata_tbl <- strata_audit$strata |>
  mutate(
    se_h  = sqrt(s2_h / n_h),
    label = paste0("N=", N_h, "\nn=", n_h)
  )

ggplot(strata_tbl, aes(x = stratum, y = ybar_h, fill = stratum)) +
  geom_col(width = 0.5) +
  geom_errorbar(
    aes(ymin = ybar_h - se_h, ymax = ybar_h + se_h),
    width = 0.15
  ) +
  geom_text(aes(label = label, y = 0.05), vjust = 0, size = 3) +
  scale_fill_manual(values = creel_palette(2), guide = "none") +
  labs(
    x     = "Stratum",
    y     = "Mean angler count per sampled day",
    title = "Harlan Reservoir 2022 — stratum summary"
  ) +
  theme_creel()
```

The mean count per sampled day is similar across strata —
`r round(strata_tbl$ybar_h[strata_tbl$stratum == "weekday"], 2)` for weekdays and
`r round(strata_tbl$ybar_h[strata_tbl$stratum == "weekend"], 2)` for weekends.
Similar means across strata suggest that post-stratification by day type recovers
little variance in this survey — the main value of the weekday/weekend distinction
may be as a design guarantee that both day types are represented, not as a
variance-reduction mechanism.

## Design immutability {#sec-design-immutable}

Each `add_*` function returns a new `creel_design` object rather than modifying
its argument. This design choice prevents accidental overwriting and makes it safe
to branch from any intermediate state.

```{r}
#| label: design-immutability
#| message: false
# Build a counts-only design
design_counts_only <- creel_design(
  calendar = harlan_schedule,
  date     = "date",
  strata   = "day_type"
) |>
  add_counts(counts = harlan_counts, count_time_col = "count_time")

# Build a full design branching from the same intermediate object
design_full <- add_interviews(
  design         = design_counts_only,
  interviews     = harlan_interviews_with_catch,
  catch          = "total_harvest",
  effort         = "hours_fished",
  trip_duration  = "hours_fished",
  trip_status    = "trip_status",
  n_anglers      = "n_anglers",
  angler_type    = "angler_type",
  angler_method  = "angler_method",
  interview_type = "access"
)

# design_counts_only is unchanged — interviews were added to design_full, not to it
is.null(design_counts_only$interviews)
```

Immutability means the counts-only design survives the interview-addition step
intact. This matters when the same count data feed multiple analysis branches —
for example, effort estimation over all species (counts only) alongside harvest
estimation for a target species (counts + interviews + catch). Both branches share
the same intermediate object as a starting point rather than each rebuilding from
scratch.

## The design object as the estimation handoff {#sec-design-handoff}

Every estimation function from @sec-effort-why through @sec-uncertainty-anatomy takes a `creel_design` as its
first argument. The design carries the information those functions need without
requiring the analyst to re-specify it at each call:

- Which days were sampled and in which strata (`calendar`)
- The realized count observations and their timing (`counts`)
- Trip-level effort and catch totals (`interviews`)
- Species-level catch by outcome (`catch`)
- The underlying `survey.design2` with correct strata weights

The functions that were held as `eval: false` placeholders in @sec-roving —
`estimate_catch_rate()`, `validate_incomplete_trips()`, and `summarize_trips()` —
all require a `creel_design` object of this form. The design built in this chapter
is the prerequisite for running those functions with real output.

The step-by-step build earlier in this chapter printed each intermediate state for
teaching purposes. In production code, the same pipeline is written as a single
pipe chain:

```{r}
#| label: full-pipeline
#| message: false
harlan_design_final <- creel_design(
  calendar = harlan_schedule,
  date     = "date",
  strata   = "day_type"
) |>
  add_counts(
    counts         = harlan_counts,
    count_time_col = "count_time"
  ) |>
  add_interviews(
    interviews     = harlan_interviews_with_catch,
    catch          = "total_harvest",
    effort         = "hours_fished",
    trip_duration  = "hours_fished",
    trip_status    = "trip_status",
    n_anglers      = "n_anglers",
    angler_type    = "angler_type",
    angler_method  = "angler_method",
    interview_type = "access"
  ) |>
  add_catch(
    data          = harlan_catch,
    catch_uid     = "interview_id",
    interview_uid = "interview_id",
    species       = "species",
    count         = "n_fish",
    catch_type    = "catch_type"
  )

harlan_design_final
```

The pipe chain is the production form. Each step adds one layer, and the result
is a single named object that can be passed to any estimation function in the
book.

## What the design object stores {#sec-design-slots}

The `creel_design` object is a named list. Knowing which slot holds which
information makes it possible to inspect any piece of the survey without
re-running estimation.

```{r}
#| label: design-internal
tibble(
  slot     = c("calendar", "counts", "interviews", "catch",
               "survey", "interview_survey",
               "strata_cols", "date_col", "design_type"),
  n_rows   = c(
    nrow(harlan_design_final$calendar),
    nrow(harlan_design_final$counts),
    nrow(harlan_design_final$interviews),
    nrow(harlan_design_final$catch),
    NA_integer_, NA_integer_, NA_integer_, NA_integer_, NA_integer_
  ),
  contents = c(
    "sampling schedule with sampled flag",
    "count observations from all rounds",
    "party-level interview records",
    "species-level catch records",
    "survey.design2 for count-based effort estimation",
    "survey.design2 for interview-based estimation",
    "stratum variable name(s)",
    "date column name",
    "survey design type (instantaneous, bus_route, ...)"
  )
)
```

The two `survey.design2` objects — `design$survey` and `design$interview_survey`
— are the most important internal slots. The former carries the probability
structure for estimating effort from counts; the latter carries the structure for
estimating catch rates from interviews. Estimation functions pull from the
appropriate slot depending on what they are estimating. Direct access to these
slots is rarely needed; the estimation functions handle it internally.

## When to rebuild the design {#sec-design-rebuild}

Because the design object is immutable, any change to the underlying data
requires rebuilding from the affected step forward. This keeps the design
aligned with the raw tables.

The most common triggers for a rebuild are:

**A correction to the sampling schedule.** If a day's `day_type` is corrected or
a day is added or removed from the sampled set, the base design must be rebuilt. A
partial fix — editing the schedule but passing the old design object to estimation
— will silently use the incorrect strata sizes for the N_h weights.

**A catch data correction.** If species totals in the catch table are revised after
the design was built, `add_catch()` must be called again on the updated data. The
species-level records inside the design object are a snapshot of the catch table at
the time `add_catch()` was called; they do not update automatically.

**An interview exclusion.** @sec-cleaning-inference covers data cleaning decisions that remove
or adjust specific interviews. Any such exclusion should be made on the raw
interview table before the design is built, not by operating on the design object
after the fact. The correct workflow is:

```
raw tables → clean → derive total_harvest → creel_design pipeline → estimate
```

not:

```
creel_design pipeline → estimate → discover problem → patch design object
```

The `harlan_design_final` object built in @sec-design-handoff marks the end of
the data preparation phase for the Harlan 2022 survey. When the next chapter's
cleaning decisions are applied, rebuild the design from the cleaned tables with
the same four-step pipeline.

## Takeaway {#sec-design-takeaway}

The `creel_design` object collects everything an estimator needs to produce a
design-consistent result: the sampling frame, the strata, the realized sample
counts, the party-level effort and catch, and the species-level records. Four
explicit steps keep the design choices visible and verifiable.

`audit_strata()`, `check_completeness()`, and `plot_design()` provide immediate
feedback on whether the assembled design is consistent and complete before any
estimates are computed. Problems surfaced at this stage — a stratum below the
minimum interview count, count dates outside the schedule, or a very high design
effect — are easier to diagnose here than after estimation has already run.

The next chapter addresses data cleaning decisions that can change estimates:
how to identify and handle extreme catch values, implausible trip durations, and
interviews that do not pass internal consistency checks. Cleaning is done on the
raw tables before building the design object.
