---
title: "Cleaning Data That Changes Estimates"
---

```{r}
#| label: setup-ch11
#| include: false
library(tidycreel)
library(tidyverse)
library(lubridate)
library(here)

harlan_schedule   <- readRDS(here("data/harlan_schedule.rds"))
harlan_counts     <- readRDS(here("data/harlan_counts.rds"))
harlan_interviews <- readRDS(here("data/harlan_interviews.rds"))
harlan_catch      <- readRDS(here("data/harlan_catch.rds"))

shift_hours <- c("1" = 8L, "2" = 6L)

harlan_counts_effort <- harlan_counts |>
  mutate(
    period        = as.character(period),
    total_anglers = boat_anglers + bank_anglers,
    angler_hours  = total_anglers * shift_hours[period]
  ) |>
  select(reservoir, date, day_type, period, section, count_time, angler_hours)

harvest_by_interview <- harlan_catch |>
  filter(catch_type == "harvested") |>
  group_by(interview_id) |>
  summarise(total_harvest = sum(n_fish), .groups = "drop")

harlan_interviews_with_catch <- harlan_interviews |>
  left_join(harvest_by_interview, by = "interview_id") |>
  mutate(total_harvest = if_else(is.na(total_harvest), 0L, as.integer(total_harvest)))

build_design <- function(interviews, catch_data) {
  suppressWarnings(suppressMessages(
    creel_design(calendar = harlan_schedule, date = "date", strata = "day_type") |>
      add_counts(harlan_counts_effort, count_time_col = "count_time") |>
      add_interviews(
        interviews     = interviews,
        catch          = "total_harvest",
        effort         = "hours_fished",
        trip_duration  = "hours_fished",
        trip_status    = "trip_status",
        n_anglers      = "n_anglers",
        angler_type    = "angler_type",
        angler_method  = "angler_method",
        interview_type = "access"
      ) |>
      add_catch(
        data          = catch_data,
        catch_uid     = "interview_id",
        interview_uid = "interview_id",
        species       = "species",
        count         = "n_fish",
        catch_type    = "catch_type"
      )
  ))
}

harlan_design_ch11 <- build_design(harlan_interviews_with_catch, harlan_catch)

# Ch13-style harvest design (needed for estimate_total_harvest in what-if)
prep_harvest_interviews <- function(catch_df) {
  catch_pivot <- catch_df |>
    group_by(interview_id, catch_type) |>
    summarise(n_fish = sum(n_fish), .groups = "drop") |>
    pivot_wider(names_from = catch_type, values_from = n_fish, values_fill = 0L) |>
    mutate(total_catch = harvested + released)
  harlan_interviews |>
    left_join(catch_pivot, by = "interview_id") |>
    mutate(
      harvested   = if_else(is.na(harvested),   0L, as.integer(harvested)),
      released    = if_else(is.na(released),    0L, as.integer(released)),
      total_catch = if_else(is.na(total_catch), 0L, as.integer(total_catch))
    )
}

build_design_harvest <- function(catch_df) {
  ints <- prep_harvest_interviews(catch_df)
  suppressWarnings(suppressMessages(
    creel_design(calendar = harlan_schedule, date = "date", strata = "day_type") |>
      add_counts(harlan_counts_effort, count_time_col = "count_time") |>
      add_interviews(
        interviews     = ints,
        catch          = "total_catch",
        harvest        = "harvested",
        effort         = "hours_fished",
        trip_duration  = "hours_fished",
        trip_status    = "trip_status",
        n_anglers      = "n_anglers",
        angler_type    = "angler_type",
        angler_method  = "angler_method",
        interview_type = "access"
      ) |>
      add_catch(
        data          = catch_df,
        catch_uid     = "interview_id",
        interview_uid = "interview_id",
        species       = "species",
        count         = "n_fish",
        catch_type    = "catch_type"
      )
  ))
}

harlan_design_ch11_harvest <- build_design_harvest(harlan_catch)
```

## Cleaning as part of inference {#sec-cleaning-inference}

Every creel analysis involves cleaning. Species names get entered inconsistently.
Trip durations sometimes fall outside what is physically possible for a survey
shift. Interview records occasionally contain party sizes that conflict with count
data. Catch entries arrive with values that seem impossibly large.

None of those issues resolves itself. The analyst must decide whether to
remove, recode, or retain each record, and those decisions are not neutral.
Changing which records enter the estimator changes the numerator, the
denominator, or both. Cleaning decisions are methodological decisions. They
belong in the analysis record alongside model, stratification, and estimator
choices.

This chapter works through the most consequential cleaning steps for a creel
dataset: species standardization, trip duration plausibility, catch quantity
review, and trip status verification. For each step, the focus is not just on
how to perform the check but on how the choice propagates through to the final
estimate.

The Harlan Reservoir 2022 dataset is clean by real-world standards. It was
collected under a quality-controlled protocol and reviewed before archiving. That
makes it a realistic starting point: it will show the kinds of issues that
survive pre-analysis review, not synthetic data constructed to contain errors.

## Species names and standardization {#sec-species-codes}

### Why species names matter for estimation

Species catch data in creel programs often arrives in free-text form. Interview
clerks enter what anglers say, and anglers use common names that vary by region,
familiarity, and culture. "Drum" and "sheepshead" describe the same fish in
different places. "Crappie" may refer to Black Crappie, White Crappie, or an
undifferentiated pooled entry.

Those naming differences are minor within one survey running on consistent
entry rules. They matter when surveys are compared across time, merged into
multi-site databases, or linked to external data sources (bag limit tables,
stock assessment databases, CreelCat). A merge that pairs "White Bass" with
"WHB" will succeed; a merge that pairs "Whitebass" with "WHB" will silently
fail, and those records will appear as missing rather than mismatched.

The American Fisheries Society publishes standard species codes for exactly this
reason. `tidycreel` provides `standardize_species()`, which maps free-text
species names to AFS codes and identifies entries that do not match any known
code.

### Applying standardization to Harlan catch data

The Harlan catch table contains 11 species entered as free-text common names.
Most of these map cleanly to AFS codes:

```{r}
#| label: species-standardize
#| echo: true
standardize_species(harlan_catch) |>
  select(species, species_code) |>
  distinct() |>
  arrange(is.na(species_code), species)
```

Nine of eleven species resolve to AFS codes. Two do not: Wiper and Crappie.
(Freshwater Drum, which lacked an AFS code in earlier versions, now resolves
automatically to `"FRD"` in `tidycreel` ≥ 2.2.0.)

Wiper is a Striped Bass × White Bass hybrid that varies in naming by region and
lacks a single-code standard. Crappie entered without species designation is a
data quality issue: Black Crappie and White Crappie are distinct management
units in many systems, and pooled entries cannot be split after the fact.

These are different problems. Wiper needs a project-specific code assignment.
Crappie needs a conversation with field staff about whether species-level
identification was attempted and, if so, how to recover it from supplemental
records.

```{r}
#| label: fig-species-codes
#| fig-cap: "Species in the Harlan Reservoir 2022 catch records by AFS code match status. Matched species resolve to standard codes and are ready for cross-survey comparison. Unmatched species require manual resolution before merging with external data."
#| fig-alt: "Horizontal bar chart showing total catch records for each of 11 species, colored by whether the species matched an AFS code. White Bass, Walleye, and Freshwater Drum have the most records. Two species (Wiper and Crappie) are shown in a different color indicating no AFS match."
#| echo: false
standardize_species(harlan_catch) |>
  mutate(
    n_records  = n(),
    code_status = if_else(!is.na(species_code), "Matched", "No AFS match"),
    species = fct_reorder(species, n_records, .fun = function(s) sum(s))
  ) |>
  count(species, code_status) |>
  ggplot(aes(x = n, y = fct_reorder(species, n), fill = code_status)) +
  geom_col() +
  scale_fill_manual(
    values = c("Matched" = "#2E86AB", "No AFS match" = "#E84855"),
    name   = "Code status"
  ) +
  labs(
    x = "Catch records",
    y = NULL
  ) +
  theme_creel()
```

### What to do when codes are missing

When `standardize_species()` returns `NA` for a species, the analyst has three
options, and the choice should be documented:

1. **Assign a project code** for species that have no universal standard but
   can be coded consistently (Wiper). Pass a named character vector via the
   `custom_codes` argument — `standardize_species(data, custom_codes = c("Wiper" = "WPR"))` —
   or apply a `case_when()` override to the `species_code` column after the
   function runs.
2. **Create a pooled category** for entries that mix species (Crappie without
   species designation). A project-specific label like `"CRP-POOL"` flags the
   record as unresolved without discarding the catch count.
3. **Flag as unresolved and exclude from species totals** if the entry cannot be
   reliably coded without returning to field records. Report species totals only
   for rows where `!is.na(species_code)`.

A `case_when()` override handles the two remaining unresolved species after
`standardize_species()` runs. For single-species overrides, `custom_codes =
c("Wiper" = "WPR")` is equivalent; `case_when()` is used here because `Crappie`
requires a pooled label (`"CRP-POOL"`) that warrants explicit documentation in
the code:

```{r}
#| label: species-code-override
#| echo: true
harlan_catch_coded <- harlan_catch |>
  standardize_species() |>
  mutate(species_code = case_when(
    species == "Wiper"   ~ "WPR",      # project code; no AFS standard
    species == "Crappie" ~ "CRP-POOL", # pooled: species unknown
    TRUE                 ~ species_code
  ))

harlan_catch_coded |>
  select(species, species_code) |>
  distinct() |>
  arrange(is.na(species_code), species)
```

After the override, all 11 species in the Harlan dataset have a code. (Nine resolved automatically via the AFS lookup; two required the `case_when()` step.) The
`"CRP-POOL"` label is a signal to downstream reporting code that this entry
should not be summed with species-level Crappie totals from other sources.
Any consistent, documented label works; what matters is that the convention is
applied identically in every reporting cycle.

## Referential integrity between tables {#sec-referential-integrity}

A creel dataset is a set of linked tables: a schedule table, an interview table,
and a catch table. Cleaning a single table without checking the others can leave
the dataset in an inconsistent state. The most common failure modes are:

- **Orphan catch records** — catch rows whose `interview_id` does not match any
  interview. These inflate species totals without contributing effort or trip
  information.
- **Zero-catch interviews** — interview rows with no matching catch records.
  These are usually valid (skunked anglers), but a large count may signal a data
  entry gap.

A pair of `anti_join()` calls checks both conditions:

```{r}
#| label: referential-integrity
#| echo: true
# Catch records with no parent interview
anti_join(harlan_catch, harlan_interviews, by = "interview_id")

# Interviews with no catch records at all
anti_join(harlan_interviews, harlan_catch, by = "interview_id") |>
  count(trip_status)
```

The Harlan data passes both checks. The 158 interviews with no catch records are
not a data gap: 56 are completed trips and 102 are incomplete trips where the
angler reported zero fish caught. Zero-catch completed trips are valid
observations and must remain in the analysis.

The second check is worth running any time the catch table and interview table
arrive from different sources — for example, when interview data is collected
in a mobile app and catch data is added later from a separate entry screen. In
that workflow, late-arriving catch records that reference deleted interview IDs
would appear as orphans here.

When an orphan catch record is found, the response depends on whether the
interview can be recovered. If the interview was not entered, the catch record
should be removed. If the interview was entered under a different ID (a key
mismatch, common in manual entry), the catch record should be relinked. Either
way, the decision should be logged.

## Trip duration: plausibility and outlier review {#sec-trip-duration}

### Why hours_fished is sensitive

The `hours_fished` field is the effort denominator for catch-per-unit-effort
estimates. Its mean value among complete trips drives the CPUE calculation
directly: CPUE = (total catch) / (total hours fished). An extreme value in
`hours_fished` inflates the denominator, deflates the CPUE, and, if the
associated catch is high, distorts both the effort and catch components.

Access-point surveys at Harlan are scheduled in two periods: morning (period 1,
8-hour window) and afternoon/evening (period 2, 6-hour window). A trip duration
exceeding the scheduled window for that period is physically impossible if the
angler's entire trip occurred within the survey shift. Durations approaching the
shift ceiling are plausible but warrant review.

### Identifying outliers with flag_outliers()

`flag_outliers()` implements Tukey's IQR fence method. For a numeric column with
first quartile $Q_1$ and third quartile $Q_3$:

$$\text{fence\_high} = Q_3 + k \times \text{IQR}$$

Values above the upper fence are flagged as outliers. The multiplier $k$
controls sensitivity. Tukey's conventional choices are $k = 1.5$ for mild
outliers and $k = 3.0$ for extreme outliers.

```{r}
#| label: flag-hours-k15
#| echo: true
flag_outliers(harlan_interviews, hours_fished, k = 1.5) |>
  filter(is_outlier) |>
  select(interview_id, date, day_type, period, trip_status,
         hours_fished, fence_high, outlier_reason)
```

At $k = 1.5$, one interview is flagged: interview 388, recorded on 14 July 2022,
with `hours_fished = 13.82`. The upper fence at $k = 1.5$ is
$Q_3 + 1.5 \times \text{IQR} = 5.53 + 1.5 \times 4.42 = 12.17$ hours. Interview
388 exceeds that fence.

```{r}
#| label: flag-hours-k30
#| echo: true
flag_outliers(harlan_interviews, hours_fished, k = 3.0) |>
  filter(is_outlier) |>
  select(interview_id, date, day_type, trip_status, hours_fished,
         fence_high, outlier_reason)
```

At $k = 3.0$, the upper fence rises to
$5.53 + 3.0 \times 4.42 = 18.80$ hours. No records are flagged. A 13.82-hour
trip falls within the extreme-outlier fence.

The choice between $k = 1.5$ and $k = 3.0$ is not mechanical. For a survey
with 14-hour interview windows, a 13.82-hour trip is unusual but physically
possible. Whether to treat it as an error depends on whether the survey period
for that record permitted a trip that long. The right question is not whether the
value is statistically unusual but whether it is substantively implausible given
the field conditions.

In this case, interview 388 is classified as a completed trip on a weekday in
period 2 (the 6-hour window). A completed trip of 13.82 hours in a 6-hour window
is impossible — the angler could not have completed a 13.82-hour trip within the
survey period if the period only ran 6 hours. That physical constraint justifies
flagging and reviewing, regardless of which $k$ value the IQR fence produced.

### Visualizing duration against catch

A univariate view of `hours_fished` shows where the fence falls. A bivariate
view — hours fished against total fish caught — reveals whether a flagged record
is unusual in one dimension only or in both simultaneously.

```{r}
#| label: fig-hours-catch-scatter
#| fig-cap: "Hours fished versus total fish caught (harvest + release) per interview at Harlan Reservoir 2022. Each point is one interview, colored by trip status. Interview 388 (13.82 hr, 49 fish) sits at the far right — the longest trip in the dataset. The dashed vertical line marks the k = 1.5 IQR outlier fence for hours_fished (12.17 hr). Several other complete trips show high catch counts (up to 189 fish) with shorter durations, indicating that duration and catch quantity are weakly correlated at Harlan."
#| fig-alt: "Scatter plot with hours fished on the x-axis (0 to 14) and total fish caught on the y-axis (0 to 200). Points are colored by trip status (complete vs incomplete). Most points cluster below 8 hours and below 50 fish. Interview 388 is the rightmost point at 13.82 hours and 49 fish, labeled and highlighted. A dashed vertical line at 12.17 hours marks the k=1.5 fence. Several other unlabeled points show catch counts above 100 fish with shorter durations."
#| echo: false
total_catch_per_int <- harlan_catch |>
  group_by(interview_id) |>
  summarise(total_catch = sum(n_fish), .groups = "drop")

scatter_data <- harlan_interviews |>
  left_join(total_catch_per_int, by = "interview_id") |>
  mutate(
    total_catch = replace_na(total_catch, 0L),
    highlight   = interview_id == 388
  )

fence_15 <- unique(flag_outliers(harlan_interviews, hours_fished, k = 1.5)$fence_high)

ggplot(scatter_data, aes(x = hours_fished, y = total_catch,
                         color = trip_status, shape = trip_status)) +
  geom_point(data = filter(scatter_data, !highlight),
             alpha = 0.55, size = 1.8) +
  geom_point(data = filter(scatter_data, highlight),
             color = "#E84855", size = 4, shape = 18) +
  geom_vline(xintercept = fence_15, color = "#E84855", linetype = "dashed",
             linewidth = 0.8) +
  annotate("text", x = fence_15 - 0.3, y = Inf,
           label = "k = 1.5 fence\n(12.17 hr)", hjust = 1, vjust = 1.3,
           color = "#E84855", size = 2.8) +
  annotate("text", x = 13.82, y = 55, label = "Interview 388",
           hjust = 0.5, vjust = -0.3, color = "#E84855", size = 2.8) +
  scale_color_manual(
    values = c(complete = "#2E86AB", incomplete = "#A8DADC"),
    name   = "Trip status"
  ) +
  scale_shape_manual(
    values = c(complete = 16, incomplete = 17),
    name   = "Trip status"
  ) +
  labs(x = "Hours fished", y = "Total fish (harvest + release)") +
  theme_creel()
```

Interview 388 is the rightmost point: the longest trip in the dataset at 13.82
hours with a moderate catch of 49 fish. It is unusual for its duration, not for
its catch count. Trips with far higher catch totals (up to 189 fish) appear
throughout the dataset with much shorter durations, demonstrating that hours
fished and total catch are only weakly correlated at Harlan. That weak
correlation is itself informative: it means no single threshold on either
variable alone is sufficient to identify implausible combinations. A trip with
low duration and unusually high catch (for example, 189 fish in 1.92 hours,
visible in the upper-left region of the plot) deserves the same scrutiny as
a trip with unusually long duration.

### What to do with flagged duration records

The appropriate response to a flagged trip duration depends on the source of the
anomaly:

- **Transcription error** — if field sheets show a different value, correct to
  the field record and document the correction.
- **Plausible but unusual** — if the value is consistent with the field record,
  retain the record and note in the analysis log that one complete trip exceeded
  the $k = 1.5$ fence. No correction is needed.
- **Inconsistent with survey design** — if the duration is impossible given the
  survey period (as in the period-2 case above), reclassify the trip as
  incomplete, truncate `hours_fished` to the period boundary, or remove the
  record. Each option changes the estimate differently.

The conservative default is to flag, document, and retain — then report
sensitivity to the analyst's choice (see @sec-estimation-impact).

## Catch quantities: large bags and zero-catch trips {#sec-catch-quantities}

### Flag_outliers on fish counts

The same IQR fence logic applies to fish counts. Large bag totals can arise from
two distinct causes: data entry error (clerk transcribed "32" instead of "2") or
genuine large catch (angler found a schooling White Bass aggregation and limited
out on multiple species). The fence identifies which records are statistically
unusual; field knowledge determines which are substantively suspicious.

```{r}
#| label: flag-nfish
#| echo: true
flag_outliers(harlan_catch, n_fish, k = 1.5) |>
  filter(is_outlier) |>
  count(species) |>
  arrange(desc(n))
```

At $k = 1.5$, 99 of 987 catch records exceed the fence at 21 fish. The
distribution is dominated by White Bass and Walleye. White Bass at Harlan school
aggressively during spring and fall; bag limits in Nebraska permit daily limits
that can produce interview-level totals well above 21. Walleye with high counts
may reflect multi-angler parties (the `n_anglers` field records the party, not
the individual, at Harlan) or productive early-season trips.

These records are not errors in the conventional sense. They are real, they are
high, and they are the result of field conditions that the survey is designed to
capture. Removing them would bias harvest estimates downward by systematically
excluding high-catch parties.

### Zero-catch and zero-effort records

Zero catches are common and valid. An angler who fished all morning and caught
nothing represents a real data point: zero catch, positive effort. That zero
enters the CPUE denominator appropriately.

What requires attention is zero effort on a completed trip. A record with
`trip_status == "complete"` and `hours_fished == 0` is internally inconsistent:
a completed trip must have positive effort. The Harlan data contains no such
records:

```{r}
#| label: zero-effort-check
#| echo: true
harlan_interviews |>
  filter(hours_fished == 0 | (trip_status == "complete" & hours_fished < 0.1))
```

Zero-catch records, on the other hand, are present and expected. Interviews with
no associated catch records in `harlan_catch` represent skunked parties — valid
observations that enter the estimator with catch = 0.

### Context-dependent review

Flagging catch counts at $k = 1.5$ casts a wide net. A more defensible approach
pairs the flag with domain knowledge:

1. Compare flagged values against legal bag limits for each species. A White Bass
   count of 32 is plausible (Nebraska has no closed season and a liberal bag
   limit). A Walleye count of 49 in a single interview warrants a second look at
   field sheets.
2. Check whether the party size (`n_anglers`) is large enough to explain the
   count. An interview with `n_anglers = 4` and 32 White Bass is 8 fish per
   person — straightforward. An interview with `n_anglers = 1` and 49 Walleye
   is 49 individual fish — warranting field record review.

The Harlan catch record with the highest outlier flag is 49 Walleye released by
a single party (interview 388 — the same record flagged for unusual trip
duration). The combination of an implausible trip duration and an unusually large
Walleye release count in the same interview strengthens the case for field record
review.

## Party size and angler counts {#sec-party-size}

The `n_anglers` field records the number of anglers in the interviewed party. At
Harlan, parties range from 1 to 9:

```{r}
#| label: nanglers-distribution
#| echo: true
harlan_interviews |>
  count(n_anglers) |>
  mutate(pct = round(n / sum(n) * 100, 1))
```

Most parties are 1–2 anglers (75.5%). Parties of 7–9 exist but are rare (0.8%
of interviews). At $k = 1.5$, `flag_outliers()` sets the fence at 2 anglers,
flagging all parties of 3 or more — nearly half the dataset. This illustrates
why $k$ must be calibrated to the variable's natural distribution: `n_anglers`
is right-skewed but parties of 3 or more are not errors. For discrete count
variables with low expected maxima, domain-specific absolute bounds are more
useful than IQR fences.

At Harlan, a reasonable absolute upper bound is 8 anglers per interview — a
figure consistent with a typical boat capacity plus family bank fishing. A
single record exceeds that bound:

```{r}
#| label: nanglers-absolute-bound
#| echo: true
harlan_interviews |>
  filter(n_anglers > 8) |>
  select(interview_id, date, angler_type, n_anglers, hours_fished, trip_status)
```

Interview 273 recorded 9 anglers on a weekday in May. Nine anglers is
physically plausible for a group outing but worth confirming against the field
sheet — both because party size affects the angler-effort denominator and
because the interview represents 9 angler-trips in the catch rate calculation.
One misrecorded party of 9 that should have been 2 would contribute 4.5× the
weight it deserves.

A more reliable check is joint consistency between party size and catch. A
party of 2 anglers reporting 6 Walleye harvested is plausible. A party of 2
reporting 90 Walleye harvested is not. That joint check requires pairing the
interview table with the catch table:

```{r}
#| label: party-catch-consistency
#| echo: true
harlan_interviews |>
  left_join(
    harlan_catch |>
      filter(catch_type == "harvested") |>
      group_by(interview_id) |>
      summarise(total_harvest = sum(n_fish), .groups = "drop"),
    by = "interview_id"
  ) |>
  mutate(
    total_harvest      = replace_na(total_harvest, 0),
    harvest_per_angler = total_harvest / n_anglers
  ) |>
  filter(harvest_per_angler > 15) |>
  select(interview_id, date, n_anglers, total_harvest, harvest_per_angler,
         trip_status, hours_fished)
```

Any interview with more than 15 harvested fish per angler — a threshold that
exceeds legal bag limits for most target species in this system — surfaces as a
candidate for field record review.

## Trip status as a cleaning variable {#sec-trip-status}

Trip status (complete versus incomplete) is not just a classification field. It
determines which records enter which estimator:

- Complete trips contribute to CPUE estimation (ratio-of-means uses
  completed trips only; see @sec-effort-formula and @sec-cpue-estimators).
- Incomplete trips are used by some estimators to validate that complete
  and incomplete anglers are exchangeable (see `validate_incomplete_trips()` in
  @sec-roving-creel-surveys).

A misclassified trip status is therefore a cleaning decision with a direct
pathway into the estimate. An incomplete trip recorded as complete inflates the
hours-fished denominator of CPUE (incomplete trips are shorter on average). A
complete trip recorded as incomplete removes a valid observation from the ratio
estimator.

The Harlan data shows the expected pattern: complete trips are longer than
incomplete trips, with little overlap in their medians:

```{r}
#| label: summarize-trips-full
#| echo: true
summarize_trips(harlan_design_ch11)
```

The complete trip mean is 4.74 hours; the incomplete trip mean is 1.71 hours.
The ratio (2.77) is consistent with access-point survey literature where
incomplete trips typically represent early-arriving anglers intercepted before
they finish. If those means were reversed — or if incomplete trip durations
clustered near the shift ceiling — misclassification would be a prime suspect.

`validate_incomplete_trips()` tests whether the two groups are
statistically exchangeable for catch-rate estimation using two one-sided tests
(TOST) within a specified equivalence margin:

```{r}
#| label: validate-trips-ch11
#| echo: true
validate_incomplete_trips(
  harlan_design_ch11,
  catch      = total_harvest,
  effort     = hours_fished,
  truncate_at = 0.5
)
```

The test fails. Incomplete trips at Harlan catch at a meaningfully higher rate
than complete trips (0.792 vs. 0.585 fish/hr), and the difference falls outside
the ±20% equivalence margin. The TOST failure does not indicate that trip status
was misclassified — it indicates that incomplete anglers systematically differ
from complete anglers, a common pattern at access-point surveys where early-
morning arrivals target faster-biting periods. Keep the groups separate. Any
record misclassified as "complete" that truly represents a shorter,
higher-intensity trip would inflate the complete-trip catch rate and bias CPUE
estimates upward.

## Building a cleaning log {#sec-cleaning-log}

Any cleaning step that changes a record's contribution to an estimate should be
logged before it is applied. A minimal cleaning log has four columns per
changed record: the record identifier, the field affected, the original value,
and the new value. A fifth column should capture the rule applied.

The dplyr approach keeps the raw data intact and stores changes as a separate
audit table:

```{r}
#| label: cleaning-log
#| echo: true
cleaning_log <- harlan_interviews |>
  left_join(
    flag_outliers(harlan_interviews, hours_fished, k = 1.5) |>
      select(interview_id, is_outlier, fence_high),
    by = "interview_id"
  ) |>
  filter(is_outlier) |>
  mutate(
    field         = "hours_fished",
    original_value = hours_fished,
    proposed_value = fence_high,
    rule           = "IQR k=1.5: hours_fished exceeds upper fence; review against field record",
    action         = "REVIEW — do not apply automatically"
  ) |>
  select(interview_id, date, day_type, period, trip_status,
         field, original_value, proposed_value, rule, action)

cleaning_log
```

The key discipline is the `action` column. A cleaning log is not a deletion
list. It is a record of anomalies and proposed responses. The final decision —
accept, correct, or remove — should be recorded before the analysis runs, not
reconstructed after the fact.

For the one flagged duration record (interview 388), the proposed action is field
record review. Until that review is complete, the defensible choice is to run
the analysis twice: once with the record retained and once with it excluded,
and report the sensitivity.

### Applying a cleaning decision in code

Once a cleaning decision is made, the implementation follows one of three
patterns depending on the action taken. All three keep the raw data unchanged
and produce a derived cleaned dataset:

```{r}
#| label: cleaning-apply
#| echo: true
# Option 1: truncate hours_fished to the period boundary (6 hr for period 2)
interviews_truncated <- harlan_interviews |>
  mutate(
    hours_fished = case_when(
      interview_id == 388 ~ 6.0,   # period 2 maximum
      TRUE                ~ hours_fished
    )
  )

# Option 2: remove the record entirely
interviews_removed <- harlan_interviews |>
  filter(interview_id != 388)

# Option 3: reclassify trip status to incomplete (then drop from CPUE)
interviews_reclassified <- harlan_interviews |>
  mutate(
    trip_status = if_else(interview_id == 388, "incomplete", trip_status)
  )

# Show effect on mean complete trip duration
list(
  original      = harlan_interviews,
  truncated     = interviews_truncated,
  removed       = interviews_removed,
  reclassified  = interviews_reclassified
) |>
  map_dfr(
    ~ filter(.x, trip_status == "complete") |>
        summarise(n = n(), mean_hr = mean(hours_fished), max_hr = max(hours_fished)),
    .id = "approach"
  )
```

Each option shifts the mean complete trip duration by a different amount.
Truncation to 6.0 hours reduces the mean most aggressively because 6.0 is
well below interview 388's recorded value. Removal reduces it more modestly
because it eliminates one observation. Reclassification to incomplete removes
the record from the complete-trip pool entirely, producing the same result as
removal for CPUE estimation. Which option matches the field record determines
which is correct — not statistical convenience.

## The estimation impact {#sec-estimation-impact}

### How a single record propagates

Interview 388 (13.82 hours, complete, 49 Walleye released, 0 harvested) enters
the analysis in two places: the trip duration distribution used to compute mean
hours_fished among complete trips, and the catch table as 49 released Walleye.
Removing it produces two changes:

**Trip duration summary** — the mean complete trip duration shifts from 4.74 to
4.72 hours (removing 0.02 hours from the mean of 390 complete trips):

```{r}
#| label: trips-before-after
#| echo: true
# Design without interview 388
design_clean <- build_design(
  filter(harlan_interviews_with_catch, interview_id != 388),
  filter(harlan_catch, interview_id != 388)
)

# Compare trip summaries
cat("--- Full dataset ---\n")
summarize_trips(harlan_design_ch11)
cat("\n--- Interview 388 removed ---\n")
summarize_trips(design_clean)
```

**Catch rate (CPUE)** — removing the 13.82-hour zero-harvest trip reduces the
hours-fished denominator slightly, and CPUE rises from 0.585 to 0.589 fish/hr:

```{r}
#| label: cpue-before-after
#| echo: true
cpue_full  <- estimate_catch_rate(harlan_design_ch11)$estimates
cpue_clean <- estimate_catch_rate(design_clean)$estimates

bind_rows(
  cpue_full  |> mutate(dataset = "Full (n = 390 complete trips)"),
  cpue_clean |> mutate(dataset = "Interview 388 removed (n = 389)")
) |>
  select(dataset, estimate, se, ci_lower, ci_upper, n)
```

The absolute difference is 0.004 fish/hr — small relative to the 95%
confidence interval width of 0.18 fish/hr. In this case, retaining or removing
interview 388 does not materially change the conclusion. That result should be
reported explicitly: "Sensitivity analysis: removing the one trip duration
outlier changed the CPUE estimate by 0.004 fish/hr (0.7%), within the
uncertainty of the estimate."

### What if the catch classification were wrong?

The CPUE analysis above tests the impact of removing a record. A different
class of cleaning error — a catch-type misclassification — has a different
propagation path. Consider the question: what if interview 388 had the
49 Walleye entered as *harvested* rather than *released*, and the
misclassification went undetected?

That scenario does not change the hours-fished denominator. It does change the
harvest numerator. The effect is isolated to `estimate_total_harvest()`:

```{r}
#| label: whatif-harvest
#| echo: true
# Hypothetical: 49 released Walleye entered as harvested (catch_type error)
catch_whatif <- harlan_catch |>
  mutate(
    catch_type = if_else(
      interview_id == 388 & species == "Walleye",
      "harvested",
      catch_type
    )
  )

design_whatif <- build_design_harvest(catch_whatif)

bind_rows(
  estimate_total_harvest(harlan_design_ch11_harvest)$estimates |>
    mutate(scenario = "Actual (49 Walleye released)"),
  estimate_total_harvest(design_whatif)$estimates |>
    mutate(scenario = "What-if (49 Walleye entered as harvested)")
) |>
  select(scenario, estimate, se, ci_lower, ci_upper)
```

Reclassifying 49 released Walleye as harvested in one interview adds
approximately 177 fish to the season harvest estimate — a 1.6% upward bias.
The reason the expansion factor is roughly 3.6× (not 1.0×) is that the
stratified estimator treats this interview's harvest rate as representative
of other complete trips in the same stratum. One high-catch record inflates
the stratum-level mean, which then multiplies against the total estimated
effort.

Whether 177 fish is material depends on the management context. For a fishery
with a Walleye quota or a slot-limit assessment, a 1.6% harvest overestimate
matters. For a general community harvest summary, it may be within acceptable
reporting uncertainty. The point is not the magnitude in isolation. The
analyst should compute it and make the judgment explicitly rather than assume
catch-type errors are harmless.

### When the impact is not small

The Harlan example is clean enough that one outlier has minimal effect. In
programs with smaller sample sizes (50–80 interviews per stratum), a single
anomalous record can move a CPUE estimate by 5–10%. The same analysis — run
twice, compare the difference — is still the right approach. The magnitude of
the sensitivity result determines whether the record requires resolution before
the report is finalized.

A cleaning decision that shifts the estimate by more than the reported
uncertainty (i.e., by more than the half-width of the confidence interval) is
a decision that should be documented in the methods section, not buried in a
data appendix.

### Effort estimates are less sensitive

The total effort estimate (angler-hours) comes from instantaneous count
expansion, not from `hours_fished` in the interview table. Removing interview
388 does not change the count-based effort estimate:

```{r}
#| label: effort-before-after
#| echo: true
bind_rows(
  estimate_effort(harlan_design_ch11, target = "stratum_total")$estimates |>
    mutate(dataset = "Full"),
  estimate_effort(design_clean, target = "stratum_total")$estimates |>
    mutate(dataset = "Interview 388 removed")
) |>
  select(dataset, estimate, se, ci_lower, ci_upper)
```

Both designs produce the same 33,052 angler-hour estimate. This illustrates an
important structural feature of the access-point estimator: effort comes from
counts, not from interviews. Cleaning decisions that affect the interview table
propagate into catch-rate estimates (CPUE, total harvest) but not into the
count-based effort expansion. The two components should be diagnosed separately.

## A conservative cleaning rule {#sec-conservative-cleaning}

The practical guidance from this chapter reduces to four steps:

1. **Flag before removing.** Use `flag_outliers()` and `standardize_species()`
   to identify anomalies. Do not delete records at the flagging step.

2. **Return to field records for flagged entries.** A statistical flag is a
   referral, not a verdict. Check the original data sheet before altering any
   value.

3. **Log every change.** Keep the original value, the corrected value, the
   rule applied, and the interview ID in a cleaning log that travels with the
   analysis.

4. **Report sensitivity.** Run the analysis with and without each cleaning
   decision that materially affects an estimate. If the difference is within the
   confidence interval, say so. If it is not, the decision requires explicit
   justification in the methods section.

These steps are not additional overhead. They are the documentation that makes
an estimate reproducible — by the same analyst next season, by a reviewer, or
by a manager who needs to understand why this year's estimate differs from
last year's.

## Takeaway {#sec-cleaning-takeaway}

Data cleaning in creel surveys is part of estimation. Decisions about which
records to accept, recode, or remove propagate through catch rates, harvest
totals, and the uncertainty reported around them. The species name an angler
gives in the field, the trip duration a clerk records, and the catch count that
enters the database are all measurements, and they benefit from systematic
review before being embedded in an estimate.

`tidycreel` provides `flag_outliers()` for detecting unusual numeric values and
`standardize_species()` for aligning species names to a common code system.
Neither function makes the decision for the analyst. Both functions make the
anomalies visible, document the detection rule, and create a foundation for the
cleaning log that should accompany every creel analysis.

The Harlan 2022 data required minimal cleaning. One trip duration flagged at
$k = 1.5$; three species names lacked AFS codes; no zero-effort completed
trips; no impossible catch counts given the party sizes observed. Those results
matter too: a dataset that passes systematic review gives the analyst
confidence that the estimates reflect the fishery, not the data entry process.
