---
title: "Effort Estimation"
---

```{r}
#| label: setup-ch12
#| include: false
library(tidycreel)
library(tidyverse)
library(lubridate)
library(here)

harlan_schedule   <- readRDS(here("data/harlan_schedule.rds"))
harlan_counts     <- readRDS(here("data/harlan_counts.rds"))
harlan_interviews <- readRDS(here("data/harlan_interviews.rds"))
harlan_catch      <- readRDS(here("data/harlan_catch.rds"))

# Shift lengths (hours) for each period at Harlan
shift_hours <- c("1" = 8L, "2" = 6L)

# Pre-compute angler_hours = total_anglers * shift_length
# and convert period to character so it is not mistaken for a count variable
harlan_counts_effort <- harlan_counts |>
  mutate(
    period        = as.character(period),
    total_anglers = boat_anglers + bank_anglers,
    angler_hours  = total_anglers * shift_hours[period]
  ) |>
  select(reservoir, date, day_type, period, section, count_time, angler_hours)

# Derive total_harvest per interview for design pipeline
harvest_by_interview <- harlan_catch |>
  filter(catch_type == "harvested") |>
  group_by(interview_id) |>
  summarise(total_harvest = sum(n_fish), .groups = "drop")

harlan_interviews_with_catch <- harlan_interviews |>
  left_join(harvest_by_interview, by = "interview_id") |>
  mutate(total_harvest = if_else(is.na(total_harvest), 0L, as.integer(total_harvest)))

# Full design with effort-preprocessed counts
suppressMessages({
  harlan_design_effort <- creel_design(
    calendar = harlan_schedule,
    date     = "date",
    strata   = "day_type"
  ) |>
    add_counts(harlan_counts_effort, count_time_col = "count_time") |>
    add_interviews(
      interviews     = harlan_interviews_with_catch,
      catch          = "total_harvest",
      effort         = "hours_fished",
      trip_duration  = "hours_fished",
      trip_status    = "trip_status",
      n_anglers      = "n_anglers",
      angler_type    = "angler_type",
      angler_method  = "angler_method",
      interview_type = "access"
    ) |>
    add_catch(
      data          = harlan_catch,
      catch_uid     = "interview_id",
      interview_uid = "interview_id",
      species       = "species",
      count         = "n_fish",
      catch_type    = "catch_type"
    )
})
```

## Why effort comes first {#sec-effort-why}

Every estimate produced by a creel survey is either effort itself or a product
that uses effort as a scaling factor. If effort is wrong, every downstream
number built on it is wrong too.

Effort estimation therefore sits at the foundation of the creel workflow. The
count schedule — which days to sample, how many periods per day, how many
instantaneous counts per period — is determined by the precision requirements
for effort, not by the requirements for catch rate or harvest. Getting effort
right means understanding what the instantaneous count estimator targets, how
stratification turns counts into a season total, and where variance comes from.

This chapter works through effort estimation for the Harlan Reservoir 2022
survey: the formula that converts an instantaneous angler count into an
angler-hour total, the stratified expansion that converts sampled days into a
full-season estimate, the variance decomposition that separates between-day
uncertainty from within-day uncertainty, and the `estimate_effort()` function
in `tidycreel` that assembles all of these pieces in a single call.

## What effort means in a creel survey {#sec-effort-units}

Creel surveys measure effort as the amount of fishing activity in a defined
domain over a defined time period. That simple statement still requires three
explicit choices before any arithmetic can begin.

**The unit of effort.** Angler-hours — one angler fishing for one hour — is the
most common unit in North American inland fisheries [@pollock1994]. It is natural
for the instantaneous count estimator because an angler count at a point in time,
multiplied by the length of the fishing period, gives angler-hours directly. Some
designs count boats rather than individual anglers and report boat-hours, then
convert to angler-hours using a mean party size from the interview record. Harlan
records both `boat_anglers` (individual anglers on boats) and `bank_anglers`
(shore anglers), so the natural unit is total angler-hours from all modes combined.

**The domain.** The domain specifies what population the estimate covers — which
water, which species, which access modes, which time of day. The Harlan schedule
covers a single reservoir section (`section = "0"`) surveyed over two daily
shifts (morning period 1 and afternoon period 2) from April through October 2022.
Effort estimated from Harlan counts represents fishing pressure within that
section-shift-season domain; extrapolation beyond it requires assumptions that the
data cannot support.

**The reporting period.** Managers typically want a seasonal or annual total, not
just effort on sampled days. The stratified estimator expands from sampled days to
the full calendar. That expansion rests on an assumption: sampled days within a
stratum are representative of all days in that stratum. The sampling design is
responsible for making this assumption defensible; stratification by day type
(weekday vs. weekend) reduces the risk by keeping groups with different effort
levels in separate strata.

Fishing pressure on a given day-period can also be reported as angler-trips
rather than angler-hours. An angler-trip counts each party as one unit regardless
of how long they fish; it answers "how many trips happened" rather than "how much
fishing time was expended." Angler-hours and angler-trips measure different things
and can diverge substantially when trip lengths vary — a fishery dominated by long
overnight trips has fewer trips but more hours than one dominated by quick
half-day outings. Which unit a manager needs depends on the question being asked.

## The instantaneous count estimator {#sec-effort-formula}

An instantaneous count records the number of anglers active in the survey area at
a single moment in time. Multiplying by the length of the fishing period converts
a count to an angler-hour estimate for that period [@pope_chapter]:

$$\hat{E}_{d,p} = \bar{I}_{d,p} \times T_p$$

where $\hat{E}_{d,p}$ is the estimated angler-hours on day $d$ during period $p$,
$\bar{I}_{d,p}$ is the mean of the $K_{d,p}$ instantaneous counts taken within
that period, and $T_p$ is the total length of the period in hours. The mean
rather than the sum of counts is used because each count estimates the
instantaneous stock of anglers, not a cumulative total; averaging multiple counts
within a period reduces the effect of moment-to-moment fluctuation.

At Harlan, morning shifts (period 1) run 8 hours and afternoon shifts (period 2)
run 6 hours, with two instantaneous counts per shift. For the morning of
September 2, 2022, the two counts recorded 38 and 45 boat anglers (no bank
anglers active):

```{r}
#| label: single-period-example
#| echo: true

sep2_am <- harlan_counts |>
  filter(date == as.Date("2022-09-02"), period == 1) |>
  mutate(total_anglers = boat_anglers + bank_anglers)

sep2_am |>
  select(count_time, bank_anglers, boat_anglers, total_anglers)

mean_count <- mean(sep2_am$total_anglers)
angler_hours_sep2_am <- mean_count * shift_hours["1"]

cat(
  "Mean count:", mean_count, "anglers\n",
  "Shift length:", shift_hours["1"], "hours\n",
  "Estimated effort:", angler_hours_sep2_am, "angler-hours\n"
)
```

The 332 angler-hours estimated for that single morning shift illustrates the
scale of the effort signal. A sampled period with 41.5 mean anglers and an 8-hour
shift contributes more than three person-days of fishing activity to the seasonal
total.

### Why two counts, not one?

With only one count per period, the daily effort estimate has no within-period
variance and the analyst cannot assess how much moment-to-moment fluctuation
exists in angler activity [@mccormick2013]. Two counts allow estimation of
within-day variance, which is a component of total effort variance. The
recommendation from @mccormick2013 is to take at least two counts per period,
spread across the shift, so that the within-day variance can be estimated and
included in the uncertainty budget.

At Harlan, the two counts per period are separated by approximately two hours,
providing coverage of both the early-shift and mid-shift activity levels. The
`add_counts()` call in `tidycreel` records both counts, averages them to a
single period-day mean, and retains the within-day variance for inclusion in
`estimate_effort()` output.

## Stratified expansion to the season total {#sec-effort-strat}

Effort on sampled days must be expanded to the full survey period to produce the
seasonal total managers need. The stratified expansion works by scaling the
stratum sample mean up to the stratum population size [@rasmussen1998]:

$$\hat{E}_h = N_h \cdot \frac{1}{n_h} \sum_{d \in s_h} \hat{E}_{d,p}$$

where $N_h$ is the number of days available in stratum $h$ (from the sampling
calendar), $n_h$ is the number of days actually sampled, and the sum runs over
sampled days $s_h$. The ratio $N_h / n_h$ is the expansion weight: it says "each
sampled day in this stratum represents $N_h / n_h$ days in the population."

The overall season estimate sums across strata:

$$\hat{E} = \sum_h \hat{E}_h$$

For Harlan, the sampling calendar covers 211 days (April 2 to October 29, 2022),
divided into weekday ($N_{\text{weekday}} = 150$) and weekend
($N_{\text{weekend}} = 61$) strata. Of these, 82 weekday period-days and
34 weekend period-days were sampled. Because Harlan runs two periods per day, the
count data has 118 sampled period-day records (83 weekday + 35 weekend, combining
both periods).

```{r}
#| label: stratum-sizes
#| echo: true

harlan_schedule |>
  group_by(day_type) |>
  summarise(
    N_h      = n(),
    n_h      = sum(sampled),
    f_h      = round(n_h / N_h, 3),
    .groups  = "drop"
  )
```

Weekdays are sampled at a slightly higher rate than weekends (55% vs. 56%), but
the stratum sizes are very different: 150 weekday days vs. 61 weekend days. The
weekend stratum is smaller in both total days and sampled days, which matters for
precision — smaller strata with fewer sampled days produce wider confidence
intervals even when their sampling fraction is similar.

### The daily estimator versus the stratum estimator

Two estimators for seasonal effort appear in the creel literature, and they
behave differently in practice. @rasmussen1998 compared them directly using
simulation on the Escanaba Lake dataset.

The **daily estimator** [@malvestuto1978] first computes a mean daily effort for
the full sampling period and then multiplies by the total number of days:

$$\hat{E}_{\text{daily}} = N \cdot \frac{1}{n} \sum_{d=1}^{n} \hat{E}_d$$

This estimator ignores stratum membership and treats all sampled days as
interchangeable. When effort differs systematically between weekdays and weekends
— as it almost always does — the daily estimator will be biased if sampled days
are not proportionally distributed across day types. Random variation in day-type
representation across the sample can produce large positive or negative swings in
the estimate.

The **stratum estimator** [@rasmussen1998] estimates effort separately within
each stratum and then sums. It is unbiased regardless of how day-type composition
varies across the sample, because weekday and weekend days contribute to their
respective stratum means independently. Rasmussen et al. demonstrated that the
stratum estimator produced bias ratios near zero across 1,000 simulations, while
the daily estimator showed substantially larger deviations from census values in
years with low overall effort. All effort estimates in this chapter use the
stratum estimator.

## Variance of the effort estimate {#sec-effort-variance}

The variance of a stratified effort estimate has two distinct sources, and
understanding their magnitudes guides decisions about survey design
[@mccormick2013].

**Between-day variance** is the variance in daily effort across sampled days
within a stratum. It reflects genuine day-to-day variation in how many anglers
fish — driven by weather, season, weekday patterns, and stochastic factors. The
between-day variance estimator for stratum $h$ is the standard stratified
sampling result with a finite population correction:

$$\hat{V}_{\text{between}}(\hat{E}_h) = N_h^2 \cdot \frac{s_{E,h}^2}{n_h}
\left(1 - \frac{n_h}{N_h}\right)$$

where $s_{E,h}^2$ is the sample variance of daily effort estimates across sampled
days in stratum $h$. This term dominates when days differ widely in angler
activity — common for weekend strata where a rainy October weekend and a sunny
June weekend can have an order-of-magnitude difference in effort.

**Within-day variance** is the variance across multiple instantaneous counts
taken within a single period. It reflects moment-to-moment fluctuation in angler
presence during the shift: anglers departing for lunch, boats moving between
sites, brief periods of low activity following a weather event. The within-day
variance contribution is:

$$\hat{V}_{\text{within}} = \sum_h \sum_{d \in s_h}
\left(\frac{T_p^2}{K_d}\right) s_{d}^2$$

where $K_d$ is the number of counts on day $d$ and $s_d^2$ is their sample
variance. With only two counts per period-day, this is just $T_p^2 \cdot
(c_1 - c_2)^2 / 2$.

The total variance combines both:

$$\hat{V}(\hat{E}) = \hat{V}_{\text{between}} + \hat{V}_{\text{within}}$$

`estimate_effort()` reports both components as `se_between` and `se_within`,
making it straightforward to determine which source of uncertainty is more
important. When `se_between` dominates, increasing the number of sampled days
reduces uncertainty more efficiently than adding more count rounds per day. When
`se_within` dominates — rare in practice but possible in highly variable systems
— adding within-day count rounds is the more efficient path [@mccormick2013].

## Preparing count data for effort estimation {#sec-effort-prep}

The `estimate_effort()` function operates on whatever numeric column comes first
in the design's count data that is not a date, strata, or sampling unit
identifier. For Harlan, the raw count table records `bank_anglers`,
`angler_boats`, and `boat_anglers` separately. Passing these columns directly
would cause `estimate_effort()` to estimate effort in units of bank anglers alone
— the first numeric column — rather than in total angler-hours.

The correct workflow is to pre-compute `angler_hours` before building the design:

```{r}
#| label: prep-counts
#| echo: true

# Shift lengths for each period
shift_hours <- c("1" = 8L, "2" = 6L)

harlan_counts_effort <- harlan_counts |>
  mutate(
    period        = as.character(period),        # prevent period from being treated as a count
    total_anglers = boat_anglers + bank_anglers,
    angler_hours  = total_anglers * shift_hours[period]
  ) |>
  select(reservoir, date, day_type, period, section, count_time, angler_hours)

harlan_counts_effort |>
  filter(date == as.Date("2022-09-02")) |>
  select(date, day_type, period, count_time, angler_hours)
```

Two things happen here. First, `period` is converted from numeric to character
so that `estimate_effort()` does not interpret it as a count variable (period
values of 1 and 2 are identifiers, not observations). Second, `angler_hours`
is computed as `total_anglers × shift_hours[period]`, encoding the shift-length
multiplication directly in the data. `add_counts()` then averages the two
within-period records to a single period-day mean — matching the formula
$\bar{I}_{d,p} \times T_p$ at the data preparation stage rather than in the
estimator.

The full design pipeline with effort-preprocessed counts:

```{r}
#| label: design-effort
#| message: false

harlan_design_effort <- creel_design(
  calendar = harlan_schedule,
  date     = "date",
  strata   = "day_type"
) |>
  add_counts(harlan_counts_effort, count_time_col = "count_time") |>
  add_interviews(
    interviews     = harlan_interviews_with_catch,
    catch          = "total_harvest",
    effort         = "hours_fished",
    trip_duration  = "hours_fished",
    trip_status    = "trip_status",
    n_anglers      = "n_anglers",
    angler_type    = "angler_type",
    angler_method  = "angler_method",
    interview_type = "access"
  ) |>
  add_catch(
    data          = harlan_catch,
    catch_uid     = "interview_id",
    interview_uid = "interview_id",
    species       = "species",
    count         = "n_fish",
    catch_type    = "catch_type"
  )
```

The design holds the same survey structure built in Chapter 10 — calendar,
strata, interviews, catch — but the `counts` slot now carries `angler_hours`
as its single numeric count variable. This ensures all downstream effort
estimates are in angler-hours rather than raw angler counts.

## Effort on sampled days {#sec-effort-sampled}

`estimate_effort()` returns effort for the sampled days only when
`target = "sampled_days"` (the default). This is the design-weighted sum of
angler-hours actually observed across all sampled period-day combinations,
without extrapolating to unsampled days.

```{r}
#| label: effort-sampled
#| echo: true

eff_sampled <- estimate_effort(harlan_design_effort, target = "sampled_days")
eff_sampled
```

The 118 sampled period-days produced `r formatC(round(eff_sampled$estimates$estimate), big.mark=",")` observed angler-hours, with a standard error of
`r formatC(round(eff_sampled$estimates$se), big.mark=",")` angler-hours (CV =
`r round(eff_sampled$estimates$se / eff_sampled$estimates$estimate * 100, 1)`%).
This is not a seasonal estimate — it covers only the periods that were actually
surveyed — but it benchmarks the observed effort against which the full-season
expansion can be compared.

The SE decomposes into a between-day component of
`r formatC(round(eff_sampled$estimates$se_between), big.mark=",")` angler-hours
and a within-day component of
`r formatC(round(eff_sampled$estimates$se_within), big.mark=",")` angler-hours.
Between-day variance accounts for the larger share: day-to-day differences in
how many anglers visit Harlan are a more important source of uncertainty than
moment-to-moment fluctuation within a shift. This is the typical pattern for
reservoir fisheries where weather and seasonality drive large swings in daily
attendance.

## Full-season effort estimate {#sec-effort-season}

The primary management deliverable is the full-season angler-hour estimate —
expanded from sampled days to the complete sampling frame using the stratum
weights $N_h / n_h$. The `target = "stratum_total"` argument applies this
expansion:

```{r}
#| label: effort-stratum-total
#| echo: true

eff_season <- estimate_effort(harlan_design_effort, target = "stratum_total")
eff_season
```

Expanding to all 211 calendar days yields an estimated
`r formatC(round(eff_season$estimates$estimate), big.mark=",")` angler-hours
for the Harlan 2022 season (April–October), with a 95% confidence interval of
approximately `r formatC(round(eff_season$estimates$ci_lower), big.mark=",")`
to `r formatC(round(eff_season$estimates$ci_upper), big.mark=",")` angler-hours.
The CV of `r round(eff_season$estimates$se / eff_season$estimates$estimate * 100, 1)`%
satisfies the commonly cited 20% target for management-grade estimates
[@mccormick2013], though it falls in the range that some agencies consider
acceptable rather than comfortable.

The ratio of stratum_total to sampled_days reflects the proportion of the season
that was surveyed. Harlan sampled
`r round(100 * 116 / 211, 0)`% of available days
(`r 116` of 211), so the expansion factor is approximately
`r round(211 / 116, 2)`. The actual expansion differs slightly between strata
because weekday and weekend days were sampled at slightly different rates.

## Effort by stratum {#sec-effort-by-stratum}

The four-cell breakdown by day type and period reveals how effort is distributed
across the sampling frame:

```{r}
#| label: effort-by-stratum
#| echo: true

eff_strata <- estimate_effort(
  harlan_design_effort,
  by     = c(day_type, period),
  target = "stratum_total"
)
eff_strata
```

Three patterns are immediately visible. First, weekday effort substantially
exceeds weekend effort in every period — weekday morning effort alone
(`r formatC(round(eff_strata$estimates$estimate[1]), big.mark=",")` angler-hours)
is larger than total weekend effort across both periods. This is the reverse of
the pattern typical for urban-access reservoirs and reflects Harlan's location
in rural south-central Nebraska, where mid-week fishing by retirees, nearby
residents, and agricultural workers is common.

Second, morning effort (period 1) exceeds afternoon effort (period 2) for both
day types. Morning periods are 8 hours long versus 6 hours for afternoons, but
the count-per-unit-time is also higher in the morning, so the difference reflects
genuine differences in angler behavior rather than just shift-length arithmetic.

Third, the weekend strata carry wider confidence intervals relative to their
estimates (higher CV). The weekend stratum has fewer sampled days and more
variable day-to-day attendance — sunny summer weekends draw large crowds while
cold autumn weekends can be nearly empty — so the between-day variance is larger
relative to the mean.

The marginal day-type totals collapse across periods:

```{r}
#| label: effort-by-daytype
#| echo: true

eff_daytype <- estimate_effort(
  harlan_design_effort,
  by     = day_type,
  target = "stratum_total"
)
eff_daytype
```

Weekdays account for `r round(100 * eff_daytype$estimates$estimate[1] / sum(eff_daytype$estimates$estimate), 0)`%
of total estimated angler-hours
(`r formatC(round(eff_daytype$estimates$estimate[1]), big.mark=",")` AH) against
`r round(100 * eff_daytype$estimates$estimate[2] / sum(eff_daytype$estimates$estimate), 0)`%
for weekends
(`r formatC(round(eff_daytype$estimates$estimate[2]), big.mark=",")` AH). This
split has direct management implications: a regulation targeting weekend fishing
pressure would affect a minority of total season effort, while a midweek closure
would affect the majority.

## Seasonal pattern in effort {#sec-effort-seasonal}

The survey calendar spans April through October. Within that window, effort is
not evenly distributed — spring and early summer typically see higher fishing
pressure than late fall. The mean angler-hours per sampled period-day by month
describes this seasonal arc directly from the observed counts:

```{r}
#| label: fig-seasonal-effort
#| echo: true
#| fig-cap: "Mean angler-hours per sampled period-day by month and day type, Harlan Reservoir 2022. Each bar represents the average of all sampled period-days within a month-day-type combination; error bars show one standard error. The survey covers April through October. Weekday effort (dark) consistently exceeds weekend effort (light) across the season; both peak in May and decline through autumn."
#| fig-alt: "Grouped bar chart with months on the x-axis and mean angler-hours on the y-axis, bars colored by day type (weekday vs. weekend), error bars showing one SE."

monthly_effort <- harlan_design_effort$counts |>
  mutate(month = month(date, label = TRUE, abbr = TRUE)) |>
  group_by(month, day_type) |>
  summarise(
    n       = n(),
    mean_ah = mean(angler_hours, na.rm = TRUE),
    se_ah   = sd(angler_hours, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(monthly_effort, aes(x = month, y = mean_ah, fill = day_type)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(
    aes(ymin = mean_ah - se_ah, ymax = mean_ah + se_ah),
    position = position_dodge(width = 0.8),
    width = 0.25
  ) +
  scale_fill_manual(
    values = creel_palette(2),
    labels = c("weekday" = "Weekday", "weekend" = "Weekend")
  ) +
  labs(
    x    = NULL,
    y    = "Mean angler-hours per sampled period-day",
    fill = "Day type"
  ) +
  theme_creel() +
  theme(legend.position = "top")
```

May records the highest mean effort on sampled period-days for both day types,
with weekday mornings averaging over 300 angler-hours. This likely reflects the
opening of the main spring fishing season at Harlan — walleye and white bass are
highly targeted in May — combined with the long spring days that extend the
productive morning period. Effort declines through summer and into autumn, with
October weekday averages roughly one-fifth of the May peak.

The April values are near zero because most April sampled days fell in early
April before anglers had established regular patterns. This emphasizes a practical
point: sampled days near the margins of the season carry a disproportionate share
of the between-day variance (a zero-effort day and a 300-angler-hour day in the
same stratum drive a large $s_{E,h}^2$). If the calendar boundary is set too
early or too late relative to when fishing actually occurs, the expansion weights
amplify this edge-effect noise into the seasonal total.

## Stratum-level effort estimates with confidence intervals {#sec-effort-ci}

The point estimates and uncertainty intervals for the full-season breakdown
visualize which strata carry the most effort and where estimation uncertainty is
largest:

```{r}
#| label: fig-stratum-effort
#| echo: true
#| fig-cap: "Estimated full-season angler-hours by day type and period, Harlan Reservoir 2022. Bars show the stratum_total estimate (expanded to all calendar days in each stratum). Error bars show 95% confidence intervals from the Taylor linearization variance estimator. Numbers above bars show the point estimate in angler-hours. Weekday mornings (period 1) dominate; weekend effort in either period is a small fraction of weekday effort."
#| fig-alt: "Four grouped bars — weekday period 1, weekday period 2, weekend period 1, weekend period 2 — with error bars showing 95% confidence intervals and numeric labels."

eff_4cell <- eff_strata$estimates |>
  mutate(
    label = paste0(day_type, "\nPeriod ", period),
    label = factor(label, levels = c(
      "weekday\nPeriod 1", "weekday\nPeriod 2",
      "weekend\nPeriod 1", "weekend\nPeriod 2"
    ))
  )

ggplot(eff_4cell, aes(x = label, y = estimate, fill = day_type)) +
  geom_col(width = 0.6) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    width = 0.2
  ) +
  geom_text(
    aes(label = formatC(round(estimate), big.mark = ",")),
    vjust = -0.8, size = 3.2
  ) +
  scale_fill_manual(
    values = creel_palette(2),
    labels = c("weekday" = "Weekday", "weekend" = "Weekend"),
    guide  = "none"
  ) +
  labs(
    x = NULL,
    y = "Estimated angler-hours (full season)"
  ) +
  theme_creel()
```

The confidence intervals reveal why survey design decisions matter. The weekday
period 1 CI spans roughly 14,000 angler-hours — a range large enough to affect
management conclusions if the estimate were used to compare against a target
fishing pressure threshold. The weekend period 2 CI is narrower in absolute terms
but represents a similar proportional uncertainty because weekend effort is
inherently lower. Reducing these intervals requires increasing the number of
sampled days, not adding more count rounds per already-sampled day — because
between-day variance dominates across all strata.

## Precision and reporting {#sec-effort-precision}

The coefficient of variation for the full-season estimate is
`r round(eff_season$estimates$se / eff_season$estimates$estimate * 100, 1)`%.
Whether this is adequate depends on how the estimate will be used. For detecting
year-over-year trends of 20–30%, a CV of 10–15% is generally sufficient because
the trend signal is larger than the noise. For detecting changes of 10% or less —
common when managing regulated fisheries close to exploitation targets — a CV
below 10% is preferable [@mccormick2013].

The standard error decomposes into between-day and within-day components:

```{r}
#| label: se-decomposition
#| echo: true

eff_season$estimates |>
  select(estimate, se, se_between, se_within) |>
  mutate(
    pct_between = round(100 * se_between^2 / se^2, 1),
    pct_within  = round(100 * se_within^2  / se^2, 1)
  )
```

Between-day variance accounts for
`r round(100 * eff_season$estimates$se_between^2 / eff_season$estimates$se^2, 0)`%
of total variance. Adding a third count round to already-sampled days would
reduce the within-day component but leave the dominant between-day component
unchanged. The most efficient way to reduce overall uncertainty is to sample more
days — either by extending the survey season or by increasing the sampling
fraction within existing strata.

Survey programs that want explicit precision targets before the season can use the
variance decomposition from historical data to plan sample sizes. If prior surveys
show $s_{E,h}^2$ for each stratum and the target CV is 15%, the required $n_h$
follows directly from the standard stratified sampling formula [@mccormick2013].

## `tidycreel` support {#sec-effort-tidycreel}

**Implemented for instantaneous count designs:**

- `estimate_effort()` — stratified effort estimator with Taylor variance,
  between/within decomposition, multiple targets (`sampled_days`, `stratum_total`,
  `period_total`), and grouping via `by`
- `estimate_effort_grouped()` — called internally by `estimate_effort()` when `by`
  is specified
- `add_counts()` with `count_time_col` — averages multiple within-period counts
  to a single period-day mean and computes within-day variance contribution

**Constraints and workflow notes:**

- The count variable used by `estimate_effort()` is always the first numeric
  column in the design's count data that is not a date, strata, or PSU column.
  Pre-compute the target unit (here, `angler_hours`) and drop component columns
  before calling `add_counts()` to ensure the correct variable is selected.
- Convert `period` (and any similar integer identifiers) to character before
  calling `add_counts()` to prevent them from being selected as count variables.
- The `stratum_total` target requires that the design have a strata column and
  that the calendar cover the full survey period. Days not in the calendar cannot
  be expanded to.

**Not in scope for this chapter:**

- Bus-route effort estimation — handled by `estimate_effort_br()`, covered in
  Chapter 4 context but not demonstrated here
- Aerial and camera-trap designs — handled by `estimate_effort_aerial()` and
  `estimate_effort_camera()`, introduced in Chapter 6
- Sample size planning — requires external calculation using variance estimates
  from this chapter as inputs; `tidycreel` does not currently provide an
  `optimal_n()` function (see `TIDYCREEL-WISHLIST.md`)

## Takeaway {#sec-effort-takeaway}

Effort estimation is the point of the count schedule and the foundation for
every harvest and catch total the survey produces. Getting it right means three
things.

First, the instantaneous count estimator assumes that each count represents the
instantaneous angler stock during the period. That assumption holds when counts
are taken at random or representative times within the period. Counts clustered
at the same time of day — always at 9:00 a.m., always at midday — can produce
biased effort estimates if angler activity peaks and valleys are systematic.

Second, stratification by day type is not optional decoration. Without it, the
random day-to-day composition of the sample can produce large swings in the
estimated total. With it, weekday and weekend effort are estimated from samples of
their respective populations, and the expansion to the full season is unbiased
regardless of how any particular week was sampled.

Third, between-day variance almost always dominates. The lesson for survey design
is that sampling more days per season is more effective than adding count rounds
to days already sampled. A second or third count round per shift reduces within-day
noise, but if days vary substantially in effort — as they do at any reservoir
influenced by weather and season — only more days can reduce the dominant
between-day component.

The Harlan 2022 season estimate of approximately
`r formatC(round(eff_season$estimates$estimate), big.mark=",")` angler-hours
(95% CI `r formatC(round(eff_season$estimates$ci_lower), big.mark=",")`–`r formatC(round(eff_season$estimates$ci_upper), big.mark=",")`) is the
number that Chapter 13 will use to scale harvest and catch totals. The
downstream confidence intervals are wider because effort uncertainty
multiplies through $\hat{E} \times \hat{R}$.
