---
title: "Catch, Harvest, and Release Estimation"
---

```{r}
#| label: setup-ch13
#| include: false
library(tidycreel)
library(tidyverse)
library(lubridate)
library(here)

harlan_schedule   <- readRDS(here("data/harlan_schedule.rds"))
harlan_counts     <- readRDS(here("data/harlan_counts.rds"))
harlan_interviews <- readRDS(here("data/harlan_interviews.rds"))
harlan_catch      <- readRDS(here("data/harlan_catch.rds"))

# Shift lengths used in Chapter 12
shift_hours <- c("1" = 8L, "2" = 6L)

# Pre-compute angler_hours and drop component columns
harlan_counts_ch13 <- harlan_counts |>
  mutate(
    period        = as.character(period),
    total_anglers = boat_anglers + bank_anglers,
    angler_hours  = total_anglers * shift_hours[period]
  ) |>
  select(reservoir, date, day_type, period, section, count_time, angler_hours)

# Pivot catch data to one row per interview with harvested / released columns
catch_by_interview <- harlan_catch |>
  group_by(interview_id, catch_type) |>
  summarise(n_fish = sum(n_fish), .groups = "drop") |>
  pivot_wider(
    names_from  = catch_type,
    values_from = n_fish,
    values_fill = 0L
  ) |>
  mutate(total_catch = harvested + released)

# Join catch summary to interviews and fill zeros for no-catch interviews
harlan_interviews_ch13 <- harlan_interviews |>
  left_join(catch_by_interview, by = "interview_id") |>
  mutate(
    harvested   = if_else(is.na(harvested),   0L, as.integer(harvested)),
    released    = if_else(is.na(released),    0L, as.integer(released)),
    total_catch = if_else(is.na(total_catch), 0L, as.integer(total_catch))
  )

# Build the design object
# catch = total fish encountered (harvested + released)
# harvest = fish retained
suppressMessages({
  harlan_design_ch13 <- creel_design(
    calendar = harlan_schedule,
    date     = "date",
    strata   = "day_type"
  ) |>
    add_counts(harlan_counts_ch13, count_time_col = "count_time") |>
    add_interviews(
      interviews     = harlan_interviews_ch13,
      catch          = "total_catch",
      harvest        = "harvested",
      effort         = "hours_fished",
      trip_duration  = "hours_fished",
      trip_status    = "trip_status",
      n_anglers      = "n_anglers",
      angler_type    = "angler_type",
      angler_method  = "angler_method",
      interview_type = "access"
    ) |>
    add_catch(
      data          = harlan_catch,
      catch_uid     = "interview_id",
      interview_uid = "interview_id",
      species       = "species",
      count         = "n_fish",
      catch_type    = "catch_type"
    )
})
```

## Three quantities, one fishery {#sec-catch-distinctions}

Effort is the foundation of creel estimation, but managers usually ask about
catch, harvest, and species. Those questions sound similar but they require
different estimators and different data columns.

**Catch** is the total number of fish that left the water during a fishing
trip, regardless of what happened to them afterward. A walleye that was
measured and returned is a caught fish. So is one that went in the livewell.
Catch includes both harvest and release.

**Harvest** is the subset of catch that anglers retained — fish taken home,
eaten, or otherwise removed from the fishery permanently. Harvest is the
quantity most directly tied to fishing mortality, and it is what regulations
target when they set bag limits or size restrictions.

**Release** is the complement: fish caught and returned. Release is more
subject to measurement error than harvest because creel clerks can inspect
retained fish directly but must rely on angler recall for fish already back in
the water. Access-point surveys reduce this error by interviewing anglers
immediately after they leave the water, when memory of released fish is still
fresh [@pollock1994].

The distinction matters. An angler who catches fifteen white bass but keeps
only three contributes a harvest total of three and a release total of twelve,
even though the trip produced one catch total of fifteen. The estimator and
its denominator have to match the question being answered.

This chapter shows how the Harlan Reservoir 2022 survey data produce estimates
of all three quantities using `tidycreel`, and how the relationship between
them reveals the character of the fishery.

## The harvest estimator {#sec-harvest-estimator}

Total harvest is estimated as the product of total angling effort and a
per-unit-effort rate. The rate is the **harvest per unit effort** (HPUE):

$$
\widehat{H}_j = \widehat{E}_j \times \widehat{\text{HPUE}}_j
$$

where $\widehat{E}_j$ is the estimated total angling effort in stratum $j$
(angler-hours) and $\widehat{\text{HPUE}}_j$ is the ratio-of-means HPUE for
that stratum. The season total sums across strata:

$$
\widehat{H} = \sum_j \widehat{H}_j
$$

with variance propagated through the delta method, combining uncertainty from
both the effort and HPUE estimates.

The ratio-of-means HPUE is defined as total observed harvest divided by total
observed effort across all interviews in a stratum [@pollock1994; @hoenig1997]:

$$
\widehat{\text{HPUE}}_j = \frac{\sum_{i} h_i}{\sum_{i} l_i}
$$

where $h_i$ is fish harvested on trip $i$ and $l_i$ is hours fished.
`estimate_harvest_rate()` defaults to completed trips, matching the CPUE
estimator. Harvest from an incomplete trip is theoretically observable — what
is in the livewell at interview time is harvested regardless of whether the
angler keeps fishing — but using the same interview pool as CPUE makes the
HPUE + RPUE ≈ CPUE consistency check exact rather than approximate.

The same product structure applies to release:
$\widehat{R}_j = \widehat{E}_j \times \widehat{\text{RPUE}}_j$, and to total
catch: $\widehat{C}_j = \widehat{E}_j \times \widehat{\text{CPUE}}_j$. Because
all three rate estimators default to the same completed-trip interview pool, the
season totals satisfy $\widehat{C} \approx \widehat{H} + \widehat{R}$ up to
rounding. Roving designs use a mean-of-ratios estimator; see Chapter 17 for
that case.

## Preparing catch data for estimation {#sec-catch-prep}

The design construction in this chapter differs from Chapter 12 in one
important respect. Chapter 12 passed a single harvest column to
`add_interviews()` as the `catch` argument. Here, we need separate quantities
for total catch and harvest. Released fish follow automatically from their
difference.

The `harlan_catch` table stores species-level fish counts with a `catch_type`
column that takes values `"harvested"` or `"released"`. Pivot that table to
one row per interview with separate columns for each type.

```{r}
#| label: catch-prep
#| echo: true
catch_by_interview <- harlan_catch |>
  group_by(interview_id, catch_type) |>
  summarise(n_fish = sum(n_fish), .groups = "drop") |>
  pivot_wider(
    names_from  = catch_type,
    values_from = n_fish,
    values_fill = 0L
  ) |>
  mutate(total_catch = harvested + released)

catch_by_interview
```

The result has one row per interview that recorded at least one fish. When we
join it to the full interview table, any missing interview receives zeros for
all three columns.

```{r}
#| label: catch-join
#| echo: true
harlan_interviews_ch13 <- harlan_interviews |>
  left_join(catch_by_interview, by = "interview_id") |>
  mutate(
    harvested   = if_else(is.na(harvested),   0L, as.integer(harvested)),
    released    = if_else(is.na(released),    0L, as.integer(released)),
    total_catch = if_else(is.na(total_catch), 0L, as.integer(total_catch))
  )

harlan_interviews_ch13 |>
  select(interview_id, date, day_type, hours_fished, trip_status,
         total_catch, harvested, released)
```

The interview table now carries three catch columns at the trip level. Each
feeds a different estimation function downstream.

## Building the design object {#sec-ch13-design}

With three catch columns available, the design pipeline is identical to
Chapter 12 except that we now pass both `catch` and `harvest` to
`add_interviews()`. The `harvest` argument tells `tidycreel` which column
represents retained fish.

```{r}
#| label: design-build
#| echo: true
suppressMessages({
  harlan_design_ch13 <- creel_design(
    calendar = harlan_schedule,
    date     = "date",
    strata   = "day_type"
  ) |>
    add_counts(harlan_counts_ch13, count_time_col = "count_time") |>
    add_interviews(
      interviews     = harlan_interviews_ch13,
      catch          = "total_catch",
      harvest        = "harvested",
      effort         = "hours_fished",
      trip_duration  = "hours_fished",
      trip_status    = "trip_status",
      n_anglers      = "n_anglers",
      angler_type    = "angler_type",
      angler_method  = "angler_method",
      interview_type = "access"
    ) |>
    add_catch(
      data          = harlan_catch,
      catch_uid     = "interview_id",
      interview_uid = "interview_id",
      species       = "species",
      count         = "n_fish",
      catch_type    = "catch_type"
    )
})

harlan_design_ch13
```

Count data preparation mirrors Chapter 12: shift lengths convert instantaneous
angler counts to angler-hours, so effort and rates share the same time units.

## Harvest rates by day type {#sec-harvest-rate}

The harvest rate — fish harvested per angler-hour — is the ratio-of-means
HPUE from 390 completed-trip interviews. `estimate_harvest_rate()` defaults to
completed trips, matching the CPUE estimator (see @sec-harvest-estimator).

```{r}
#| label: harvest-rate-overall
#| echo: true
estimate_harvest_rate(harlan_design_ch13)
```

The overall Harlan 2022 harvest rate is 0.585 fish per angler-hour
(95% CI: 0.495–0.676). Disaggregating by day type shows that weekend anglers
harvest at a substantially higher rate than weekday anglers:

```{r}
#| label: harvest-rate-daytype
#| echo: true
estimate_harvest_rate(harlan_design_ch13, by = day_type)
```

```{r}
#| label: release-rate-daytype
#| echo: true
estimate_release_rate(harlan_design_ch13, by = day_type)
```

```{r}
#| label: fig-rates-daytype
#| echo: true
#| fig-cap: "Estimated harvest rate (HPUE) and release rate (RPUE) by day type at Harlan Reservoir, 2022. Bars show the ratio-of-means estimate from completed-trip interviews; error bars show 95% confidence intervals. Weekend anglers harvested at a noticeably higher rate than weekday anglers, while release rates were similar across day types."
#| fig-alt: "Grouped bar chart with HPUE and RPUE on the y axis and weekday/weekend on the x axis. Weekend HPUE (0.745) is higher than weekday HPUE (0.558). Release rates are similar between day types at approximately 0.924 (weekday) and 0.897 (weekend)."

hpue_by_dt <- estimate_harvest_rate(harlan_design_ch13, by = day_type)
rpue_by_dt <- estimate_release_rate(harlan_design_ch13, by = day_type)

rates_dt <- bind_rows(
  hpue_by_dt$estimates |> mutate(rate_type = "Harvest rate (HPUE)"),
  rpue_by_dt$estimates |> mutate(rate_type = "Release rate (RPUE)")
)

ggplot(rates_dt, aes(x = day_type, y = estimate, fill = rate_type)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.7),
    width    = 0.25,
    colour   = "grey30"
  ) +
  scale_fill_manual(
    values = c("Harvest rate (HPUE)" = "#2171b5",
               "Release rate (RPUE)" = "#74c476"),
    name   = NULL
  ) +
  labs(x = "Day type", y = "Fish per angler-hour") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

The weekend harvest rate (0.745 fish/hour, 95% CI: 0.499–0.991) is about
33% higher than weekdays (0.558, 95% CI: 0.459–0.656). Release rates are
nearly identical — 0.924 on weekdays and 0.897 on weekends.
Weekend anglers both catch more (CPUE: 1.64 vs 1.48 fish/hr weekday) and
retain a larger fraction of what they catch. The weekend fishery at
Harlan draws more anglers actively targeting harvestable species, while
weekday fishing is often more incidental in character.

The stratum-specific rates matter because they multiply against different effort
volumes to produce the season harvest totals. A higher per-hour rate on weekends
does not mean weekends dominate the harvest — that depends on how many
angler-hours each stratum contributes.

## Estimating total harvest {#sec-total-harvest}

`estimate_total_harvest()` combines effort and HPUE via the delta method. The
`target` argument controls which effort domain is multiplied by the rate:
`"sampled_days"` (days when at least one count occurred) or `"stratum_total"`
(all days in the stratum, sampled or not). Use `"stratum_total"` when reporting
seasonal harvest for the full fishery.

```{r}
#| label: total-harvest-sampled
#| echo: true
estimate_total_harvest(harlan_design_ch13)
```

```{r}
#| label: total-harvest-stratum
#| echo: true
estimate_total_harvest(harlan_design_ch13, target = "stratum_total")
```

On sampled days only, estimated harvest was 11,052 fish (95% CI: 8,092–14,012).
Expanding to all days in the season yields 19,825 fish (95% CI: 14,643–25,006).
The difference reflects the unsampled days: the creel did not cover every day
on the calendar, but the expansion assumes those days had fishing activity at
the same rate structure as sampled days. The sampled-days total describes what
happened on surveyed days; the stratum-total estimates season-level removal from
the stock.

Grouping by day type connects the rates from @sec-harvest-rate to the season
totals:

```{r}
#| label: total-harvest-by-daytype
#| echo: true
estimate_total_harvest(
  harlan_design_ch13,
  by     = day_type,
  target = "stratum_total"
)
```

```{r}
#| label: fig-harvest-totals
#| echo: true
#| fig-cap: "Estimated season harvest by day type at Harlan Reservoir, 2022, using the stratum-total effort expansion. Error bars show 95% confidence intervals. Despite weekends having a higher harvest rate per angler-hour, weekdays contribute roughly four times more harvest in absolute terms because weekday fishing pressure is substantially greater."
#| fig-alt: "Horizontal bar chart with weekday and weekend on the y axis and estimated harvest on the x axis. Weekday bar (15,793 fish) is roughly four times longer than weekend bar (4,032 fish). Error bars on both bars do not overlap zero."

harvest_by_dt <- estimate_total_harvest(
  harlan_design_ch13,
  by     = day_type,
  target = "stratum_total"
)

ggplot(harvest_by_dt$estimates,
       aes(x = estimate, y = fct_rev(day_type))) +
  geom_col(fill = "#2171b5", width = 0.5) +
  geom_errorbar(
    aes(xmin = ci_lower, xmax = ci_upper),
    width  = 0.2,
    colour = "grey30"
  ) +
  labs(
    x = "Estimated harvest (fish)",
    y = NULL
  ) +
  theme_bw(base_size = 12)
```

Weekdays account for roughly four times more harvest than weekends in absolute
terms (15,793 vs. 4,032 fish). Even though the per-hour harvest rate is higher
on weekends, weekday effort at Harlan is substantially larger — enough to
reverse the relationship between rate and total. This is the harvest analog of
the effort result from Chapter 12: a higher rate does not guarantee a higher
total when the two strata differ in total effort by a wide margin.

## Estimating total catch and release {#sec-catch-release}

Harvest and release are the two components of total catch. Estimating each
separately allows the analyst to report how much fishing activity ended with
fish returned to the water — an increasingly important management metric in
fisheries where exploitation targets are low or catch-and-release is promoted.

```{r}
#| label: total-catch
#| echo: true
estimate_total_catch(harlan_design_ch13, target = "stratum_total")
```

```{r}
#| label: total-release
#| echo: true
estimate_total_release(harlan_design_ch13, target = "stratum_total")
```

The estimated total catch for the 2022 season was 49,756 fish
(95% CI: 37,272–62,239). Of that, approximately 19,825 were harvested and
30,310 released — roughly 60% of all fish caught were returned. That is a
large catch-and-release fraction, and it raises a question the estimate alone
cannot answer: do those released fish survive? The creel measures the number
returned; post-release mortality is a separate quantity requiring tagging
studies or laboratory experiments and is applied outside the creel workflow.

The catch rate (CPUE) and release rate (RPUE) confirm the partition:

```{r}
#| label: release-rate-overall
#| echo: true
estimate_release_rate(harlan_design_ch13)
```

```{r}
#| label: catch-rate-overall
#| echo: true
estimate_catch_rate(harlan_design_ch13)
```

All three rate estimators (`estimate_catch_rate()`, `estimate_harvest_rate()`,
`estimate_release_rate()`) default to the 390 completed-trip interviews (65%
of 600 parties). Using the same interview pool makes the consistency check
exact: HPUE (0.585) + RPUE (0.920) = 1.505, which rounds to CPUE (1.51),
confirming the catch partition is internally coherent.

## Reporting catch, harvest, and release together {#sec-combined-reporting}

Management reports often need all three quantities in a single table with
consistent confidence intervals. Collect the stratum-total estimates for each
and bind them by row.

```{r}
#| label: combined-table
#| echo: true
harvest_est <- estimate_total_harvest(harlan_design_ch13, target = "stratum_total")
catch_est   <- estimate_total_catch(harlan_design_ch13,   target = "stratum_total")
release_est <- estimate_total_release(harlan_design_ch13, target = "stratum_total")

bind_rows(
  harvest_est$estimates |> mutate(quantity = "Total harvest"),
  release_est$estimates |> mutate(quantity = "Total release"),
  catch_est$estimates   |> mutate(quantity = "Total catch")
) |>
  select(quantity, estimate, se, ci_lower, ci_upper, n) |>
  mutate(across(estimate:ci_upper, \(x) round(x)))
```

The combined table expresses the accounting identity of the fishery: catch
equals harvest plus release. The estimates do not sum exactly because the
component estimators differ slightly in their interview pools and variance
calculations, but they share the same effort base and the same interview
records.

When presenting this table in a management report, the relevant context is
what the estimates imply for fishing mortality. An estimated 19,825 harvested
fish represents direct removal from the Harlan population in 2022 — the number
an exploitation rate or yield model requires as input. The 30,310 released fish
contribute to fishing mortality only through post-release mortality, a
species-specific quantity not estimated by the creel.

## Species composition of the harvest {#sec-species-composition}

The estimates above report totals across all species combined. Species-specific
harvest estimates require species-level catch rates from
`estimate_catch_rate(by = species)`, which Chapter 14 covers in detail.
For composition analysis — understanding which species anglers encounter and
what fraction they retain — the raw catch table is the right tool. It records
the actual species and catch type for every fish without expansion to the
population level.

```{r}
#| label: species-comp-table
#| echo: true
harlan_catch |>
  group_by(species, catch_type) |>
  summarise(n_fish = sum(n_fish), .groups = "drop") |>
  pivot_wider(
    names_from  = catch_type,
    values_from = n_fish,
    values_fill = 0L
  ) |>
  mutate(
    total        = harvested + released,
    pct_released = round(released / total * 100, 1)
  ) |>
  arrange(desc(total))
```

White bass dominate by both total catch (4,428 fish) and harvest (2,509),
with 43% released. Walleye show a starkly different pattern: 1,899 fish caught,
200 retained, 89.5% released. Freshwater drum are similarly high at 91%
released. These patterns reflect angler preference and regulation structure:
walleye have a size limit and are a prized food fish, so sub-legal fish are
released; freshwater drum are often returned regardless of size. Gizzard shad
show 0% released — every shad caught was harvested, consistent with their use
as cut bait.

```{r}
#| label: fig-species-comp
#| echo: true
#| fig-cap: "Species catch composition at Harlan Reservoir, 2022. Bars show recorded fish counts from interview catch records, split by harvest (blue) and release (green). Species are ordered by total catch. White bass dominate both harvest and total catch; walleye and freshwater drum show high release fractions consistent with size-limit regulations and angler preferences."
#| fig-alt: "Horizontal stacked bar chart with species on the y axis ordered from most-caught (White Bass) to least-caught (Crappie). Each bar is split into harvested (blue) and released (green) segments. White Bass has the longest bar at over 4,000 fish; Walleye has a very large green (released) segment."

species_comp <- harlan_catch |>
  group_by(species, catch_type) |>
  summarise(n_fish = sum(n_fish), .groups = "drop") |>
  group_by(species) |>
  mutate(total = sum(n_fish)) |>
  ungroup() |>
  mutate(
    species    = fct_reorder(species, total),
    catch_type = factor(catch_type, levels = c("released", "harvested"))
  )

ggplot(species_comp, aes(x = n_fish, y = species, fill = catch_type)) +
  geom_col() +
  scale_fill_manual(
    values = c("harvested" = "#2171b5", "released" = "#74c476"),
    labels = c("Harvested", "Released"),
    name   = NULL
  ) +
  labs(
    x = "Number of fish (raw interview records)",
    y = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

The species composition in raw catch records describes which species anglers
encounter and what they do with them. It does not, by itself, produce an
expanded season total per species — that requires design-based species-level
CPUE estimates applied to total effort, which Chapter 14 covers using
`estimate_catch_rate(by = species)`.

## What the creel can and cannot measure {#sec-ch13-limits}

A well-executed creel survey provides design-unbiased estimates of harvest,
catch, and release at the stratum level. Three limitations are worth stating
clearly.

**Post-release mortality is not measured.** The creel records that a fish was
returned to the water; it does not follow that fish. For species with high
post-release survival under typical conditions (most warm-water species in
summer), the released total approximates the number that survive. For species
with elevated post-release mortality — fish caught on bait, held in warm
water, or deeply hooked — some fraction of the release total contributes to
fishing mortality even though it was not recorded as harvest. Adjustments for
post-release mortality require independent data from tagging studies or
laboratory experiments and are applied outside the creel estimation workflow.

**Release counts depend on angler recall.** In an access-point survey, creel
clerks inspect retained fish directly, so harvest counts and species
identification are accurate for kept fish. Released fish are reported from
memory. At Harlan, where anglers are interviewed immediately at the boat ramp,
that recall is fresh — but an angler who released fifty white bass over a
four-hour trip may not have counted precisely. For common rough fish that
anglers return automatically, the released count is more likely to be
approximated than the retained count [@pollock1994].

**Interviews with no catch record mean zero catch, not missing data.** An
interview with no rows in `harlan_catch` means the angler reported catching
nothing — not that the question was skipped. The `left_join` to
`catch_by_interview` and the subsequent zero-fill treat those interviews
correctly: they contribute a zero to both the numerator and the denominator of
the ratio-of-means estimators.

## Takeaway {#sec-ch13-takeaway}

Catch, harvest, and release are distinct quantities produced by the same set
of interviews. `estimate_total_harvest()`, `estimate_total_catch()`, and
`estimate_total_release()` each return a design-based season estimate with
propagated variance. `add_interviews()` needs explicit `catch` and `harvest`
arguments.

By default, `estimate_harvest_rate()` and `estimate_release_rate()` use
completed-trip interviews, matching `estimate_catch_rate()`. Harvest and
release are theoretically observable for incomplete trips — the livewell
contents and any fish already returned are directly countable — but the
default promotes exact HPUE + RPUE = CPUE consistency. For access-point
designs, CPUE cannot include incomplete trips because the final catch total is
unknown. Roving designs use a mean-of-ratios estimator that treats each
mid-trip contact as an instantaneous rate observation.

At Harlan in 2022, roughly 60% of all fish caught were released. Weekend
harvest rates were higher than weekday rates, but weekday fishing pressure was
large enough that weekdays still account for about 80% of season harvest. Rate
and total tell different stories about the fishery, and both are necessary for
a complete picture.
