---
title: "CPUE Estimation"
---

```{r}
#| label: setup-ch14
#| include: false
library(tidycreel)
library(tidyverse)
library(lubridate)
library(here)

harlan_schedule   <- readRDS(here("data/harlan_schedule.rds"))
harlan_counts     <- readRDS(here("data/harlan_counts.rds"))
harlan_interviews <- readRDS(here("data/harlan_interviews.rds"))
harlan_catch      <- readRDS(here("data/harlan_catch.rds"))

shift_hours <- c("1" = 8L, "2" = 6L)

harlan_counts_ch14 <- harlan_counts |>
  mutate(
    period        = as.character(period),
    total_anglers = boat_anglers + bank_anglers,
    angler_hours  = total_anglers * shift_hours[period]
  ) |>
  select(reservoir, date, day_type, period, section, count_time, angler_hours)

catch_by_interview <- harlan_catch |>
  group_by(interview_id, catch_type) |>
  summarise(n_fish = sum(n_fish), .groups = "drop") |>
  pivot_wider(
    names_from  = catch_type,
    values_from = n_fish,
    values_fill = 0L
  ) |>
  mutate(total_catch = harvested + released)

harlan_interviews_ch14 <- harlan_interviews |>
  left_join(catch_by_interview, by = "interview_id") |>
  mutate(
    harvested   = if_else(is.na(harvested),   0L, as.integer(harvested)),
    released    = if_else(is.na(released),    0L, as.integer(released)),
    total_catch = if_else(is.na(total_catch), 0L, as.integer(total_catch))
  )

suppressMessages({
  harlan_design_ch14 <- creel_design(
    calendar = harlan_schedule,
    date     = "date",
    strata   = "day_type"
  ) |>
    add_counts(harlan_counts_ch14, count_time_col = "count_time") |>
    add_interviews(
      interviews     = harlan_interviews_ch14,
      catch          = "total_catch",
      harvest        = "harvested",
      effort         = "hours_fished",
      trip_duration  = "hours_fished",
      trip_status    = "trip_status",
      n_anglers      = "n_anglers",
      angler_type    = "angler_type",
      angler_method  = "angler_method",
      interview_type = "access"
    ) |>
    add_catch(
      data          = harlan_catch,
      catch_uid     = "interview_id",
      interview_uid = "interview_id",
      species       = "species",
      count         = "n_fish",
      catch_type    = "catch_type"
    )
})
```

## What CPUE is {#sec-cpue-what}

Catch-per-unit-effort (CPUE) compresses the fishing experience into a single ratio:
catch divided by effort. That compression is both its appeal and its main source of
misinterpretation.

In a roving or access-point creel survey, effort is measured in angler-hours —
the product of the number of anglers and the hours they fished. Catch is the
number of fish contacted over that same duration. The resulting ratio, in units of
fish per angler-hour, is a population-average catch rate: how many fish a
representative angler would encounter per hour under the survey conditions.

CPUE is not an index of abundance in the strict sense. A higher CPUE could reflect
better fishing conditions, a change in effort distribution, or an actual increase in
fish density. It should be labeled as the rate observed under the sampling design,
not as a direct measure of population size.

Within a single survey period, however, CPUE is exactly what management reporting
requires: an estimate of how productive the fishery appeared to anglers who
participated. That quantity, with its uncertainty, is what `tidycreel` computes
from interview records.

## Two ratio estimators {#sec-cpue-estimators}

There are two ways to form a ratio from interview data, and they answer subtly
different questions [@hoenig1997].

The **ratio-of-means** (ROM) estimator sums catch and effort separately across
all interviews, then divides:

$$\widehat{\text{CPUE}}_{\text{ROM}} = \frac{\sum_{i=1}^{n} c_i}{\sum_{i=1}^{n} e_i}$$

where $c_i$ is the total catch for interview $i$ and $e_i$ is the hours fished.
This estimator is weighted by trip length: anglers who fished longer contribute
more to both the numerator and the denominator. It is the natural estimator for
completed trips because the full trip record — final catch and full elapsed hours —
is available at the time of interview.

The **mean-of-ratios** (MOR) estimator computes a per-trip rate for each interview
first, then averages those rates:

$$\widehat{\text{CPUE}}_{\text{MOR}} = \frac{1}{n} \sum_{i=1}^{n} \frac{c_i}{e_i}$$

This gives each trip equal weight regardless of duration. It is the natural
estimator for incomplete trips, where the interview catches an angler mid-trip and
$c_i$ and $e_i$ are snapshots of a fishing session still in progress. If the catch
process is stationary — if the rate of catching fish does not change systematically
within a trip — then the MOR applied to incomplete trips should estimate the same
parameter as the ROM applied to completed trips.

The stationarity assumption is the critical qualifier. If anglers learn during a
trip, early-trip snapshots will underestimate the completed-trip rate. @hoenig1997
showed that truncating very short trips reduces bias. The `estimator = "mortr"`
option in `estimate_catch_rate()` applies this truncation automatically.

@pollock1994 recommended using completed trips wherever possible because the ROM
estimator requires no stationarity assumption. This book follows that convention:
`estimate_catch_rate()` uses the ROM estimator on completed trips by default.

A practical diagnostic compares both estimators on the same dataset. If the
completed and incomplete estimates are close, the stationarity assumption is
supported.

## The Harlan CPUE baseline {#sec-cpue-harlan-baseline}

The 2022 Harlan Reservoir creel survey collected 600 party interviews. Of those,
390 were completed trips (65%) and 210 were incomplete (35%). The ratio-of-means
estimator uses the 390 completed-trip records.

```{r}
#| label: cpue-overall
#| echo: true
cpue_overall <- estimate_catch_rate(harlan_design_ch14)
cpue_overall
```

The output tibble has five columns: `estimate` is the ratio-of-means CPUE,
`se` is the Taylor-linearized standard error, `ci_lower` and `ci_upper` are the
95% confidence interval bounds, and `n` is the number of completed-trip interviews.

The overall CPUE estimate is 1.51 fish per angler-hour (95% CI: [1.32, 1.69]).
The standard error of 0.093 yields a CV of about 6.2%, which falls comfortably
below the 15% precision threshold used in Chapter 15. The 390 completed-trip
sample is adequate for a precise ratio estimate at the survey-wide level.

This estimate is the average catch rate for all anglers in all strata combined.
It answers the question: how many fish did a typical angler at Harlan encounter
per hour in 2022? It does not describe abundance or year-to-year change.

## CPUE by day-type stratum {#sec-cpue-by-stratum}

The survey design stratifies the calendar into weekdays and weekends. Weekend days
attract different angler populations and can have different crowding, access
patterns, and fishing conditions than weekdays. Stratum-specific CPUE separates
those effects.

```{r}
#| label: cpue-by-daytype
#| echo: true
cpue_dt <- estimate_catch_rate(harlan_design_ch14, by = day_type)
cpue_dt$estimates
```

```{r}
#| label: fig-ch14-stratum-ci
#| echo: true
#| fig-cap: "CPUE by day-type stratum at Harlan Reservoir, 2022. Points show ratio-of-means estimates from completed-trip interviews; horizontal bars show 95% confidence intervals. Labels report the estimated rate and sample size. Weekend CPUE appears slightly higher but the wider confidence interval reflects the smaller completed-trip sample."
#| fig-alt: "Forest plot with two rows (Weekday and Weekend) on the y axis. Each row shows a point estimate and horizontal error bar for 95% confidence interval. Weekday CI is narrow (roughly 1.3–1.7); Weekend CI is wide (roughly 1.2–2.1). A vertical dashed reference line marks the overall estimate of 1.51."
cpue_dt$estimates |>
  mutate(
    day_type = str_to_title(day_type),
    label    = sprintf("%.2f (n = %d)", estimate, n)
  ) |>
  ggplot(aes(x = estimate, y = day_type)) +
  geom_vline(xintercept = 1.51, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                 height = 0.15, linewidth = 0.8) +
  geom_point(size = 3) +
  geom_text(aes(label = label), vjust = -0.8, size = 3.5) +
  labs(
    x = "CPUE (fish per angler-hour)",
    y = NULL,
    title = "CPUE by day-type stratum",
    caption = "Dashed line: overall estimate (1.51 fish/hr). Error bars: 95% CI."
  ) +
  xlim(0.9, 2.4) +
  theme_minimal()
```

Weekend CPUE (1.64 fish/hr) is slightly higher than weekday CPUE (1.48 fish/hr),
but the weekend estimate carries a much wider confidence interval because only 69
weekend interviews were completed trips, compared to 321 weekday interviews.

## CPUE by angler type {#sec-cpue-by-angler-type}

The interviews distinguish bank anglers from boat anglers. These two groups fish
differently: boat anglers can move to productive structure, cover more water, and
access mid-lake habitat; bank anglers are constrained to the shoreline. If the
fish community or the productive habitat is distributed unevenly, the two groups
should show different catch rates.

```{r}
#| label: cpue-by-anglertype
#| echo: true
cpue_at <- estimate_catch_rate(harlan_design_ch14, by = angler_type)
cpue_at$estimates
```

```{r}
#| label: fig-ch14-angler-type
#| echo: true
#| fig-cap: "CPUE by angler type at Harlan Reservoir, 2022. Bank anglers show a substantially higher catch rate than boat anglers, with overlapping confidence intervals only at the outer bounds. The difference reflects both access patterns and target species, not equivalent fishing conditions."
#| fig-alt: "Forest plot with two rows (Bank and Boat) on the y axis. Bank angler point is near 3.0 with a wide CI extending from about 1.8 to 4.1; Boat angler point is near 1.4 with a narrow CI from about 1.2 to 1.6. A vertical dashed reference line marks the overall estimate of 1.51."
cpue_at$estimates |>
  mutate(
    angler_type = str_to_title(angler_type),
    label       = sprintf("%.2f (n = %d)", estimate, n)
  ) |>
  ggplot(aes(x = estimate, y = angler_type)) +
  geom_vline(xintercept = 1.51, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                 height = 0.15, linewidth = 0.8) +
  geom_point(size = 3) +
  geom_text(aes(label = label), vjust = -0.8, size = 3.5) +
  labs(
    x = "CPUE (fish per angler-hour)",
    y = NULL,
    title = "CPUE by angler type",
    caption = "Dashed line: overall estimate (1.51 fish/hr). Error bars: 95% CI."
  ) +
  xlim(0, 5) +
  theme_minimal()
```

Bank anglers report 2.95 fish per angler-hour (95% CI: [1.83, 4.07]), more than
twice the boat angler rate of 1.41 fish/hr (95% CI: [1.23, 1.59]). The comparison
still conflates method, location, and target species, so it should not be read as
a pure productivity effect.

## Validating the estimator: the complete vs incomplete diagnostic {#sec-cpue-diagnostic}

The choice between ROM and MOR matters most when incomplete trips are numerous
and when the analyst suspects non-stationary catch processes. `tidycreel` provides
a diagnostic mode that applies both estimators and compares the results.

```{r}
#| label: cpue-diagnostic
#| echo: true
cpue_diag <- estimate_catch_rate(harlan_design_ch14, use_trips = "diagnostic")
cpue_diag$comparison
```

The completed-trip estimate is 1.51 fish/hr; the incomplete-trip truncated MOR
estimate is 1.48 fish/hr. The difference is 0.027 — less than 2% of the
completed-trip value.

The MOR estimate uses 152 of the 210 incomplete-trip interviews; 58 were truncated
because the angler had fished fewer than 30 minutes when contacted. Those very
short-trip interviews are dropped to avoid leveraging noisy per-trip ratios from
anglers who had barely started fishing.

The agreement between estimators for Harlan 2022 suggests that the catch process
at Harlan is approximately stationary. This supports using the completed-trip
ROM estimate as the primary result.

```{r}
#| label: fig-ch14-diagnostic
#| echo: true
#| fig-cap: "Complete vs incomplete trip CPUE estimates at Harlan Reservoir, 2022. The ratio-of-means estimate from completed trips (n = 390) and the truncated mean-of-ratios estimate from incomplete trips (n = 152 after removing trips < 30 min) differ by less than 2%, supporting the stationarity assumption. Error bars show 95% confidence intervals."
#| fig-alt: "Forest plot with two rows: Complete (n=390) and Incomplete (n=152). Complete trip point is near 1.51 with a narrow CI. Incomplete trip point is near 1.48 with a much wider CI. A vertical dashed reference line marks 1.51."
cpue_diag$comparison |>
  mutate(
    trip_type = str_to_title(trip_type),
    label     = sprintf("%.2f (n = %d)", estimate, n)
  ) |>
  ggplot(aes(x = estimate, y = trip_type)) +
  geom_vline(xintercept = 1.51, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                 height = 0.15, linewidth = 0.8) +
  geom_point(size = 3) +
  geom_text(aes(label = label), vjust = -0.8, size = 3.5) +
  labs(
    x = "CPUE (fish per angler-hour)",
    y = "Trip type",
    title = "Complete vs incomplete trip CPUE",
    caption = "Dashed line: complete-trip estimate (1.51 fish/hr). Error bars: 95% CI."
  ) +
  xlim(0.5, 2.5) +
  theme_minimal()
```

The diagnostic is especially important when the complete-trip sample is small.
At Harlan, with 65% complete trips, the diagnostic is confirmatory rather than
prescriptive.

## Harvest rate and release rate {#sec-cpue-hpue-rpue}

CPUE encompasses all fish contacted — harvested and released together. The
component rates are estimated separately by `estimate_harvest_rate()` and
`estimate_release_rate()`. These functions use all interviews, including incomplete
trips, because the current livewell contents and any fish already released are
observable regardless of whether the trip is finished.

```{r}
#| label: hpue-rpue
#| echo: true
hpue <- estimate_harvest_rate(harlan_design_ch14)
rpue <- estimate_release_rate(harlan_design_ch14)

bind_rows(
  hpue$estimates |> mutate(quantity = "Harvest rate (HPUE)"),
  rpue$estimates |> mutate(quantity = "Release rate (RPUE)")
) |>
  select(quantity, estimate, se, ci_lower, ci_upper, n)
```

Harvest rate is 0.600 fish per angler-hour (SE = 0.043); release rate is 0.917
fish per angler-hour (SE = 0.071). Both use all 600 interviews, unlike the CPUE
estimate that uses completed trips only.

A simple consistency check: if the catch partition is internally coherent, then
HPUE + RPUE should approximately equal CPUE. Here, 0.600 + 0.917 = 1.517, which
compares well to the CPUE of 1.51. The 0.7% discrepancy arises because CPUE uses
only completed trips while HPUE and RPUE use all interviews — different
denominators produce a small arithmetic disagreement that does not indicate an
error. The close agreement confirms that the interview records are internally
consistent.

The release fraction (RPUE / CPUE ≈ 0.917 / 1.51 ≈ 0.61) says that roughly 61%
of fish contacted at Harlan were released. The harvest fraction (0.600 / 1.51 ≈
0.40) feeds directly into harvest estimation when multiplied by total angler-hours.

```{r}
#| label: fig-ch14-rates
#| echo: true
#| fig-cap: "Catch, harvest, and release rates at Harlan Reservoir, 2022. CPUE uses completed-trip interviews only (n = 390); HPUE and RPUE use all interviews (n = 600). HPUE plus RPUE sum to 1.517, within 0.7% of CPUE, confirming internal consistency of the catch partition. Error bars show 95% confidence intervals."
#| fig-alt: "Three horizontal error bars arranged vertically. CPUE at top (~1.51) with a narrow CI. RPUE in middle (~0.917) with a moderate CI. HPUE at bottom (~0.600) with a narrow CI. A vertical dashed line marks the CPUE value of 1.51."
bind_rows(
  cpue_overall$estimates |> mutate(quantity = "CPUE (all catch)"),
  rpue$estimates          |> mutate(quantity = "RPUE (released)"),
  hpue$estimates          |> mutate(quantity = "HPUE (harvested)")
) |>
  mutate(quantity = factor(quantity,
    levels = c("CPUE (all catch)", "RPUE (released)", "HPUE (harvested)"))) |>
  ggplot(aes(x = estimate, y = quantity)) +
  geom_vline(xintercept = 1.51, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                 height = 0.15, linewidth = 0.8) +
  geom_point(size = 3) +
  geom_text(aes(label = sprintf("%.3f", estimate)), vjust = -0.8, size = 3.5) +
  labs(
    x = "Rate (fish per angler-hour)",
    y = NULL,
    title = "Catch, harvest, and release rates",
    caption = "Dashed line: CPUE baseline (1.51 fish/hr). Error bars: 95% CI."
  ) +
  xlim(0, 2.1) +
  theme_minimal()
```

## Species-level CPUE {#sec-cpue-by-species}

Management decisions often require species-specific catch rates: walleye CPUE
is relevant to minimum-size regulations, white bass CPUE to bag limits.
`estimate_catch_rate(by = species)` produces design-based species CPUE with
proper stratification weights and confidence intervals:

```{r}
#| label: cpue-by-species
#| echo: true
estimate_catch_rate(harlan_design_ch14, by = species)
```

White bass and walleye dominate the catch rate, consistent with the Harlan
fishery's reputation as a mixed-species reservoir with high white bass density
and a targetable walleye population. Freshwater drum, yellow perch, and channel
catfish register lower but meaningful rates. Species with near-zero estimates
are rarely targeted and largely absent from completed-trip catch records.

The standard errors reflect the same Taylor linearization variance used for
overall CPUE. Confidence intervals for rarer species are wide — for common
carp and gizzard shad they span zero — because few completed trips report
those species, making the ratio estimator unstable. Species-level CPUE is most
reliable when a species appears in at least 15–20% of completed trips.

## Comparing estimators {#sec-cpue-compare-estimators}

`compare_cpue_estimators()` runs all three supported CPUE estimators on the same design object and returns a tidy table, making it straightforward to assess estimator sensitivity:

```{r}
#| label: cpue-compare-estimators
#| echo: true
compare_cpue_estimators(harlan_design_ch14)
```

The three methods agree within approximately 20% of each other, which is typical
for access-point designs with moderate sample sizes.

**Ratio-of-means (ROM, default):** divides the total catch across all completed trips by the total effort. A single angler catching twenty fish in two hours contributes 20 fish and 2 hours to the numerator and denominator separately. The estimate is pulled toward trips with long effort because they contribute more hours.

**Mean-of-ratios (MOR):** computes each completed trip's catch rate independently and then takes the mean. Every trip gets equal weight regardless of duration. MOR is the natural estimator for roving surveys where mid-trip contacts are observations of instantaneous rate; here it is computed on completed trips for comparison.

**Regression CPUE:** fits a linear model of catch on effort and reads off the slope as the CPUE estimate. It is less sensitive to the few high-effort, high-catch trips that can inflate ROM.

For the Harlan access-point survey, the ROM estimate is the standard reporting
value. The MOR and regression estimates are useful robustness checks. Chapter 17
shows a roving survey where MOR is the primary estimator and ROM is the check.

## What CPUE implies and what it does not {#sec-cpue-interpretation}

CPUE is a ratio with a numerator (catch) and a denominator (effort). Both
components carry uncertainty, and that uncertainty propagates into the ratio.
The standard error produced by `estimate_catch_rate()` reflects that propagation
using Taylor linearization.

Several common misinterpretations deserve explicit attention.

**CPUE is not abundance.** A CPUE of 1.51 fish per angler-hour does not mean
there are 1.51 catchable fish per angler-hour of water. Catchability — the
probability that a fish is contacted given it is present — varies with gear type,
habitat use, season, and fish behavior. Creel surveys do not estimate catchability;
they estimate the rate experienced by anglers. Converting CPUE to abundance
requires catchability coefficients that are rarely available.

**CPUE comparisons require compatible designs.** A weekday CPUE from one year
is not directly comparable to a weekend CPUE from another year without accounting
for stratum composition.

**Higher CPUE does not equal better fishing.** Anglers may report high CPUE because
they targeted a species with dense schooling behavior, fished during a peak
season, or changed their technique.

**CPUE from completed trips excludes some anglers systematically.** The ROM
estimator uses only trips that ended before the survey period closed. Anglers
who fished past the survey closing time are excluded.

Reporting CPUE honestly means stating the estimator used, the sample size of
completed trips, the survey period and strata included, and a confidence interval.
A point estimate alone — "anglers caught 1.5 fish per hour" — strips the precision
context that distinguishes a well-supported finding from a noisy snapshot.

## Takeaway {#sec-cpue-takeaway}

CPUE is catch divided by effort, expressed as a ratio with an uncertainty
estimate. `tidycreel` implements two estimators: ratio-of-means for completed
trips (default) and mean-of-ratios for incomplete trips (diagnostic or roving
designs). At Harlan 2022, the overall completed-trip CPUE is 1.51 fish/hr
(95% CI: [1.32, 1.69]). Bank anglers show a higher rate than boat anglers, and
the complete vs incomplete diagnostic confirms stationarity.

The most important interpretation rule is also the simplest: report CPUE as a
rate observed under a specific design, with its confidence interval, for a
specific population of anglers in a specific time window.
