---
title: "CPUE Estimation"
---

```{r}
#| label: setup-ch14
#| include: false
library(tidycreel)
library(tidyverse)
library(lubridate)
library(here)

harlan_schedule   <- readRDS(here("data/harlan_schedule.rds"))
harlan_counts     <- readRDS(here("data/harlan_counts.rds"))
harlan_interviews <- readRDS(here("data/harlan_interviews.rds"))
harlan_catch      <- readRDS(here("data/harlan_catch.rds"))

shift_hours <- c("1" = 8L, "2" = 6L)

harlan_counts_ch14 <- harlan_counts |>
  mutate(
    period        = as.character(period),
    total_anglers = boat_anglers + bank_anglers,
    angler_hours  = total_anglers * shift_hours[period]
  ) |>
  select(reservoir, date, day_type, period, section, count_time, angler_hours)

catch_by_interview <- harlan_catch |>
  group_by(interview_id, catch_type) |>
  summarise(n_fish = sum(n_fish), .groups = "drop") |>
  pivot_wider(
    names_from  = catch_type,
    values_from = n_fish,
    values_fill = 0L
  ) |>
  mutate(total_catch = harvested + released)

harlan_interviews_ch14 <- harlan_interviews |>
  left_join(catch_by_interview, by = "interview_id") |>
  mutate(
    harvested   = if_else(is.na(harvested),   0L, as.integer(harvested)),
    released    = if_else(is.na(released),    0L, as.integer(released)),
    total_catch = if_else(is.na(total_catch), 0L, as.integer(total_catch))
  )

suppressMessages({
  harlan_design_ch14 <- creel_design(
    calendar = harlan_schedule,
    date     = "date",
    strata   = "day_type"
  ) |>
    add_counts(harlan_counts_ch14, count_time_col = "count_time") |>
    add_interviews(
      interviews     = harlan_interviews_ch14,
      catch          = "total_catch",
      harvest        = "harvested",
      effort         = "hours_fished",
      trip_duration  = "hours_fished",
      trip_status    = "trip_status",
      n_anglers      = "n_anglers",
      angler_type    = "angler_type",
      angler_method  = "angler_method",
      interview_type = "access"
    ) |>
    add_catch(
      data          = harlan_catch,
      catch_uid     = "interview_id",
      interview_uid = "interview_id",
      species       = "species",
      count         = "n_fish",
      catch_type    = "catch_type"
    )
})
```

## What CPUE is {#sec-cpue-what}

Catch-per-unit-effort (CPUE) compresses the fishing experience into a single ratio:
catch divided by effort. That compression is both its appeal and its main source of
misinterpretation.

In a roving or access-point creel survey, effort is measured in angler-hours —
the product of the number of anglers and the hours they fished. Catch is the
number of fish contacted over that same duration. The resulting ratio, in units of
fish per angler-hour, is a population-average catch rate: how many fish a
representative angler would encounter per hour under the survey conditions.

CPUE is not an index of abundance in the strict sense. A higher CPUE could reflect
better fishing conditions, a change in effort distribution, or an actual increase in
fish density. Without a stable relationship between catchability and true abundance,
CPUE should be labeled as the rate observed under the sampling design, not as a
direct measure of population size. The distinction matters most when comparing
CPUE across years, strata, or water bodies where survey conditions differ.

Within a single survey period, however, CPUE is exactly what management reporting
requires: an estimate of how productive the fishery appeared to anglers who
participated. That quantity, with its uncertainty, is what `tidycreel` computes
from interview records.

## Two ratio estimators {#sec-cpue-estimators}

There are two ways to form a ratio from interview data, and they answer subtly
different questions [@hoenig1997].

The **ratio-of-means** (ROM) estimator sums catch and effort separately across
all interviews, then divides:

$$\widehat{\text{CPUE}}_{\text{ROM}} = \frac{\sum_{i=1}^{n} c_i}{\sum_{i=1}^{n} e_i}$$

where $c_i$ is the total catch for interview $i$ and $e_i$ is the hours fished.
This estimator is weighted by trip length: anglers who fished longer contribute
more to both the numerator and the denominator. It is the natural estimator for
completed trips because the full trip record — final catch and full elapsed hours —
is available at the time of interview.

The **mean-of-ratios** (MOR) estimator computes a per-trip rate for each interview
first, then averages those rates:

$$\widehat{\text{CPUE}}_{\text{MOR}} = \frac{1}{n} \sum_{i=1}^{n} \frac{c_i}{e_i}$$

This gives each trip equal weight regardless of duration. It is the natural
estimator for incomplete trips, where the interview catches an angler mid-trip and
$c_i$ and $e_i$ are snapshots of a fishing session still in progress. If the catch
process is stationary — if the rate of catching fish does not change systematically
within a trip — then the MOR applied to incomplete trips should estimate the same
parameter as the ROM applied to completed trips.

The stationarity assumption is the critical qualifier. If anglers learn during a
trip (improving their catch rate over time), then early-trip snapshots will
underestimate the completed-trip rate. @hoenig1997 analyzed this under a learner
model and found that truncating very short trips — removing interviews where the
angler has fished fewer than 30 minutes — substantially reduced bias by excluding
anglers who had barely started and thus had few catches relative to their eventual
rate. The `estimator = "mortr"` option in `estimate_catch_rate()` applies this
truncation automatically.

@pollock1994 recommended using completed trips wherever possible because the ROM
estimator requires no stationarity assumption. This book follows that convention:
`estimate_catch_rate()` uses the ROM estimator on completed trips by default.

A practical diagnostic compares both estimators using the same dataset. If the
completed and incomplete estimates are close — within 10% of the complete-trip
value — the stationarity assumption is supported and incomplete-trip data can
supplement the analysis if sample sizes are small. If they diverge substantially,
the analyst should investigate whether the catch process is non-stationary before
using MOR-based estimates.

## The Harlan CPUE baseline {#sec-cpue-harlan-baseline}

The 2022 Harlan Reservoir creel survey collected 600 party interviews. Of those,
390 were completed trips (65%) and 210 were incomplete (35%). The ratio-of-means
estimator uses the 390 completed-trip records.

```{r}
#| label: cpue-overall
#| echo: true
cpue_overall <- estimate_catch_rate(harlan_design_ch14)
cpue_overall
```

The output tibble has five columns: `estimate` is the ratio-of-means CPUE,
`se` is the Taylor-linearized standard error, `ci_lower` and `ci_upper` are the
95% confidence interval bounds, and `n` is the number of completed-trip interviews
used. Every subsequent call to `estimate_catch_rate()` returns the same structure.

The overall CPUE estimate is 1.51 fish per angler-hour (95% CI: [1.32, 1.69]).
The standard error of 0.093 yields a CV of about 6.2%, which falls comfortably
below the 15% precision threshold used in Chapter 15. The 390 completed-trip
sample is adequate for a precise ratio estimate at the survey-wide level.

This estimate is the average catch rate for all anglers in all strata combined.
It answers the question: how many fish did a typical angler at Harlan encounter
per hour in 2022? It does not say anything about the total number of fish in the
reservoir, about whether fishing improved or declined from prior years, or about
whether any particular species drove the rate.

## CPUE by day-type stratum {#sec-cpue-by-stratum}

The survey design stratifies the calendar into weekdays and weekends. Weekend days
attract different angler populations — often more casual anglers or families —
and can have different crowding, access patterns, and fishing conditions than
weekdays. Stratum-specific CPUE separates those effects.

```{r}
#| label: cpue-by-daytype
#| echo: true
cpue_dt <- estimate_catch_rate(harlan_design_ch14, by = day_type)
cpue_dt$estimates
```

```{r}
#| label: fig-ch14-stratum-ci
#| echo: true
#| fig-cap: "CPUE by day-type stratum at Harlan Reservoir, 2022. Points show ratio-of-means estimates from completed-trip interviews; horizontal bars show 95% confidence intervals. Labels report the estimated rate and sample size. Weekend CPUE appears slightly higher but the wider confidence interval reflects the smaller completed-trip sample."
#| fig-alt: "Forest plot with two rows (Weekday and Weekend) on the y axis. Each row shows a point estimate and horizontal error bar for 95% confidence interval. Weekday CI is narrow (roughly 1.3–1.7); Weekend CI is wide (roughly 1.2–2.1). A vertical dashed reference line marks the overall estimate of 1.51."
cpue_dt$estimates |>
  mutate(
    day_type = str_to_title(day_type),
    label    = sprintf("%.2f (n = %d)", estimate, n)
  ) |>
  ggplot(aes(x = estimate, y = day_type)) +
  geom_vline(xintercept = 1.51, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                 height = 0.15, linewidth = 0.8) +
  geom_point(size = 3) +
  geom_text(aes(label = label), vjust = -0.8, size = 3.5) +
  labs(
    x = "CPUE (fish per angler-hour)",
    y = NULL,
    title = "CPUE by day-type stratum",
    caption = "Dashed line: overall estimate (1.51 fish/hr). Error bars: 95% CI."
  ) +
  xlim(0.9, 2.4) +
  theme_minimal()
```

Weekend CPUE (1.64 fish/hr) is slightly higher than weekday CPUE (1.48 fish/hr),
but the weekend estimate carries a much wider confidence interval because only 69
weekend interviews were completed trips, compared to 321 weekday interviews. The
weekend SE is 0.22, more than twice the weekday SE of 0.10. The weekend CI extends
from 1.21 to 2.08 — a range of nearly one fish per hour — while the weekday CI
spans only 0.40 units.

Whether the weekend and weekday rates are genuinely different is not resolved by
this data alone. The overlap in confidence intervals is substantial. A formal
comparison would require a pooled-variance test or a model-based contrast, but
for reporting purposes the key observation is that weekend estimates carry higher
uncertainty and should be interpreted with that caveat.

## CPUE by angler type {#sec-cpue-by-angler-type}

The interviews distinguish bank anglers from boat anglers. These two groups fish
differently: boat anglers can move to productive structure, cover more water, and
access mid-lake habitat; bank anglers are constrained to the shoreline. If the
fish community or the productive habitat is distributed unevenly, the two groups
should show different catch rates.

```{r}
#| label: cpue-by-anglertype
#| echo: true
cpue_at <- estimate_catch_rate(harlan_design_ch14, by = angler_type)
cpue_at$estimates
```

```{r}
#| label: fig-ch14-angler-type
#| echo: true
#| fig-cap: "CPUE by angler type at Harlan Reservoir, 2022. Bank anglers show a substantially higher catch rate than boat anglers, with overlapping confidence intervals only at the outer bounds. The difference reflects both access patterns and target species, not equivalent fishing conditions."
#| fig-alt: "Forest plot with two rows (Bank and Boat) on the y axis. Bank angler point is near 3.0 with a wide CI extending from about 1.8 to 4.1; Boat angler point is near 1.4 with a narrow CI from about 1.2 to 1.6. A vertical dashed reference line marks the overall estimate of 1.51."
cpue_at$estimates |>
  mutate(
    angler_type = str_to_title(angler_type),
    label       = sprintf("%.2f (n = %d)", estimate, n)
  ) |>
  ggplot(aes(x = estimate, y = angler_type)) +
  geom_vline(xintercept = 1.51, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                 height = 0.15, linewidth = 0.8) +
  geom_point(size = 3) +
  geom_text(aes(label = label), vjust = -0.8, size = 3.5) +
  labs(
    x = "CPUE (fish per angler-hour)",
    y = NULL,
    title = "CPUE by angler type",
    caption = "Dashed line: overall estimate (1.51 fish/hr). Error bars: 95% CI."
  ) +
  xlim(0, 5) +
  theme_minimal()
```

Bank anglers report 2.95 fish per angler-hour (95% CI: [1.83, 4.07]), more than
twice the boat angler rate of 1.41 fish/hr (95% CI: [1.23, 1.59]). The difference
is large and the confidence intervals do not overlap.

The interpretation requires care. It is tempting to conclude that bank fishing is
simply more productive, but the comparison conflates method with location and
target species. Bank anglers at Harlan skew toward white bass, which school near
the dam face and rocky banks — habitat that is highly accessible from shore and
episodically very productive when fish are surface feeding. Boat anglers are more
likely to be pursuing walleye in deeper mid-lake structure, a lower-frequency catch
but a more deliberately targeted species. The raw CPUE ratio reflects both the
biology and the behavior, not productivity of the water body alone.

The sample sizes also differ: only 71 bank-angler completed-trip interviews vs 319
for boat anglers. The bank angler estimate is less precise (SE = 0.57, CV ≈ 19%)
and should be interpreted in that light.

When reporting CPUE by subgroup, the analyst should always report the sample size
and CI alongside the point estimate. A 2.95 with a CI of [1.83, 4.07] is a
different level of evidence than a 1.41 with [1.23, 1.59]. Omitting the interval
suggests a precision that the data do not support.

## Validating the estimator: the complete vs incomplete diagnostic {#sec-cpue-diagnostic}

The choice between ROM and MOR matters most when incomplete trips are numerous
and when the analyst suspects non-stationary catch processes. `tidycreel` provides
a diagnostic mode that applies both estimators and compares the results.

```{r}
#| label: cpue-diagnostic
#| echo: true
cpue_diag <- estimate_catch_rate(harlan_design_ch14, use_trips = "diagnostic")
cpue_diag$comparison
```

The completed-trip estimate is 1.51 fish/hr; the incomplete-trip truncated MOR
estimate is 1.48 fish/hr. The difference is 0.027 — less than 2% of the
completed-trip value — and falls well within the 10% threshold used as a
stationarity benchmark [@hoenig1997]. The diagnostic interpretation returned by
the function confirms this: "Estimates are similar."

The MOR estimate uses 152 of the 210 incomplete-trip interviews; 58 were truncated
because the angler had fished fewer than 30 minutes when contacted. Those very
short-trip interviews are dropped to avoid leveraging noisy per-trip ratios from
anglers who had barely started fishing.

The agreement between estimators for Harlan 2022 is reassuring. It suggests that
the catch process at Harlan is approximately stationary: anglers were not
systematically more or less successful later in their trips relative to earlier.
This supports using the completed-trip ROM estimate as the primary result without
adjustment.

```{r}
#| label: fig-ch14-diagnostic
#| echo: true
#| fig-cap: "Complete vs incomplete trip CPUE estimates at Harlan Reservoir, 2022. The ratio-of-means estimate from completed trips (n = 390) and the truncated mean-of-ratios estimate from incomplete trips (n = 152 after removing trips < 30 min) differ by less than 2%, supporting the stationarity assumption. Error bars show 95% confidence intervals."
#| fig-alt: "Forest plot with two rows: Complete (n=390) and Incomplete (n=152). Complete trip point is near 1.51 with a narrow CI. Incomplete trip point is near 1.48 with a much wider CI. A vertical dashed reference line marks 1.51."
cpue_diag$comparison |>
  mutate(
    trip_type = str_to_title(trip_type),
    label     = sprintf("%.2f (n = %d)", estimate, n)
  ) |>
  ggplot(aes(x = estimate, y = trip_type)) +
  geom_vline(xintercept = 1.51, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                 height = 0.15, linewidth = 0.8) +
  geom_point(size = 3) +
  geom_text(aes(label = label), vjust = -0.8, size = 3.5) +
  labs(
    x = "CPUE (fish per angler-hour)",
    y = "Trip type",
    title = "Complete vs incomplete trip CPUE",
    caption = "Dashed line: complete-trip estimate (1.51 fish/hr). Error bars: 95% CI."
  ) +
  xlim(0.5, 2.5) +
  theme_minimal()
```

The diagnostic is especially important when the complete-trip sample is small.
If fewer than 30% of interviews are completed trips, the ROM estimate may be
imprecise and the analyst should explicitly test whether MOR-based supplementation
would improve precision without introducing systematic bias. At Harlan, with 65%
complete trips, the diagnostic is confirmatory rather than diagnostic in a strict
sense — it validates the design, not rescues it.

## Harvest rate and release rate {#sec-cpue-hpue-rpue}

CPUE encompasses all fish contacted — harvested and released together. The
component rates are estimated separately by `estimate_harvest_rate()` and
`estimate_release_rate()`. These functions use all interviews, including incomplete
trips, because the current livewell contents and any fish already released are
observable regardless of whether the trip is finished.

```{r}
#| label: hpue-rpue
#| echo: true
hpue <- estimate_harvest_rate(harlan_design_ch14)
rpue <- estimate_release_rate(harlan_design_ch14)

bind_rows(
  hpue$estimates |> mutate(quantity = "Harvest rate (HPUE)"),
  rpue$estimates |> mutate(quantity = "Release rate (RPUE)")
) |>
  select(quantity, estimate, se, ci_lower, ci_upper, n)
```

Harvest rate is 0.600 fish per angler-hour (SE = 0.043); release rate is 0.917
fish per angler-hour (SE = 0.071). Both use all 600 interviews (n = 600), which
is why the sample sizes differ from the CPUE n of 390.

A simple consistency check: if the catch partition is internally coherent, then
HPUE + RPUE should approximately equal CPUE. Here, 0.600 + 0.917 = 1.517, which
compares well to the CPUE of 1.51. The 0.7% discrepancy arises because CPUE uses
only completed trips while HPUE and RPUE use all interviews — different
denominators produce a small arithmetic disagreement that does not indicate an
error. The close agreement confirms that the interview records are internally
consistent.

The release fraction (RPUE / CPUE ≈ 0.917 / 1.51 ≈ 0.61) says that roughly 61%
of fish contacted at Harlan were released. This is a useful planning number: if
an angler at Harlan catches 10 fish in a session, about 6 are expected to go back.
The harvest fraction (0.600 / 1.51 ≈ 0.40) reflects the direct extraction rate
and feeds directly into harvest estimation when multiplied by total angler-hours
(see Chapter 13).

```{r}
#| label: fig-ch14-rates
#| echo: true
#| fig-cap: "Catch, harvest, and release rates at Harlan Reservoir, 2022. CPUE uses completed-trip interviews only (n = 390); HPUE and RPUE use all interviews (n = 600). HPUE plus RPUE sum to 1.517, within 0.7% of CPUE, confirming internal consistency of the catch partition. Error bars show 95% confidence intervals."
#| fig-alt: "Three horizontal error bars arranged vertically. CPUE at top (~1.51) with a narrow CI. RPUE in middle (~0.917) with a moderate CI. HPUE at bottom (~0.600) with a narrow CI. A vertical dashed line marks the CPUE value of 1.51."
bind_rows(
  cpue_overall$estimates |> mutate(quantity = "CPUE (all catch)"),
  rpue$estimates          |> mutate(quantity = "RPUE (released)"),
  hpue$estimates          |> mutate(quantity = "HPUE (harvested)")
) |>
  mutate(quantity = factor(quantity,
    levels = c("CPUE (all catch)", "RPUE (released)", "HPUE (harvested)"))) |>
  ggplot(aes(x = estimate, y = quantity)) +
  geom_vline(xintercept = 1.51, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                 height = 0.15, linewidth = 0.8) +
  geom_point(size = 3) +
  geom_text(aes(label = sprintf("%.3f", estimate)), vjust = -0.8, size = 3.5) +
  labs(
    x = "Rate (fish per angler-hour)",
    y = NULL,
    title = "Catch, harvest, and release rates",
    caption = "Dashed line: CPUE baseline (1.51 fish/hr). Error bars: 95% CI."
  ) +
  xlim(0, 2.1) +
  theme_minimal()
```

## Species-level CPUE {#sec-cpue-by-species}

Management decisions often require species-specific catch rates: walleye CPUE
is relevant to minimum-size regulations, white bass CPUE to bag limits. The
natural call is `estimate_catch_rate(by = species)`, but a known issue in the
current version of `tidycreel` (GitHub issue #64) causes this to return zeros
for all species. The internal species-catch lookup matches on a catch type label
that does not align with the field values in the Harlan dataset.

```{r}
#| label: cpue-species-issue
#| echo: true
estimate_catch_rate(harlan_design_ch14, by = species)
```

All estimates are zero. This is not a property of the Harlan data; it is a package
limitation. Until issue #64 is resolved, species-level CPUE must be approximated
from the raw catch table rather than from the estimation pipeline.

The raw approach computes an approximate catch rate per species by pooling catch
and effort across completed-trip interviews:

```{r}
#| label: species-cpue-approx
#| echo: true
species_catch <- harlan_catch |>
  filter(catch_type %in% c("harvested", "released")) |>
  group_by(interview_id, species) |>
  summarise(n_fish = sum(n_fish), .groups = "drop")

completed_ids <- harlan_interviews |>
  filter(trip_status == "complete") |>
  pull(interview_id)

species_catch |>
  filter(interview_id %in% completed_ids) |>
  left_join(
    harlan_interviews |> select(interview_id, hours_fished),
    by = "interview_id"
  ) |>
  group_by(species) |>
  summarise(
    total_fish  = sum(n_fish),
    total_hours = sum(hours_fished),
    approx_cpue = round(total_fish / total_hours, 3),
    n_interviews = n_distinct(interview_id),
    .groups = "drop"
  ) |>
  arrange(desc(approx_cpue))
```

This is a descriptive approximation, not a design-based estimate. It ignores
stratification weights and does not produce a standard error. The values should
be read as a relative ranking — white bass dominate the catch rate, walleye are
the second most frequently encountered species — rather than as calibrated
population estimates.

When issue #64 is resolved in a future package version,
`estimate_catch_rate(by = species)` will replace this workaround and produce
design-based species CPUE with proper confidence intervals.

## What CPUE implies and what it does not {#sec-cpue-interpretation}

CPUE is a ratio with a numerator (catch) and a denominator (effort). Both
components carry uncertainty, and that uncertainty propagates into the ratio.
The standard error produced by `estimate_catch_rate()` reflects that propagation
using Taylor linearization, which is appropriate for smooth statistics like ratios
under the assumption that both components are large enough relative to their
variability.

Several common misinterpretations deserve explicit attention.

**CPUE is not abundance.** A CPUE of 1.51 fish per angler-hour does not mean
there are 1.51 catchable fish per angler-hour of water. Catchability — the
probability that a fish is contacted given it is present — varies with gear type,
habitat use, season, and fish behavior. Creel surveys do not estimate catchability;
they estimate the rate experienced by anglers. Converting CPUE to abundance
requires catchability coefficients that are rarely available.

**CPUE comparisons require compatible designs.** A weekday CPUE from one year
is not directly comparable to a weekend CPUE from another year without accounting
for stratum composition. If the proportion of weekday surveys changes between years,
the pooled CPUE changes even if neither weekday nor weekend rates change. Comparing
overall rates across years is only defensible if the survey design and stratum
weights are held constant, or if stratum-specific rates are compared directly.

**Higher CPUE does not equal better fishing.** Anglers may report high CPUE because
they targeted a species with dense schooling behavior, fished during a peak
season, or changed their technique. Fishery health depends on biomass, age
structure, and recruitment — quantities that CPUE alone cannot reveal.

**CPUE from completed trips excludes some anglers systematically.** The ROM
estimator uses only trips that ended before the survey period closed. Anglers
who fished past the survey closing time are excluded. If late-evening fishing
has different catch rates — common when fish are crepuscular — the completed-trip
estimate may be slightly biased toward daytime conditions. The diagnostic check
(Section [-@sec-cpue-diagnostic]) helps assess the magnitude of this effect.

Reporting CPUE honestly means stating the estimator used, the sample size of
completed trips, the survey period and strata included, and a confidence interval.
A point estimate alone — "anglers caught 1.5 fish per hour" — strips the precision
context that distinguishes a well-supported finding from a noisy snapshot.

## Takeaway {#sec-cpue-takeaway}

CPUE is catch divided by effort, expressed as a ratio with an uncertainty
estimate. `tidycreel` implements two estimators: ratio-of-means for completed
trips (default) and mean-of-ratios for incomplete trips (diagnostic or roving
designs). At Harlan 2022, the overall completed-trip CPUE is 1.51 fish/hr
(95% CI: [1.32, 1.69]). Bank anglers show a markedly higher rate than boat anglers,
a difference that reflects target species and habitat access rather than overall
productivity. The complete vs incomplete diagnostic confirms stationarity: both
estimators produce estimates within 2% of each other.

The most important interpretation rule is also the simplest: report CPUE as a
rate observed under a specific design, with its confidence interval, for a specific
population of anglers in a specific time window. That framing prevents the ratio
from being overread as a universal property of the fishery.
