---
title: "Uncertainty, Diagnostics, and Precision"
---

```{r}
#| label: setup-ch15
#| include: false
library(tidycreel)
library(tidyverse)
library(here)

harlan_schedule   <- readRDS(here("data/harlan_schedule.rds"))
harlan_counts     <- readRDS(here("data/harlan_counts.rds"))
harlan_interviews <- readRDS(here("data/harlan_interviews.rds"))
harlan_catch      <- readRDS(here("data/harlan_catch.rds"))

# Shift lengths (hours): period 1 = AM (8 h), period 2 = PM (6 h)
shift_hours <- c("1" = 8L, "2" = 6L)

harlan_counts_effort <- harlan_counts |>
  mutate(
    period        = as.character(period),
    total_anglers = boat_anglers + bank_anglers,
    angler_hours  = total_anglers * shift_hours[period]
  ) |>
  select(reservoir, date, day_type, period, section, count_time, angler_hours)

catch_by_interview <- harlan_catch |>
  group_by(interview_id, catch_type) |>
  summarise(n_fish = sum(n_fish), .groups = "drop") |>
  pivot_wider(names_from = catch_type, values_from = n_fish, values_fill = 0L) |>
  mutate(total_catch = harvested + released)

harlan_interviews_with_catch <- harlan_interviews |>
  left_join(catch_by_interview, by = "interview_id") |>
  mutate(
    harvested   = if_else(is.na(harvested),   0L, as.integer(harvested)),
    released    = if_else(is.na(released),    0L, as.integer(released)),
    total_catch = if_else(is.na(total_catch), 0L, as.integer(total_catch))
  )

suppressMessages({
  harlan_design <- creel_design(
    calendar = harlan_schedule,
    date     = "date",
    strata   = "day_type"
  ) |>
    add_counts(harlan_counts_effort, count_time_col = "count_time") |>
    add_interviews(
      interviews     = harlan_interviews_with_catch,
      catch          = "total_catch",
      effort         = "hours_fished",
      trip_duration  = "hours_fished",
      trip_status    = "trip_status",
      n_anglers      = "n_anglers",
      angler_type    = "angler_type",
      angler_method  = "angler_method",
      interview_type = "access"
    ) |>
    add_catch(
      data          = harlan_catch,
      catch_uid     = "interview_id",
      interview_uid = "interview_id",
      species       = "species",
      count         = "n_fish",
      catch_type    = "catch_type"
    )
})
```

A creel estimate without uncertainty information is only half an answer. The
point estimate tells you how large the catch or effort was; the uncertainty
wrapper tells you how much faith to put in that number and whether the design
was precise enough to support the decisions that will follow. This chapter
covers where uncertainty comes from, how to communicate it, and how much
sampling is needed to hit a precision target.

## The anatomy of an estimate {#sec-uncertainty-anatomy}

When `estimate_effort()` runs, it returns not just the effort total but a full
uncertainty wrapper: a standard error, a confidence interval, and a decomposition
of variance into between-day and within-day components. Each of those pieces
carries distinct diagnostic value.

```{r}
#| label: effort-full
#| echo: true

eff <- suppressWarnings(estimate_effort(harlan_design))
eff$estimates
```

The season total for Harlan 2022 is
`r formatC(round(eff$estimates$estimate), big.mark = ",")` angler-hours,
with a standard error of
`r formatC(round(eff$estimates$se), big.mark = ",")` angler-hours.
The coefficient of variation — SE divided by the estimate — is
`r round(eff$estimates$se / eff$estimates$estimate * 100, 1)`%, which falls
well within the 15–20% threshold commonly used as an acceptable precision
benchmark for management-grade creel estimates [@mccormick2017].

```{r}
#| label: effort-cv
#| echo: true

eff$estimates |>
  mutate(cv = se / estimate) |>
  select(estimate, se, cv, ci_lower, ci_upper, n)
```

The confidence interval runs from
`r formatC(round(eff$estimates$ci_lower), big.mark = ",")` to
`r formatC(round(eff$estimates$ci_upper), big.mark = ",")` angler-hours.

## Where variance comes from {#sec-uncertainty-decomp}

Effort variance decomposes into two parts: variation between sampled days
and variation among instantaneous counts within a day. The `se_between` and
`se_within` columns in the estimates tibble expose this decomposition directly.

```{r}
#| label: variance-decomp-data
#| echo: true

eff$estimates |>
  select(se_between, se_within, se)
```

Between-day variance accounts for
`r round(eff$estimates$se_between^2 / eff$estimates$se^2 * 100)`% of total
variance at Harlan. Day-to-day differences in how many anglers arrive are a
larger source of uncertainty than moment-to-moment fluctuation within a shift.
That is the expected pattern for reservoir access-point surveys.

```{r}
#| label: fig-ch15-variance-decomp
#| fig-cap: "Variance decomposition for Harlan Reservoir 2022 season effort. Between-day variance (dark) dominates within-day variance (light), indicating that adding sampling days would reduce uncertainty more than increasing counts per day."
#| fig-alt: "Stacked bar chart showing se_between and se_within contributions to total effort SE"

eff$estimates |>
  select(se_between, se_within) |>
  pivot_longer(everything(), names_to = "component", values_to = "se") |>
  mutate(
    component = recode(component,
      "se_between" = "Between-day",
      "se_within"  = "Within-day"
    ),
    component = fct_relevel(component, "Between-day")
  ) |>
  ggplot(aes(x = "Season total", y = se^2, fill = component)) +
  geom_col(width = 0.4) +
  scale_fill_manual(
    values = c("Between-day" = "#2c6e49", "Within-day" = "#95d5b2"),
    name   = "Variance component"
  ) +
  scale_y_continuous(
    labels = scales::comma,
    name   = "Variance (angler-hours²)"
  ) +
  labs(x = NULL, title = "Effort variance decomposition") +
  theme_creel() +
  theme(legend.position = "bottom")
```

For designs where within-day variance dominates instead, the prescription is
more counts per sampled day.

## Precision by stratum {#sec-uncertainty-stratum}

Reporting a single season CV obscures whether both weekday and weekend strata
are meeting precision targets. `estimate_effort()` accepts a `by` argument to
produce stratum-specific estimates with their own uncertainty wrappers.

```{r}
#| label: effort-by-stratum
#| echo: true

eff_strata <- suppressWarnings(
  estimate_effort(harlan_design, by = "day_type")
)
eff_strata$estimates
```

Weekend effort at Harlan carries a higher CV than weekday effort. Both strata
meet the 15% precision target.

```{r}
#| label: fig-ch15-stratum-ci
#| fig-cap: "Effort estimates with 95% confidence intervals by day type, Harlan Reservoir 2022. Whiskers span the 95% CI returned by `estimate_effort(by = 'day_type')`. Weekend effort is substantially lower but also less precisely estimated."
#| fig-alt: "Bar chart with error whiskers showing effort CI for weekday and weekend strata"

eff_strata$estimates |>
  mutate(
    day_type = str_to_title(day_type),
    cv       = se / estimate
  ) |>
  ggplot(aes(x = day_type, y = estimate, fill = day_type)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    width = 0.15,
    linewidth = 0.8
  ) +
  geom_text(
    aes(
      y     = ci_upper + 300,
      label = paste0("CV = ", round(cv * 100, 1), "%")
    ),
    size = 3.5
  ) +
  scale_y_continuous(labels = scales::comma, name = "Angler-hours") +
  scale_fill_manual(values = c("Weekday" = "#2c6e49", "Weekend" = "#95d5b2")) +
  labs(x = "Day type", title = "Effort by stratum with 95% CI") +
  theme_creel()
```

## Precision reporting in practice {#sec-uncertainty-reporting}

Managers and readers need three quantities: the point estimate, the uncertainty
range, and a unitless relative measure they can compare across estimates, species,
and years. A clean precision table assembles all three.

```{r}
#| label: precision-table
#| echo: true

cpue <- suppressWarnings(estimate_catch_rate(harlan_design))

bind_rows(
  eff$estimates  |> mutate(quantity = "Effort (angler-hours)"),
  cpue$estimates |> mutate(quantity = "CPUE (fish/angler-hr)")
) |>
  mutate(cv = round(se / estimate, 3)) |>
  select(quantity, estimate, se, cv, ci_lower, ci_upper, n)
```

The CPUE estimate of
`r round(cpue$estimates$estimate, 3)` fish per angler-hour carries a tighter CV
than the effort estimate because it is based on interview data alone.

Communicating precision to non-technical audiences calls for a confidence
interval in the units of the estimate rather than the SE alone.

## Sampling coverage as a diagnostic {#sec-uncertainty-coverage}

The variance estimates produced by `estimate_effort()` describe sampling
variability under the realized design. They cannot detect a different problem:
systematic gaps in which days or months were actually sampled. A month where
no weekday samples were collected is not a precision problem — it is a
structural gap that biases the stratum mean in an unknowable direction.

Coverage is straightforward to compute: compare the days that appear in the
count data against the full season frame from the schedule.

```{r}
#| label: coverage-by-month
#| echo: true

coverage <- harlan_counts_effort |>
  distinct(date, day_type) |>
  mutate(month = lubridate::month(date, label = TRUE)) |>
  count(month, day_type, name = "n_sampled") |>
  left_join(
    harlan_schedule |>
      mutate(month = lubridate::month(date, label = TRUE)) |>
      count(month, day_type, name = "n_scheduled"),
    by = c("month", "day_type")
  ) |>
  mutate(pct_sampled = n_sampled / n_scheduled)

coverage
```

Coverage at Harlan ranges from
`r round(min(coverage$pct_sampled) * 100)`% to
`r round(max(coverage$pct_sampled) * 100)`% across month-stratum combinations,
with no complete gaps.

```{r}
#| label: fig-ch15-coverage
#| fig-cap: "Sampling coverage (proportion of scheduled days actually sampled) by month and day type, Harlan Reservoir 2022. Bars show the fraction of frame days in each cell that received at least one count. The dashed line at 50% is the approximate target under the Harlan design. No cell falls below 44%, indicating no structural coverage gaps."
#| fig-alt: "Grouped bar chart showing sampling coverage by month for weekday and weekend strata"

coverage |>
  mutate(day_type = str_to_title(day_type)) |>
  ggplot(aes(x = month, y = pct_sampled, fill = day_type)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_hline(yintercept = 0.50, linetype = "dashed", color = "grey40") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1),
                     name = "Proportion of frame days sampled") +
  scale_fill_manual(
    values = c("Weekday" = "#2c6e49", "Weekend" = "#95d5b2"),
    name = "Day type"
  ) +
  labs(x = "Month", title = "Sampling coverage by month and day type") +
  theme_creel() +
  theme(legend.position = "bottom")
```

When coverage varies widely across months, those months contribute a less
precise stratum mean to the seasonal total.

## How many sampling days are enough? {#sec-precision-samplesize}

Before the next field season, the survey coordinator needs to answer a practical
question: how many sampling days would be required to hit a target CV? This is
a design-forward question that the current estimates can inform retrospectively,
because the sample mean and sample variance from the completed survey are the
best available estimates of the population parameters needed for sample size
calculations.

The strata parameters required for `creel_n_effort()` are the stratum frame
sizes $N_h$ (total days in the season calendar by day type), the stratum
means $\bar{y}_h$ (mean daily angler-hours from the sample), and the stratum
variances $s^2_h$.

```{r}
#| label: strata-params
#| echo: true

# Aggregate daily angler-hours from count data; frame size from schedule
strata_params <- harlan_counts_effort |>
  group_by(date, day_type) |>
  summarise(daily_total = sum(angler_hours), .groups = "drop") |>
  group_by(day_type) |>
  summarise(
    n_sampled = n(),
    ybar_h    = mean(daily_total),
    s2_h      = var(daily_total),
    .groups   = "drop"
  ) |>
  left_join(
    harlan_schedule |> count(day_type, name = "N_h"),
    by = "day_type"
  )

strata_params
```

With those parameters in hand, `creel_n_effort()` solves for the sample size
allocation that minimizes variance subject to a CV target, using Neyman
proportional allocation across strata.

```{r}
#| label: n-required
#| echo: true

N_h    <- setNames(strata_params$N_h,    strata_params$day_type)
ybar_h <- setNames(strata_params$ybar_h, strata_params$day_type)
s2_h   <- setNames(strata_params$s2_h,   strata_params$day_type)

# Required n at three CV targets
bind_rows(
  creel_n_effort(cv_target = 0.10, N_h = N_h, ybar_h = ybar_h, s2_h = s2_h) |>
    as.list() |> as_tibble() |> mutate(cv_target = "10%"),
  creel_n_effort(cv_target = 0.15, N_h = N_h, ybar_h = ybar_h, s2_h = s2_h) |>
    as.list() |> as_tibble() |> mutate(cv_target = "15%"),
  creel_n_effort(cv_target = 0.20, N_h = N_h, ybar_h = ybar_h, s2_h = s2_h) |>
    as.list() |> as_tibble() |> mutate(cv_target = "20%")
) |>
  select(cv_target, weekday, weekend, total)
```

The Harlan 2022 design sampled 83 weekdays and 35 weekends, or 118 days total.
Hitting a 15% CV requires only 32 weekday days and 14 weekend days under
Neyman allocation.

## Validating the current design {#sec-precision-validate}

`validate_design()` takes the proposed or actual sample size and returns a
pass/fail verdict for each stratum against a stated CV target.

```{r}
#| label: validate-design
#| echo: true

design_report <- validate_design(
  N_h        = N_h,
  ybar_h     = ybar_h,
  s2_h       = s2_h,
  n_proposed = c(weekday = 83L, weekend = 35L),
  cv_target  = 0.15,
  type       = "effort"
)

design_report$results
```

Both strata pass against the 15% target. The report also records the minimum n
required for each stratum.

The design report also serves a documentation purpose: it shows whether the
sampling plan met its stated precision goals.

## Power to detect change {#sec-precision-power}

Precision addresses the current survey; power addresses future comparisons.
A fishery manager who wants to detect a 20% change in fishing effort between
years needs to know whether the current design is sensitive enough to distinguish
real change from sampling noise.

`creel_power()` computes the probability that a two-sample test will correctly
reject the null hypothesis of no change, given the historical CV of the
estimate, the proposed sample size, and the magnitude of the true difference
to detect.

```{r}
#| label: power-single
#| echo: true

# At current sample size, what power to detect a 30% change?
# Using CV = 0.30 to represent a more variable (e.g., multi-year) context
creel_power(
  n             = 118,
  cv_historical = 0.30,
  delta_pct     = 0.30,
  alpha         = 0.05
)
```

With a historical CV of 30% and 118 sampling days per year, power to detect
a 30% change exceeds 90%.

```{r}
#| label: fig-ch15-power-curve
#| fig-cap: "Statistical power to detect an effort change as a function of sample size and effect size, assuming three levels of historical CV. Each line shows the power curve for a given CV at a 5% significance threshold. The dashed line marks 80% power, the conventional minimum for planned comparisons."
#| fig-alt: "Line chart showing power curves for three historical CVs (0.15, 0.30, 0.50) across sample sizes 10 to 150"

crossing(
  n             = seq(10, 150, by = 5),
  cv_historical = c(0.15, 0.30, 0.50),
  delta_pct     = 0.30
) |>
  mutate(
    power    = pmap_dbl(list(n, cv_historical, delta_pct),
                        ~ creel_power(..1, ..2, ..3, alpha = 0.05)),
    cv_label = paste0("CV = ", cv_historical * 100, "%")
  ) |>
  ggplot(aes(x = n, y = power, color = cv_label, linetype = cv_label)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey50") +
  annotate("text", x = 15, y = 0.82, label = "80% power", hjust = 0,
           size = 3.2, color = "grey40") +
  scale_color_manual(
    values = c("CV = 15%" = "#2c6e49", "CV = 30%" = "#74c69d",
               "CV = 50%" = "#b7e4c7"),
    name = "Historical CV"
  ) +
  scale_linetype_manual(
    values = c("CV = 15%" = "solid", "CV = 30%" = "dashed",
               "CV = 50%" = "dotted"),
    name = "Historical CV"
  ) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    x     = "Sample size (days per year)",
    y     = "Power",
    title = "Power to detect 30% effort change"
  ) +
  theme_creel() +
  theme(legend.position = "bottom")
```

For a stable reservoir fishery like Harlan with an observed CV of 12%, 80%
power to detect a 30% change requires fewer than 20 sampling days per year.
For a more variable fishery, the same threshold requires far more days.

`power_creel()` provides a unified interface for the same computation
expressed as a table, which is more convenient when generating design
recommendations for a report.

```{r}
#| label: power-table
#| echo: true

# Sample size needed for 80% power across effect sizes
crossing(delta_pct = c(0.10, 0.20, 0.30, 0.40, 0.50)) |>
  mutate(
    n_for_80pct = map_dbl(delta_pct, function(d) {
      # Find minimum n achieving >= 0.80 power
      n_seq <- seq(5, 300, by = 5)
      powers <- sapply(n_seq, function(n)
        creel_power(n, cv_historical = 0.30, delta_pct = d))
      n_seq[which(powers >= 0.80)[1]]
    }),
    delta_label = paste0(delta_pct * 100, "%")
  ) |>
  select(delta_label, n_for_80pct) |>
  rename("Change to detect" = delta_label, "Days required (CV = 30%)" = n_for_80pct)
```

## CV sensitivity to interview sample size {#sec-precision-cpue-sensitivity}

For CPUE estimates, the relevant sample size is the number of completed
interviews rather than the number of sampling days. `cv_from_n()` with
`type = "cpue"` shows how CPUE precision responds to interview sample size
given input CVs for the catch and effort components. This is a planning tool:
calibrate the input parameters using historical data or the current survey's
observed variability, then read off the interview target.

The Harlan 2022 season produced a CPUE estimate of 1.51 fish/angler-hour
from 390 complete interviews (CV = 6.2%). Individual catch rates per interview
have high variability — many zero-catch anglers and a few high-count outliers
— so `cv_catch` for planning purposes is substantially larger than the
aggregate CPUE CV. A value of 1.2 (120%) for the individual catch-rate CV
is consistent with the observed aggregate precision at Harlan and typical of
reservoir access-point fisheries where zero-catch encounters are common.

```{r}
#| label: cv-cpue-sensitivity
#| echo: true

# cv_catch calibrated so the curve passes near the observed 6.2% CV at n = 390
cv_catch  <- 1.2
cv_effort <- round(eff$estimates$se / eff$estimates$estimate, 2)

tibble(n_interviews = seq(50, 450, by = 25)) |>
  mutate(
    cv_cpue = map_dbl(n_interviews, \(n)
      cv_from_n(type = "cpue", n = n,
                cv_catch = cv_catch, cv_effort = cv_effort))
  )
```

```{r}
#| label: fig-ch15-cv-sensitivity
#| fig-cap: "Estimated CPUE coefficient of variation as a function of interview sample size. Input CVs are calibrated to match the observed 6.2% CV at the Harlan 2022 sample size of 390 interviews (dashed vertical line). The horizontal dotted line marks 15% CV. Returns to precision become negligible beyond roughly 100 interviews for this level of variability."
#| fig-alt: "Line chart showing CPUE CV declining with interview sample size, with reference lines at n=390 and CV=15%"

tibble(n_interviews = seq(20, 500, by = 10)) |>
  mutate(
    cv_cpue = map_dbl(n_interviews, \(n)
      cv_from_n(type = "cpue", n = n,
                cv_catch = cv_catch, cv_effort = cv_effort))
  ) |>
  ggplot(aes(x = n_interviews, y = cv_cpue)) +
  geom_line(color = "#2c6e49", linewidth = 1) +
  geom_vline(xintercept = 390, linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = 0.15, linetype = "dotted", color = "#d62828") +
  annotate("text", x = 400, y = 0.11, label = "2022 survey\n(n = 390)",
           hjust = 0, size = 3.2, color = "grey40") +
  annotate("text", x = 25, y = 0.155, label = "CV = 15%",
           hjust = 0, size = 3.2, color = "#d62828") +
  scale_y_continuous(labels = scales::percent, name = "CPUE CV") +
  labs(
    x     = "Interview sample size",
    title = "CPUE precision as a function of interview sample size"
  ) +
  theme_creel()
```

The curve shows sharply diminishing returns. The 15% CV threshold is crossed
near 65 interviews, well below the 390 the 2022 survey collected.

For a new reservoir without historical data, `cv_from_n()` can be used
prospectively by borrowing `cv_catch` and `cv_effort` from a comparable fishery.

## Diagnostics alongside uncertainty {#sec-uncertainty-diagnostics}

Uncertainty quantifies sampling error. Diagnostics assess whether the data
structure behind the estimate is sound enough to trust it. A precise estimate
built on a biased or incomplete frame is a confident wrong answer.

Three diagnostic questions are worth asking before reporting any estimate:

**Did coverage match the design?** Compare sampled days to the frame.
Systematic gaps — no weekday samples in July, no AM periods in certain
months — create structural bias that the variance estimator cannot detect.

**Are any strata under-replicated?** A stratum with two or three sampled days
cannot support a reliable variance estimate. The `validate_design()` output
identifies strata where the proposed $n$ fell short of the required minimum;
`audit_strata()` (Chapter 10) flags strata present in the frame but absent
from the count data.

**Are any estimates driven by a small number of records?** A single high-harvest
day or an outlier interview can dominate a stratum mean. `flag_outliers()`
(Chapter 11) identifies those records before estimation. Including a sensitivity
note — "the harvest estimate would change by X% if the three highest-count days
were treated as anomalies" — is more informative than the interval alone when
the estimate is leverage-heavy.

These diagnostics belong in the same section of the survey report as the
precision table.

## Takeaway {#sec-uncertainty-takeaway}

Every estimate produced by a creel survey has two parts: the point estimate
and its uncertainty wrapper. `estimate_effort()` and `estimate_catch_rate()`
return both in a single call.

`creel_n_effort()`, `validate_design()`, and `creel_power()` connect the
completed survey to the next season's design.

The coefficient of variation is the most useful single summary of precision
for management communication. A CV below 15% is generally adequate for season-
total effort and harvest; a CV below 20% is acceptable for CPUE when interviews
are limited.
