---
title: "Reporting and Interpretation"
---

```{r}
#| label: setup-ch16
#| include: false
library(tidycreel)
library(tidyverse)
library(lubridate)
library(here)

harlan_schedule   <- readRDS(here("data/harlan_schedule.rds"))
harlan_counts     <- readRDS(here("data/harlan_counts.rds"))
harlan_interviews <- readRDS(here("data/harlan_interviews.rds"))
harlan_catch      <- readRDS(here("data/harlan_catch.rds"))

shift_hours <- c("1" = 8L, "2" = 6L)

harlan_counts_ch16 <- harlan_counts |>
  mutate(
    period        = as.character(period),
    total_anglers = boat_anglers + bank_anglers,
    angler_hours  = total_anglers * shift_hours[period]
  ) |>
  select(reservoir, date, day_type, period, section, count_time, angler_hours)

catch_by_interview <- harlan_catch |>
  group_by(interview_id, catch_type) |>
  summarise(n_fish = sum(n_fish), .groups = "drop") |>
  pivot_wider(
    names_from  = catch_type,
    values_from = n_fish,
    values_fill = 0L
  ) |>
  mutate(total_catch = harvested + released)

harlan_interviews_ch16 <- harlan_interviews |>
  left_join(catch_by_interview, by = "interview_id") |>
  mutate(
    harvested   = if_else(is.na(harvested),   0L, as.integer(harvested)),
    released    = if_else(is.na(released),    0L, as.integer(released)),
    total_catch = if_else(is.na(total_catch), 0L, as.integer(total_catch))
  )

suppressMessages({
  harlan_design_ch16 <- creel_design(
    calendar = harlan_schedule,
    date     = "date",
    strata   = "day_type"
  ) |>
    add_counts(harlan_counts_ch16, count_time_col = "count_time") |>
    add_interviews(
      interviews     = harlan_interviews_ch16,
      catch          = "total_catch",
      harvest        = "harvested",
      effort         = "hours_fished",
      trip_duration  = "hours_fished",
      trip_status    = "trip_status",
      n_anglers      = "n_anglers",
      angler_type    = "angler_type",
      angler_method  = "angler_method",
      interview_type = "access"
    ) |>
    add_catch(
      data          = harlan_catch,
      catch_uid     = "interview_id",
      interview_uid = "interview_id",
      species       = "species",
      count         = "n_fish",
      catch_type    = "catch_type"
    )
})

fmt_comma <- function(x) formatC(round(x), format = "f", digits = 0, big.mark = ",")
fmt_rate  <- function(x) sprintf("%.2f", x)
fmt_pct   <- function(x) sprintf("%.1f%%", x)
```

## Reporting as the final mile {#sec-reporting-intro}

A creel survey produces estimates with standard errors and confidence intervals.
Whether those numbers serve their purpose depends entirely on how they are
communicated. A well-executed survey can mislead if the report fails to state
what domain was sampled, how the estimate was constructed, or what uncertainty
attaches to the result. A less precise survey can still support a sound management
decision if the report is honest about its limitations and what the numbers cannot
answer.

Reporting is not cleanup that happens after analysis — it is the last step of
analysis. The decisions made during survey design (which strata to sample, which
estimator to apply, which interviews to include) determine what the final number
actually measures. The report is where those decisions become visible to people
who did not make them. A number without a domain statement is not an estimate;
it is a quantity of unknown scope.

This chapter shows how to assemble all Harlan Reservoir 2022 estimates into a
single report-ready structure, write defensible narrative sentences with inline R
so the prose stays synchronized with the analysis, present tables and figures
appropriate for different management audiences, and state caveats that protect
readers from overgeneralizing results to domains the survey did not cover.

## The reporting contract {#sec-report-contract}

Every creel report makes an implicit claim about what the numbers mean. Making
that claim explicit requires six elements, stated in some form before any
estimates appear.

**1. The question.** What biological or management quantity was the primary
target? Effort, total harvest, catch rate, or all three? Specifying the question
first tells the reader what to look for and what to skip. A report whose stated
question is "what was the total harvest of game fish at Harlan Reservoir in
2022?" has a different burden of evidence than one asking "how did effort vary
by day type?"

**2. The domain.** Which part of the fishery do the estimates cover? The Harlan
2022 survey sampled public-access areas using an access-point design during
standard operating hours from April through October. Estimates therefore apply
to that spatiotemporal domain — not to private shorelines, not to pre-spawn
April activity occurring before sampling began, not to effort on days when no
crew was deployed.

The most common reporting failure is leaving the domain unstated [@pollock1994].
Without it, an estimate of 19,000 fish harvested reads as though it applies to
the entire reservoir and the entire calendar year. It does not. The domain is
not a caveat to be buried at the end — it belongs in the same sentence as the
number.

**3. The design and unit of analysis.** Stratified access-point creel survey;
day-type strata (weekday and weekend); angler-hours as the effort unit;
party-based interviews; 600 total interviews across the season. A reader who
understands the design can evaluate whether the estimators are appropriate and
whether the same design could be applied to their own system.

**4. The key estimates.** Point estimates for each quantity, with labels that
include units. "Effort" is ambiguous; "33,000 angler-hours" is not.

**5. Uncertainty.** Standard errors and 95% confidence intervals for every
estimate in the report. A point estimate without uncertainty is not an estimate
— it is a guess with no information about how seriously to take it. Presenting
only the point estimate implicitly claims perfect precision, which no creel
survey achieves.

**6. Interpretation.** A brief statement of what the estimate implies for
management, constrained to what the data can support. The interpretation should
not be wider than the domain. If the survey covered public-access areas only,
the interpretation should not conclude that the total fishery is under pressure —
it should conclude that the sampled portion shows evidence of pressure.

These six elements protect both the reader and the analyst. Readers get the
information they need to evaluate how much weight to put on the numbers.
Analysts get a record of what the survey was designed to measure and a defense
against future questions about what was or was not estimated.

## Assembling the Harlan estimates {#sec-report-assemble}

Chapters 12 through 15 demonstrated each estimator separately. A report needs
all of them together, in one structure, so that the full picture of the fishery
is visible at once. The starting point is the same design object built throughout
Part IV — here reconstructed from the raw Harlan data in a hidden setup chunk
using the same pipeline from those earlier chapters.

Running all estimates in sequence and extracting the `$estimates` slot from each
produces a set of tibbles with a consistent column structure:
`estimate`, `se`, `ci_lower`, `ci_upper`. That structural consistency is what
makes assembly straightforward.

```{r}
#| label: estimates-ch16
#| echo: true

eff_ch16     <- estimate_effort(harlan_design_ch16,         target = "stratum_total")
harvest_ch16 <- estimate_total_harvest(harlan_design_ch16,  target = "stratum_total")
catch_ch16   <- estimate_total_catch(harlan_design_ch16,    target = "stratum_total")
release_ch16 <- estimate_total_release(harlan_design_ch16,  target = "stratum_total")
cpue_ch16    <- estimate_catch_rate(harlan_design_ch16)
hpue_ch16    <- estimate_harvest_rate(harlan_design_ch16)
rpue_ch16    <- estimate_release_rate(harlan_design_ch16)
```

The `target = "stratum_total"` argument expands sampled-day estimates to the
full stratum — the appropriate choice when reporting a season total rather than
a summary of surveyed days only. The rate estimators (`estimate_catch_rate`,
`estimate_harvest_rate`, `estimate_release_rate`) do not take a `target`
argument because they are ratios, not totals; they are already expressed per
unit effort and scale-free with respect to sampling intensity.

With all seven objects in hand, `bind_rows()` collects them into a single
reporting table. A `select()` call standardizes the column set across
estimators that return slightly different output structures (effort returns
`se_between` and `se_within`; other estimators return only `se`).

```{r}
#| label: summary-raw-ch16
#| echo: true

report_summary <- bind_rows(
  eff_ch16$estimates |>
    select(estimate, se, ci_lower, ci_upper) |>
    mutate(quantity = "Effort",         unit = "angler-hours"),
  harvest_ch16$estimates |>
    select(estimate, se, ci_lower, ci_upper) |>
    mutate(quantity = "Total harvest",  unit = "fish"),
  catch_ch16$estimates |>
    select(estimate, se, ci_lower, ci_upper) |>
    mutate(quantity = "Total catch",    unit = "fish"),
  release_ch16$estimates |>
    select(estimate, se, ci_lower, ci_upper) |>
    mutate(quantity = "Total release",  unit = "fish"),
  cpue_ch16$estimates |>
    select(estimate, se, ci_lower, ci_upper) |>
    mutate(quantity = "CPUE",           unit = "fish / hr"),
  hpue_ch16$estimates |>
    select(estimate, se, ci_lower, ci_upper) |>
    mutate(quantity = "HPUE",           unit = "fish / hr"),
  rpue_ch16$estimates |>
    select(estimate, se, ci_lower, ci_upper) |>
    mutate(quantity = "RPUE",           unit = "fish / hr")
) |>
  select(quantity, unit, estimate, se, ci_lower, ci_upper)

report_summary
```

The seven-row tibble is the raw material for every derived output in this
chapter — narrative sentences, formatted tables, and all three figures.
All downstream objects read from `report_summary` or from the individual estimate
objects, so any change to the estimation pipeline propagates automatically.

The estimates confirm the values established in earlier chapters:
`r fmt_comma(eff_ch16$estimates$estimate)` angler-hours of effort,
`r fmt_comma(harvest_ch16$estimates$estimate)` fish harvested,
`r fmt_comma(catch_ch16$estimates$estimate)` fish caught in total, and
a catch rate of `r fmt_rate(cpue_ch16$estimates$estimate)` fish per angler-hour.
The harvest CI spans from
`r fmt_comma(harvest_ch16$estimates$ci_lower)` to
`r fmt_comma(harvest_ch16$estimates$ci_upper)` fish —
a range of `r fmt_comma(harvest_ch16$estimates$ci_upper - harvest_ch16$estimates$ci_lower)` fish
from lower to upper bound. Reporting only the point estimate would conceal that
entire span.

## Narrative reporting with inline R {#sec-narrative}

A table of estimates is not a report. It is the evidence a report draws on.
The report also needs to tell the reader what the estimates mean, in plain
language, with the numbers embedded directly in the prose so that a reader who
never reaches the table still encounters the key result.

Quarto supports inline R expressions using backtick notation: any expression
inside `` `r ` `` is evaluated at render time and its result is inserted into
the surrounding prose. This mechanism ensures that the narrative cannot drift
from the analysis. If the data or design changes, re-rendering the document
updates every inline number automatically. There is no manual transcription
step where errors can enter.

```{r}
#| label: fmt-helpers-ch16
#| echo: true
#| eval: false

fmt_comma <- function(x) formatC(round(x), format = "f", digits = 0, big.mark = ",")
fmt_rate  <- function(x) sprintf("%.2f", x)
fmt_pct   <- function(x) sprintf("%.1f%%", x)
```

These helpers, defined once in the setup chunk, cover the three formats that
creel reports need most often. `fmt_comma()` rounds to the nearest integer and
adds thousand-separator commas — appropriate for totals where sub-unit precision
is spurious. `fmt_rate()` uses two decimal places — appropriate for CPUE and
HPUE where differences at the second decimal place can signal meaningful change
across years. `fmt_pct()` formats coefficients of variation as percentages for
precision tables.

The sentence pattern that works for creel estimates has three parts: what was
estimated, how large the estimate is, and how uncertain it is. Domain belongs
in the first part. The three examples below show correctly structured report
sentences for effort, harvest, and catch rate.

::: {.callout-note title="Report sentence: effort"}
Anglers expended an estimated
`r fmt_comma(eff_ch16$estimates$estimate)` angler-hours at Harlan
Reservoir during the 2022 open season
(95% CI: `r fmt_comma(eff_ch16$estimates$ci_lower)`–`r fmt_comma(eff_ch16$estimates$ci_upper)` angler-hours).
:::

::: {.callout-note title="Report sentence: harvest"}
Total harvest in public-access areas was estimated at
`r fmt_comma(harvest_ch16$estimates$estimate)` fish
(SE = `r fmt_comma(harvest_ch16$estimates$se)`,
95% CI: `r fmt_comma(harvest_ch16$estimates$ci_lower)`–`r fmt_comma(harvest_ch16$estimates$ci_upper)` fish).
:::

::: {.callout-note title="Report sentence: catch rate"}
The overall catch rate was
`r fmt_rate(cpue_ch16$estimates$estimate)` fish per angler-hour
(SE = `r fmt_rate(cpue_ch16$estimates$se)`,
95% CI: `r fmt_rate(cpue_ch16$estimates$ci_lower)`–`r fmt_rate(cpue_ch16$estimates$ci_upper)` fish / hr),
based on `r cpue_ch16$estimates$n` completed party interviews.
:::

Each sentence includes three things that are often omitted in practice: the
domain ("Harlan Reservoir, 2022 open season"), the unit ("angler-hours" or
"fish per angler-hour"), and the uncertainty ("SE =" and "95% CI"). Omitting
any one of them weakens the claim without necessarily making it wrong — but it
leaves the reader unable to evaluate the estimate.

Choosing precision is also a reporting decision. Reporting a harvest estimate
of `r fmt_comma(harvest_ch16$estimates$estimate)` fish implies that the model
is accurate to the individual fish, which it is not. The point estimate is an
expansion of count-derived effort and interview-derived rates, each carrying
substantial uncertainty. Rounding to the nearest hundred or even the nearest
thousand is defensible for a management report where the practical question is
"roughly how many fish were removed?" rather than "exactly how many?"

```{r}
#| label: rounding-demo-ch16
#| echo: true

harvest_est <- harvest_ch16$estimates$estimate

cat(sprintf(
  "Raw point estimate:       %s fish\n
Rounded to nearest 100:   %s fish\n
Rounded to nearest 1,000: %s fish\n",
  fmt_comma(harvest_est),
  fmt_comma(round(harvest_est, -2)),
  fmt_comma(round(harvest_est, -3))
))
```

For most management contexts, rounding to the nearest hundred preserves the
meaningful information while avoiding spurious precision. The exact rounding
choice should be stated in the report — "estimates are rounded to the nearest
100 fish" — so readers understand that the reported figure is not exact.

## The summary table {#sec-report-table}

A formatted summary table presents all estimates simultaneously with consistent
rounding and visible uncertainty. The goal is a table that a fisheries manager
can hand to a supervisor or include in an annual report without additional
editing.

```{r}
#| label: report-table-ch16
#| echo: true

report_table <- report_summary |>
  mutate(
    Estimate = case_when(
      unit %in% c("angler-hours", "fish") ~ fmt_comma(estimate),
      TRUE                                 ~ fmt_rate(estimate)
    ),
    SE = case_when(
      unit %in% c("angler-hours", "fish") ~ fmt_comma(se),
      TRUE                                 ~ fmt_rate(se)
    ),
    `95% CI` = paste0(
      case_when(
        unit %in% c("angler-hours", "fish") ~ fmt_comma(ci_lower),
        TRUE                                 ~ fmt_rate(ci_lower)
      ),
      "–",
      case_when(
        unit %in% c("angler-hours", "fish") ~ fmt_comma(ci_upper),
        TRUE                                 ~ fmt_rate(ci_upper)
      )
    ),
    CV = fmt_pct(se / estimate * 100)
  ) |>
  select(
    Quantity = quantity,
    Unit     = unit,
    Estimate,
    SE,
    `95% CI`,
    CV
  )

report_table
```

The CV column — coefficient of variation, expressed as SE / estimate × 100 —
gives readers a scale-free precision indicator. A CV below 20% is commonly
cited as management-grade precision for recreational fisheries surveys
[@pollock1994]. The Harlan estimates vary by quantity: effort benefits from
the stratified design and has the lowest CV among the totals, while harvest and
catch totals carry both effort uncertainty and rate uncertainty, producing larger
CVs. Rate estimates fall between those extremes.

The table should appear in any report with a caption that states the domain:
"Estimated effort, harvest, catch, release, and catch rates for Harlan
Reservoir 2022, covering public-access areas during the open season
(April–October). All totals use `target = "stratum_total"` expansions;
rates are ratio-of-means estimates from all eligible interviews."

## Summary figure: totals {#sec-fig-totals}

For audiences who read figures before tables, a summary plot that shows all
totals with confidence intervals communicates scale and uncertainty more
efficiently than a table. Because effort (angler-hours) and the fish counts
(harvest, catch, release) have different units and very different magnitudes,
they require separate panels with free scales.

```{r}
#| label: fig-ch16-totals
#| echo: true
#| fig-cap: "Season totals for Harlan Reservoir 2022. Left panel: total angling effort in angler-hours. Right panel: estimated total catch, harvest, and release in number of fish. Points are ratio-of-means estimates; horizontal bars show 95% confidence intervals. Catch exceeds harvest because the majority of fish were released; the CI on each total reflects combined uncertainty from effort estimation and rate estimation."
#| fig-alt: "Two-panel dot-and-CI plot. Left panel shows estimated effort with a wide CI. Right panel shows three quantities: total catch, release, and harvest in descending order; all have wide CIs relative to their point estimates."

totals_effort <- report_summary |>
  filter(quantity == "Effort") |>
  mutate(panel = "Effort (angler-hours)")

totals_fish <- report_summary |>
  filter(quantity %in% c("Total catch", "Total harvest", "Total release")) |>
  mutate(panel = "Fish (count)")

totals_plot <- bind_rows(totals_effort, totals_fish) |>
  mutate(
    quantity = fct_relevel(quantity, "Total catch", "Total release", "Total harvest"),
    panel    = fct_relevel(panel, "Effort (angler-hours)", "Fish (count)")
  )

ggplot(totals_plot, aes(x = estimate, y = quantity)) +
  geom_linerange(
    aes(xmin = ci_lower, xmax = ci_upper),
    linewidth = 0.8,
    color     = "#2171b5"
  ) +
  geom_point(
    size  = 3,
    color = "#2171b5"
  ) +
  scale_x_continuous(labels = scales::comma) +
  facet_wrap(~panel, scales = "free", ncol = 2) +
  labs(
    x     = "Estimate (with 95% CI)",
    y     = NULL,
    title = "Harlan Reservoir 2022: season totals"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

The width of the confidence intervals in both panels conveys a message that
numbers alone can obscure: the estimates are imprecise. The effort CI alone
spans a range large enough to shift harvest estimates by several thousand fish.
A manager who sees only the point estimates might treat them as facts; the
figure makes clear that they are model-derived quantities with meaningful
uncertainty attached.

## Precision profile {#sec-fig-precision}

Not all estimates from the same survey are equally reliable. Effort estimates
from a well-designed stratified survey typically have lower CVs than harvest
or catch totals, because the latter add rate uncertainty on top of effort
uncertainty. A precision profile — a horizontal bar chart of CVs — makes
these differences visible and helps managers understand which parts of the
report deserve the most caution.

```{r}
#| label: fig-ch16-precision
#| echo: true
#| fig-cap: "Coefficient of variation (CV) for each Harlan Reservoir 2022 estimate. Bars show SE / estimate × 100. The dashed vertical line at 20% marks a commonly cited threshold for management-grade precision in recreational fishery surveys. Estimates with CVs above 20% should be interpreted cautiously and may justify increased sampling effort in future seasons."
#| fig-alt: "Horizontal bar chart showing CV for seven estimates ordered from largest to smallest. Color distinguishes effort, fish totals, and rate estimates. A vertical dashed line at 20% marks the management-grade precision threshold."

cv_tbl <- report_summary |>
  mutate(cv = se / estimate * 100) |>
  mutate(
    quantity = fct_reorder(quantity, cv),
    category = case_when(
      quantity == "Effort"                              ~ "Effort",
      as.character(quantity) %in% c("Total harvest",
                                    "Total catch",
                                    "Total release")   ~ "Total (fish)",
      TRUE                                              ~ "Rate (fish / hr)"
    )
  )

ggplot(cv_tbl, aes(x = cv, y = quantity, fill = category)) +
  geom_col(width = 0.6) +
  geom_vline(xintercept = 20, linetype = "dashed", color = "gray40") +
  annotate(
    "text",
    x = 21, y = Inf,
    label    = "20% target",
    hjust    = 0,
    vjust    = 1.5,
    size     = 3.2,
    color    = "gray30"
  ) +
  scale_fill_manual(
    values = c(
      "Effort"           = "#2171b5",
      "Total (fish)"     = "#6baed6",
      "Rate (fish / hr)" = "#bdd7e7"
    )
  ) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    x    = "Coefficient of variation (CV)",
    y    = NULL,
    fill = NULL,
    title = "Harlan Reservoir 2022: estimate precision"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position    = "bottom",
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_blank()
  )
```

```{r}
#| label: cv-summary-ch16
#| echo: false

eff_cv     <- eff_ch16$estimates$se     / eff_ch16$estimates$estimate     * 100
harvest_cv <- harvest_ch16$estimates$se / harvest_ch16$estimates$estimate * 100
cpue_cv    <- cpue_ch16$estimates$se    / cpue_ch16$estimates$estimate    * 100
```

For the Harlan 2022 survey, effort precision
(CV = `r fmt_pct(eff_cv)`) falls within the range expected for a stratified
access-point survey with adequate sampling of both day types. Harvest carries
a substantially higher CV
(CV = `r fmt_pct(harvest_cv)`) because it inherits uncertainty from both
the effort expansion and the harvest rate estimation; each source of uncertainty
combines multiplicatively in the total. The catch rate, derived only from
completed trip interviews, has a CV of `r fmt_pct(cpue_cv)` — moderate
precision that reflects the number of completed interviews relative to the
variability in individual catch outcomes.

Estimates that exceed the 20% CV threshold are not wrong — they are imprecise.
An imprecise estimate is still useful if the management question is whether a
quantity is large or small, not whether it equals a specific value. An estimated
harvest with a high CV still rules out both very low and very high harvests; it
just cannot distinguish between values that differ by less than about one
standard error on either side of the point estimate.

If specific estimates consistently fall above the CV threshold across years,
the appropriate response is to revisit the sampling design rather than to
report the numbers without comment. Chapter 15 showed how `creel_power()` and
`cv_from_n()` can guide sample size decisions toward a precision target. The
precision profile here identifies which quantities need that investment.

## Stratified reporting {#sec-report-strat}

The season totals are the headline, but stratified estimates often contain
the management-relevant signal. A regulation that differentially affects
weekday or weekend fishing — a bag limit change, a closure, a catch-and-release
requirement — should be evaluated against stratified estimates, not the
pooled season total.

The same `estimate_total_harvest()` call with a `by` argument produces
stratum-level estimates. These share the design structure from the overall call,
so the stratified and unstratified results are methodologically consistent.

```{r}
#| label: harvest-by-daytype-ch16
#| echo: true

harvest_by_dt <- estimate_total_harvest(
  harlan_design_ch16,
  by     = day_type,
  target = "stratum_total"
)

harvest_by_dt$estimates
```

The stratified harvest figures answer a question the season total cannot: where
does the removal occur? A figure makes the comparison immediate.

```{r}
#| label: fig-ch16-strat
#| echo: true
#| fig-cap: "Estimated total harvest by day type at Harlan Reservoir, 2022. Points show stratum-total estimates; bars show 95% confidence intervals. Weekday harvest substantially exceeds weekend harvest in absolute terms because the weekday stratum contains more days and accumulated more angler-hours over the season."
#| fig-alt: "Horizontal dot-and-CI plot with two rows: weekday and weekend. The weekday estimate is substantially larger than the weekend estimate, and both CIs are wide but non-overlapping."

ggplot(harvest_by_dt$estimates, aes(x = estimate, y = day_type)) +
  geom_linerange(
    aes(xmin = ci_lower, xmax = ci_upper),
    linewidth = 0.8,
    color     = "#2171b5"
  ) +
  geom_point(size = 3.5, color = "#2171b5") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    x     = "Estimated harvest (fish, with 95% CI)",
    y     = NULL,
    title = "Harlan Reservoir 2022: harvest by day type"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

```{r}
#| label: strat-table-ch16
#| echo: true

strat_table <- harvest_by_dt$estimates |>
  mutate(
    `Day type` = day_type,
    Harvest    = fmt_comma(estimate),
    SE         = fmt_comma(se),
    `95% CI`   = paste0(fmt_comma(ci_lower), "–", fmt_comma(ci_upper)),
    CV         = fmt_pct(se / estimate * 100)
  ) |>
  select(`Day type`, Harvest, SE, `95% CI`, CV)

strat_table
```

The stratified harvest table answers a different question than the season total:
not how many fish were harvested overall, but whether harvest is concentrated
in weekday or weekend fishing. That distinction matters when a regulation change
would close or restrict one day type, because the season-total reduction is not
proportional to harvest per day — it depends on how many days of each type
occur and how much harvest each carries.

Interpreting stratified estimates correctly requires stating the stratum
definition in the table caption. "Weekday" in the Harlan survey means any
Monday through Friday that fell within the sampling frame; "weekend" means
Saturday and Sunday. A regulation effective only on Saturdays would affect a
subset of the "weekend" stratum, and the stratum estimate would overstate the
impact unless further post-stratified.

## Domain and scope {#sec-report-domain}

Every estimate has a scope boundary. The boundary is not a limitation to
apologize for — it is a property of the design that tells readers where the
estimate applies. Stating it clearly prevents two common errors: applying the
estimate to a domain it cannot cover, and assuming the estimate is invalid
because it does not cover everything.

For the Harlan 2022 survey, the scope boundaries are:

**Temporal scope.** Sampling ran from April through October during standard
crew operating hours. Effort and harvest on days outside the sampling frame,
during pre-dawn and late-evening hours, and on crew non-deployment days are
not in the estimate. Winter fishing is excluded entirely.

**Spatial scope.** The access-point design covers public-access areas where
departing anglers can be intercepted. Private shorelines, mid-lake boats not
associated with a sampled access point, and tributaries outside the main
reservoir are not represented.

**Species scope.** Catch and harvest rates are available for species recorded
in the catch table. Rare species may be under-sampled — a catch record requires
that at least one party in the interview sample encountered that species. Total
harvest by species inherits both effort uncertainty and the sampling uncertainty
from the catch table.

**Design scope.** The access-point method samples anglers as they depart. It
is not designed to capture effort by anglers who never return to a sampled
access point (e.g., anglers who enter by canoe and exit via a private road).
The fraction of effort in those groups is unknown; if it is large, the effort
estimate is biased downward.

Stating these boundaries in the report is not weakness — it is the information
a reader needs to decide whether the estimate answers their management question.
If the question is "how much did harvest decline after a bag limit change?",
the answer requires an estimate with the same domain in both years. If the
domain is not documented, the before-after comparison cannot be made with
confidence even if both years' surveys used identical methods.

## Writing for different audiences {#sec-report-audience}

A creel report often serves audiences with different needs and different
tolerances for statistical language. Field staff need operational feedback.
Managers need a decision summary. Researchers need methodological detail.
Stakeholders need a plain-language explanation of what the survey found.

The same estimates can support all four audiences without changing the underlying
claim — only the presentation changes.

**Manager summary (2–3 sentences).** Lead with the most policy-relevant
quantity. State the estimate, the CI, and one interpretive sentence.

::: {.callout-note title="Manager version"}
The 2022 Harlan Reservoir creel survey estimated
`r fmt_comma(harvest_ch16$estimates$estimate)` fish harvested
(95% CI: `r fmt_comma(harvest_ch16$estimates$ci_lower)`–`r fmt_comma(harvest_ch16$estimates$ci_upper)`)
from public-access areas during the April–October season.
The catch rate of `r fmt_rate(cpue_ch16$estimates$estimate)` fish per
angler-hour was within the range observed in recent years.
Total effort was approximately
`r fmt_comma(eff_ch16$estimates$estimate)` angler-hours.
:::

**Researcher disclosure.** Provide the estimator name, the variance formula,
the software pipeline, and any assumptions that were checked or violated.
State CV for each quantity and note any diagnostic findings (incomplete-trip
sensitivity, variance decomposition, coverage diagnostics). The `report_table`
from @sec-report-table, with its CV column, is the core of a researcher-facing
results section.

**Stakeholder language.** Avoid jargon, define every number in plain terms
before stating it, and lead with what the fishery looked like rather than what
the survey was designed to do.

::: {.callout-note title="Stakeholder version"}
About `r fmt_comma(harvest_ch16$estimates$estimate)` fish were kept by
anglers fishing Harlan Reservoir in 2022. That number could be anywhere
from `r fmt_comma(harvest_ch16$estimates$ci_lower)` to
`r fmt_comma(harvest_ch16$estimates$ci_upper)` fish, depending on
how representative the surveyed days were of the full season.
Of all fish caught, roughly
`r fmt_pct((catch_ch16$estimates$estimate - harvest_ch16$estimates$estimate) / catch_ch16$estimates$estimate * 100)`
were released.
:::

The estimate does not change between these versions. The confidence interval
does not change. What changes is which information gets foregrounded and what
vocabulary is used to present it.

A failure of audience adaptation is describing the survey method in detail to
a stakeholder audience that wants to know whether the fish population is healthy
— and omitting the uncertainty when writing for a manager who needs to decide
whether to tighten regulations. Both failures serve the wrong need.

## Reproducible reports {#sec-reproducibility}

The most defensible property of a Quarto-based creel report is that it can be
regenerated from the same data and code. When the data change — because a data
entry error is corrected, or a late interview is added — re-running
`quarto render` updates every table, figure, and inline number automatically.
No manual transcription step introduces new errors.

That reproducibility has practical consequences for multi-year programs:

**Year-over-year reuse.** A report template developed for the 2022 season can
be applied to 2023 data by swapping the input files. The chapter structure,
section headings, figure layouts, and table formats carry over. Only the
numbers change, and those changes are automatic.

**Audit trail.** Because the analysis lives in the same document as the report,
a future analyst can trace any number back to the code that produced it. There
is no gap between "the analysis" and "the report" where undocumented decisions
accumulate.

**Version control.** Storing the `.qmd` source and input data in a version
control system (git) creates a complete record of every change to the report
over time. If a manager asks why the 2022 harvest estimate was revised, the
git history shows exactly what changed and when.

The design object — `harlan_design_ch16` in this chapter — is the key
intermediate artifact. Saving it at the end of the analysis pipeline and
loading it at the start of the reporting pipeline creates a clean separation
between estimation and presentation without breaking the audit chain.

```{r}
#| label: save-design-ch16
#| echo: true
#| eval: false

saveRDS(harlan_design_ch16, here("data/harlan_design_2022.rds"))
```

A report template that loads a saved design object rather than rebuilding it
from raw data is faster to render and explicit about where the design
assumptions live. The data provenance statement in the report caption can then
cite the specific data files and the date they were last modified.

## Common reporting failures {#sec-report-failures}

The six-element framework from @sec-report-contract works as a checklist for
catching the most common reporting errors before the report reaches its audience.

**Leaving out the domain.** An estimate of 19,000 fish harvested with no domain
statement could refer to the entire reservoir, a single access point, or a single
species. The reader cannot tell. Fix: state the domain — water body, access type,
sampling period, species scope — in the first sentence that reports any estimate.

**Failing to state the denominator.** A catch rate is meaningless without its
denominator. "Anglers caught 1.5 fish" is not a catch rate. "Anglers caught
1.5 fish per angler-hour" is. The denominator is what makes the rate comparable
across surveys of different sizes. Fix: always report rates with explicit units.

**Presenting a point estimate without uncertainty.** A single number cannot tell
the reader whether the survey was precise enough to be useful or imprecise enough
to be misleading. Fix: report SE and 95% CI alongside every point estimate. The
CV column in @sec-report-table makes it easy to include precision for all
quantities at once.

**Mixing stratified and unstratified interpretations.** If the estimate covers
all days, the interpretation should not compare weekday versus weekend behavior
unless a stratified estimate is also reported. Interpreting the season-total CPUE
as evidence that weekday anglers catch more fish is a domain error — the total
was not computed by day type. Fix: every interpretation stays within the domain
of the estimate that supports it.

**Burying the domain in a footnote.** A domain statement that appears only in a
footnote or appendix will be missed by readers who stop at the table. Fix: the
domain belongs in the first paragraph and in every table caption.

**Describing a result as general when it covers only a subset.** "Harlan
Reservoir anglers harvested approximately 20,000 fish" implies coverage of all
anglers. A survey that covers only public-access areas cannot make that claim
without qualification. Fix: "Anglers at public-access areas covered by the 2022
creel survey harvested approximately 20,000 fish."

## Takeaway {#sec-ch16-takeaway}

The estimate is correct only if the report says what it means. A point estimate
without a domain, a domain without uncertainty, or an interpretation wider than
the evidence all undermine the value of the survey that produced the number.

The inline R approach — embedding estimates in prose at render time — protects
against transcription errors and keeps the document honest when data change.
The six-element framework protects against the interpretive errors that no
amount of statistical care can prevent if the report does not state what was
measured.

The Harlan Reservoir 2022 estimates tell a specific story about a specific
fishery during a specific season at specific access points. That story is worth
telling precisely. It is also worth telling accurately — which means telling the
reader where the story begins and where it ends.
