---
title: "Case Study: Cedar Lake Roving Creel Survey"
---

```{r}
#| label: setup
#| include: false
library(tidycreel)
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
library(forcats)
library(patchwork)
library(here)

cedar_schedule   <- readRDS(here("data/cedar_schedule.rds"))
cedar_counts     <- readRDS(here("data/cedar_counts.rds"))
cedar_interviews <- readRDS(here("data/cedar_interviews.rds"))
cedar_catch      <- readRDS(here("data/cedar_catch.rds"))

# Shift hours: both periods 6 hours (6 am–noon and noon–6 pm)
shift_hrs <- c("1" = 6L, "2" = 6L)

# Counts: pre-compute angler_hours, retain only count variable (Ch12 pattern)
cedar_counts_effort <- cedar_counts |>
  mutate(
    period       = as.character(period),
    angler_hours = (bank_anglers + boat_anglers) * shift_hrs[period]
  ) |>
  select(reservoir, date, day_type, period, section, count_time, angler_hours)

# Catch wide: harvested + released columns required by add_interviews()
catch_wide <- cedar_catch |>
  group_by(interview_id, catch_type) |>
  summarise(n = sum(n_fish), .groups = "drop") |>
  pivot_wider(names_from = catch_type, values_from = n, values_fill = 0L) |>
  mutate(total_catch = harvested + released)

cedar_int2 <- cedar_interviews |>
  mutate(period = as.character(period)) |>
  left_join(catch_wide, by = "interview_id") |>
  mutate(across(c(harvested, released, total_catch),
                ~ if_else(is.na(.), 0L, as.integer(.))))

# Build design object
suppressMessages({
  cedar_design <- creel_design(
    calendar = cedar_schedule,
    date     = "date",
    strata   = "day_type"
  ) |>
    add_counts(cedar_counts_effort, count_time_col = "count_time") |>
    add_interviews(
      interviews     = cedar_int2,
      catch          = "total_catch",
      harvest        = "harvested",
      effort         = "hours_fished",
      trip_duration  = "hours_fished",
      trip_status    = "trip_status",
      n_anglers      = "n_anglers",
      angler_type    = "angler_type",
      angler_method  = "angler_method",
      interview_type = "roving"
    ) |>
    add_catch(
      data          = cedar_catch,
      catch_uid     = "interview_id",
      interview_uid = "interview_id",
      species       = "species",
      count         = "n_fish",
      catch_type    = "catch_type"
    )
})

# Pre-compute all season-level estimates for inline R
eff_tot   <- estimate_effort(cedar_design, target = "stratum_total")
eff_strat <- estimate_effort(cedar_design, by = day_type, target = "stratum_total")
harv_tot  <- estimate_total_harvest(cedar_design, target = "stratum_total")
harv_strat <- estimate_total_harvest(cedar_design, by = day_type, target = "stratum_total")
cat_tot   <- estimate_total_catch(cedar_design, target = "stratum_total")
rel_tot   <- estimate_total_release(cedar_design, target = "stratum_total")
cpue_all  <- estimate_catch_rate(cedar_design)
cpue_dt   <- estimate_catch_rate(cedar_design, by = day_type)
cpue_at   <- estimate_catch_rate(cedar_design, by = angler_type)
hpue_all  <- estimate_harvest_rate(cedar_design)

# Formatting helpers
fmt_comma <- function(x) formatC(round(x), format = "f", digits = 0, big.mark = ",")
fmt_rate  <- function(x) sprintf("%.2f", x)
fmt_pct   <- function(x) sprintf("%.1f%%", x)
fmt_cv    <- function(est) fmt_pct(100 * est$se / est$estimate)
```

## Setting and design overview

Cedar Lake is the roving counterpart to the Harlan access-point example in Part IV. A single clerk drove the shoreline twice per sampling period, counted visible anglers, and interviewed parties encountered mid-trip. That design produces many incomplete-trip contacts and requires a rate estimator that works from truncated trip records.

The fishery targets channel catfish, bluegill, and white crappie across the April-through-October 2023 open-water season. Pressure peaks in June and July; April and October are thinner because of weather and occasional cancelled survey days.

This chapter walks the full analysis pipeline — frame to final report — using Cedar Lake as the vehicle.

## Data overview

Four linked tables describe the 2023 Cedar Lake survey:

```{r}
#| label: cedar-summary
n_days   <- nrow(cedar_schedule)
n_samp   <- sum(cedar_schedule$sampled)
n_int    <- nrow(cedar_interviews)
n_comp   <- sum(cedar_interviews$trip_status == "complete")
n_incomp <- sum(cedar_interviews$trip_status == "incomplete")
pct_bank <- 100 * mean(cedar_interviews$angler_type == "bank")

cat(
  "Survey period:   ", format(min(cedar_schedule$date), "%B %d"), "–",
  format(max(cedar_schedule$date), "%B %d, %Y"), "\n",
  "Total days:      ", n_days, "\n",
  "Sampled days:    ", n_samp, sprintf("(%.0f%%)\n", 100 * n_samp / n_days),
  "Interviews:      ", n_int,
  sprintf("(%d complete, %d incomplete)\n", n_comp, n_incomp),
  "Bank anglers:    ", sprintf("%.0f%% of interviews\n", pct_bank),
  sep = ""
)
```

The `r fmt_pct(100 * n_incomp / n_int)` incomplete-trip rate is expected in a roving design: interviews happen mid-trip, so only elapsed time and fish-so-far are observed. Complete trips account for `r fmt_pct(100 * n_comp / n_int)` of contacts. Total catch uses only completed trips, while CPUE uses every interview as an instantaneous rate observation.

### Seasonal survey coverage

Sampling was weighted toward summer months when angler pressure is highest. Eighteen survey days were cancelled or produced zero-angler counts because of weather.

```{r}
#| label: fig-cedar-season
#| fig-cap: "Monthly sampling effort and mean instantaneous angler count at Cedar Lake, 2023. Left: number of sampled days per month by stratum. Right: mean total anglers per count by month. Zero-count days from adverse weather appear in both panels."
#| fig-width: 8
#| fig-height: 3.5

monthly_days <- cedar_schedule |>
  filter(sampled) |>
  mutate(month = factor(format(date, "%b"), levels = month.abb)) |>
  count(month, day_type)

monthly_counts <- cedar_counts |>
  mutate(
    month = factor(format(date, "%b"), levels = month.abb),
    total = bank_anglers + boat_anglers
  ) |>
  group_by(month) |>
  summarise(mean_count = mean(total), .groups = "drop")

p1 <- ggplot(monthly_days, aes(month, n, fill = day_type)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c(weekday = "#6baed6", weekend = "#2171b5"),
                    name = "Day type") +
  labs(x = NULL, y = "Sampled days") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p2 <- ggplot(monthly_counts, aes(month, mean_count)) +
  geom_col(fill = "#74c476") +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(x = NULL, y = "Mean anglers per count") +
  theme_minimal(base_size = 11)

p1 + p2
```

Mean instantaneous counts rise from `r round(monthly_counts$mean_count[monthly_counts$month == "Apr"], 1)` anglers per count in April to `r round(monthly_counts$mean_count[monthly_counts$month == "Aug"], 1)` in August, then fall through October. The low-count days at the season edges line up with weather-related cancellations.

### Angler composition across the season

Bank anglers dominate the count record, with only modest month-to-month movement in the small boat contingent.

```{r}
#| label: fig-cedar-bankboat
#| fig-cap: "Monthly angler composition at Cedar Lake, 2023, from instantaneous count records. Bank anglers (dark) dominate throughout the season; the boat fraction increases slightly in midsummer. Zero-count days from adverse weather reduce monthly totals without changing the proportion."
#| fig-width: 6.5
#| fig-height: 3.2

cedar_counts |>
  mutate(month = factor(format(date, "%b"), levels = month.abb)) |>
  group_by(month) |>
  summarise(
    bank = sum(bank_anglers),
    boat = sum(boat_anglers),
    .groups = "drop"
  ) |>
  pivot_longer(c(bank, boat), names_to = "type", values_to = "count") |>
  ggplot(aes(month, count, fill = type)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = c(bank = "#74c476", boat = "#2171b5"),
                    name = "Angler type") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Fraction of counted anglers",
       title = "Monthly bank vs. boat composition") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

Bank anglers account for `r fmt_pct(100 * sum(cedar_counts$bank_anglers) / sum(cedar_counts$bank_anglers + cedar_counts$boat_anglers))` of all counted anglers across the season, so angler-type CPUE is not driven by a large seasonal shift in who is on the water.

### Incomplete trips in a roving design

```{r}
#| label: cedar-trip-summary
summarize_trips(cedar_design)
```

`summarize_trips()` reports median and mean trip duration separately for complete and incomplete trips. At Cedar Lake, complete trips last longer than incomplete contacts, which is the length-bias you expect when the clerk meets anglers mid-circuit.

```{r}
#| label: cedar-incomplete-check
validate_incomplete_trips(
  cedar_design,
  catch   = harvested,
  effort  = hours_fished,
  truncate_at = 0.5
)
```

`validate_incomplete_trips()` runs a two-one-sided t-test (TOST) comparing harvest per hour between incomplete and complete trips. Non-significance means the data are consistent with rate equivalence within the default ± 0.5 fish/hr tolerance.

## Building the design object

The Cedar Lake pipeline follows the same four-step assembly used throughout Part III and Part IV. Three details differ from the Harlan build:

1. **Both periods have equal shift length.** Harlan's morning shift runs 8 hours; Cedar Lake uses 6 hours for each period (6 am–noon, noon–6 pm). The shift-hours vector reflects that equality.
2. **Counts carry only `angler_hours`.** Bank and boat counts are summed and multiplied by shift length before calling `add_counts()`. The component columns are dropped so `tidycreel` selects only the target variable.
3. **Two catch columns required.** `estimate_total_harvest()` requires a dedicated harvest column passed via the `harvest` argument. Total catch (harvested + released) is a separate column passed via `catch`.

```{r}
#| label: cedar-counts-prep
# Pre-compute angler_hours; drop component columns
cedar_counts_effort |>
  filter(date == as.Date("2023-07-08")) |>
  select(date, day_type, period, count_time, angler_hours)
```

```{r}
#| label: cedar-int-prep
# Interview table with both catch columns
cedar_int2 |>
  select(interview_id, date, trip_status, angler_type, hours_fished,
         harvested, released, total_catch) |>
  slice(1:6)
```

The split between `harvested` and `total_catch` is what allows the estimation layer to produce HPUE (harvest per unit effort) and CPUE (catch per unit effort) as distinct quantities from the same interview record.

```{r}
#| label: cedar-design-build
#| eval: false
cedar_design <- creel_design(
  calendar = cedar_schedule,
  date     = "date",
  strata   = "day_type"
) |>
  add_counts(cedar_counts_effort, count_time_col = "count_time") |>
  add_interviews(
    interviews     = cedar_int2,
    catch          = "total_catch",
    harvest        = "harvested",
    effort         = "hours_fished",
    trip_duration  = "hours_fished",
    trip_status    = "trip_status",
    n_anglers      = "n_anglers",
    angler_type    = "angler_type",
    angler_method  = "angler_method",
    interview_type = "roving"
  ) |>
  add_catch(
    data          = cedar_catch,
    catch_uid     = "interview_id",
    interview_uid = "interview_id",
    species       = "species",
    count         = "n_fish",
    catch_type    = "catch_type"
  )
```

The design object keeps the frame, counts, interviews, and catch together so all estimation functions read the same stratum labels and day-type assignments.

## Effort estimation

Total angler-hours comes from instantaneous count data. `add_counts()` averages the two replicates within each period, converts counts to day-period effort, and expands sampled-day effort to the full survey period within each stratum.

```{r}
#| label: cedar-effort
eff_tot$estimates
```

Cedar Lake logged an estimated `r fmt_comma(eff_tot$estimates$estimate)` angler-hours during the 2023 open-water season (95% CI: `r fmt_comma(eff_tot$estimates$ci_lower)`–`r fmt_comma(eff_tot$estimates$ci_upper)`). Weekend and weekday effort are nearly balanced.

```{r}
#| label: cedar-effort-strat
eff_strat$estimates
```

```{r}
#| label: fig-cedar-effort
#| fig-cap: "Season-total effort estimates by day-type stratum, Cedar Lake 2023. Points are ratio-of-means estimates; bars are 95% confidence intervals."
#| fig-width: 5.5
#| fig-height: 2.8

eff_strat$estimates |>
  ggplot(aes(estimate, fct_rev(day_type))) +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                 height = 0.15, linewidth = 0.8) +
  geom_point(size = 3, shape = 21, fill = "#6baed6") +
  scale_x_continuous(labels = scales::comma) +
  labs(x = "Estimated effort (angler-hours)", y = NULL,
       title = "Season effort by day type") +
  theme_minimal(base_size = 11)
```

Both strata have overlapping confidence intervals, so this is a reporting display rather than a hypothesis test.

The overall effort CV is `r fmt_pct(100 * eff_tot$estimates$se / eff_tot$estimates$estimate)`, within the 10–20% management-grade range for a stratified open-water creel survey. Between-day variance dominates.

## Catch, harvest, and release

Cedar Lake targets three species with different management priorities: channel catfish, bluegill, and white crappie.

### Species catch composition

```{r}
#| label: fig-cedar-species
#| fig-cap: "Total fish recorded in Cedar Lake interviews by species and catch type. Bars show absolute counts from the interview catch table before expansion; proportions within species reflect true harvest and release fractions."
#| fig-width: 6
#| fig-height: 3.2

cedar_catch |>
  group_by(species, catch_type) |>
  summarise(total = sum(n_fish), .groups = "drop") |>
  mutate(species = fct_reorder(species, total, .fun = sum)) |>
  ggplot(aes(total, species, fill = catch_type)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c(harvested = "#fd8d3c", released = "#74c476"),
                    name = "Catch type") +
  scale_x_continuous(labels = scales::comma) +
  labs(x = "Fish recorded in interviews", y = NULL,
       title = "Interview catch composition by species") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

Bluegill dominates raw catch counts, and channel catfish show the highest harvest fraction. That matches the fishery: panfish are abundant, while catfish are kept more often than bluegill or crappie.

### Season-level estimates

```{r}
#| label: cedar-harvest
harv_tot$estimates
cat_tot$estimates
rel_tot$estimates
```

The season totals reveal asymmetry in estimation precision. Total harvest uses the full interview set (`n` = `r harv_tot$estimates$n`), total catch uses only the `r cat_tot$estimates$n` completed trips, and total release uses the full interview set (`n` = `r rel_tot$estimates$n`). Harvest plus release will not exactly equal total catch because the inputs differ.

Harvest by stratum shows modest weekday–weekend differences, both within the season total's confidence interval:

```{r}
#| label: cedar-harv-strat
harv_strat$estimates
```

Season total harvest: `r fmt_comma(harv_tot$estimates$estimate)` fish (95% CI: `r fmt_comma(harv_tot$estimates$ci_lower)`–`r fmt_comma(harv_tot$estimates$ci_upper)`). Season total catch: `r fmt_comma(cat_tot$estimates$estimate)` fish.

## CPUE and harvest rate

Roving designs use the mean-of-ratios (MOR) estimator for catch rate, not the ratio-of-means (ROM) used in access-point surveys. Each mid-trip contact is a valid instantaneous rate observation. With `r fmt_pct(100 * n_incomp / n_int)` incomplete contacts, the MOR estimator still retains `r cpue_all$estimates$n` of `r n_int` interviews; eight very short contacts under 0.5 hours are excluded as unreliable.

```{r}
#| label: cedar-cpue-all
cpue_all$estimates
hpue_all$estimates
```

Overall CPUE is `r fmt_rate(cpue_all$estimates$estimate)` fish per angler-hour (95% CI: `r fmt_rate(cpue_all$estimates$ci_lower)`–`r fmt_rate(cpue_all$estimates$ci_upper)`). The harvest rate is `r fmt_rate(hpue_all$estimates$estimate)` fish per angler-hour, so anglers kept `r fmt_pct(100 * hpue_all$estimates$estimate / cpue_all$estimates$estimate)` of all fish caught.

### CPUE by day type

```{r}
#| label: cedar-cpue-daytype
cpue_dt$estimates
```

Weekday and weekend CPUE are nearly identical — `r fmt_rate(cpue_dt$estimates$estimate[cpue_dt$estimates$day_type == "weekday"])` and `r fmt_rate(cpue_dt$estimates$estimate[cpue_dt$estimates$day_type == "weekend"])` fish per angler-hour. Overlapping confidence intervals show no detectable difference at this sample size.

### CPUE by angler type

Bank and boat anglers differ in access, target species, and technique. Bank anglers typically fish bait near cover for bluegill and catfish, while the small boat fleet leans more toward crappie. The CPUE split reflects those differences.

```{r}
#| label: fig-cedar-cpue-type
#| fig-cap: "Catch per angler-hour by angler type at Cedar Lake, 2023. Points are mean-of-ratios estimates using all interviewed trips; bars are 95% confidence intervals. Boat stratum sample size is small (n < 50 contacts); variance estimate is less stable."
#| fig-width: 5.5
#| fig-height: 2.8

cpue_at$estimates |>
  ggplot(aes(estimate, fct_rev(angler_type))) +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                 height = 0.15, linewidth = 0.8) +
  geom_point(aes(shape = angler_type), size = 3, fill = "#6baed6") +
  scale_shape_manual(values = c(bank = 21, boat = 24), guide = "none") +
  geom_vline(xintercept = cpue_all$estimates$estimate, linetype = "dashed",
             linewidth = 0.5, color = "grey50") +
  annotate("text", x = cpue_all$estimates$estimate + 0.03,
           y = Inf, vjust = 1.5, hjust = 0, size = 3,
           label = sprintf("Overall: %.2f", cpue_all$estimates$estimate)) +
  labs(x = "Catch per angler-hour", y = NULL,
       title = "CPUE by angler type") +
  theme_minimal(base_size = 11)
```

Bank anglers catch `r fmt_rate(cpue_at$estimates$estimate[cpue_at$estimates$angler_type == "bank"])` fish per angler-hour; boat anglers catch `r fmt_rate(cpue_at$estimates$estimate[cpue_at$estimates$angler_type == "boat"])` fish per angler-hour. The boat sample is small (`r cpue_at$estimates$n[cpue_at$estimates$angler_type == "boat"]` trips), so treat that contrast cautiously.

## Precision and diagnostics

A precision profile summarizes CV across all season-level estimates and highlights which quantities are solid enough for management use.

```{r}
#| label: cedar-cv-summary
cv_eff   <- 100 * eff_tot$estimates$se   / eff_tot$estimates$estimate
cv_harv  <- 100 * harv_tot$estimates$se  / harv_tot$estimates$estimate
cv_catch <- 100 * cat_tot$estimates$se   / cat_tot$estimates$estimate
cv_cpue  <- 100 * cpue_all$estimates$se  / cpue_all$estimates$estimate
cv_hpue  <- 100 * hpue_all$estimates$se  / hpue_all$estimates$estimate

cv_tbl <- tibble::tibble(
  quantity = c("Effort", "Total harvest", "Total catch", "CPUE", "HPUE"),
  category = c("effort", "total", "total", "rate", "rate"),
  cv = c(cv_eff, cv_harv, cv_catch, cv_cpue, cv_hpue)
)
cv_tbl
```

```{r}
#| label: fig-cedar-precision
#| fig-cap: "Coefficient of variation (%) for season-level estimates, Cedar Lake 2023. Dashed line at 20% is a common management-grade threshold for acceptable precision."
#| fig-width: 6
#| fig-height: 3.2

cv_tbl |>
  mutate(quantity = fct_reorder(quantity, cv)) |>
  ggplot(aes(cv, quantity, fill = category)) +
  geom_col() +
  geom_vline(xintercept = 20, linetype = "dashed", linewidth = 0.6, color = "grey30") +
  scale_fill_manual(
    values = c(effort = "#6baed6", total = "#fd8d3c", rate = "#74c476"),
    name = "Estimate type"
  ) +
  labs(x = "CV (%)", y = NULL, title = "Precision profile") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

Effort CV is `r fmt_pct(cv_eff)`, the lowest of the group because it uses all `r n_samp` sampled days. Total harvest CV is `r fmt_pct(cv_harv)`. Total catch CV is `r fmt_pct(cv_catch)`, highest because only the `r n_comp` completed trips contribute. CPUE and HPUE are the smallest — `r fmt_pct(cv_cpue)` and `r fmt_pct(cv_hpue)` — because the MOR sample is large.

Quantities above 20% CV deserve a caveated interpretation. For Cedar Lake 2023, all season-level totals and rates fall within the management-grade range, but the boat-angler CPUE sub-group should still be reported with a strong uncertainty note.

## Final report assembly

All seven estimates are assembled into a single reporting tibble, formatted for a non-technical audience. The pattern mirrors Chapter 16: bind rows, add formatted strings, compute CV.

```{r}
#| label: cedar-report-assembly
report_summary <- bind_rows(
  eff_tot$estimates   |> mutate(quantity = "Effort",         unit = "angler-hours"),
  harv_tot$estimates  |> mutate(quantity = "Total harvest",  unit = "fish"),
  cat_tot$estimates   |> mutate(quantity = "Total catch",    unit = "fish"),
  rel_tot$estimates   |> mutate(quantity = "Total release",  unit = "fish"),
  cpue_all$estimates  |> mutate(quantity = "CPUE",           unit = "fish/hr"),
  hpue_all$estimates  |> mutate(quantity = "HPUE",           unit = "fish/hr")
) |>
  select(quantity, unit, estimate, se, ci_lower, ci_upper)

report_summary
```

```{r}
#| label: cedar-report-cv
report_table <- report_summary |>
  mutate(
    cv = fmt_pct(100 * se / estimate),
    estimate_fmt = if_else(
      unit %in% c("fish/hr"),
      fmt_rate(estimate),
      fmt_comma(estimate)
    ),
    ci_fmt = paste0(
      if_else(unit %in% c("fish/hr"), fmt_rate(ci_lower), fmt_comma(ci_lower)),
      "–",
      if_else(unit %in% c("fish/hr"), fmt_rate(ci_upper), fmt_comma(ci_upper))
    )
  ) |>
  select(Quantity = quantity, Unit = unit,
         Estimate = estimate_fmt, `95% CI` = ci_fmt, CV = cv)

knitr::kable(report_table, align = "llrrr",
             caption = "Cedar Lake 2023 creel survey season estimates.")
```

### Inline narrative

A manager's summary for Cedar Lake 2023 draws directly from the computed estimates:

::: {.callout-note title="Example: manager one-page summary (excerpt)"}
Cedar Lake anglers logged an estimated `r fmt_comma(eff_tot$estimates$estimate)` angler-hours during the 2023 open-water season (April 1–October 31), with weekday and weekend effort nearly equal. Total harvest was `r fmt_comma(harv_tot$estimates$estimate)` fish (95% CI: `r fmt_comma(harv_tot$estimates$ci_lower)`–`r fmt_comma(harv_tot$estimates$ci_upper)` fish). Anglers caught `r fmt_rate(cpue_all$estimates$estimate)` fish per angler-hour and retained `r fmt_rate(hpue_all$estimates$estimate)` fish per angler-hour, based on `r n_samp` sampled days and `r n_int` contacts (`r n_comp` complete trips).
:::

::: {.callout-note title="Example: technical report footnote"}
Effort uses ratio-of-means with day-type stratification from instantaneous count data. CPUE and HPUE use the mean-of-ratios estimator with all `r cpue_all$estimates$n` contacts, excluding 8 trips under 0.5 hours. Total catch uses only the `r n_comp` completed trips (`r fmt_pct(100 * n_comp / n_int)` of contacts); total harvest and release use the full interview set. Confidence intervals are 95% normal-approximation intervals from the design-based variance estimator. Domain: Cedar Lake shoreline, daylight hours, April 1–October 31, 2023.
:::

### Species-level CPUE

Cedar Lake targets three species with different catchability profiles. `estimate_catch_rate(by = species)` returns design-based estimates with stratification weights and confidence intervals for each:

```{r}
#| label: cedar-species-cpue
estimate_catch_rate(cedar_design, by = species)
```

Bluegill has the highest CPUE, consistent with shallow vegetated habitat where panfish are abundant. Channel catfish CPUE is lower but still drives a disproportionate share of harvest. White crappie CPUE is lowest overall, reflecting lower abundance and seasonal availability.

## Cedar Lake versus Harlan: a design contrast

Cedar Lake and Harlan share the same estimation engine, but the surveys differ in most operational details. The comparison below shows which pieces of the workflow stay fixed and which depend on the lake.

```{r}
#| label: cedar-harlan-compare
compare_tbl <- tibble::tribble(
  ~Feature,                   ~`Harlan Reservoir`,        ~`Cedar Lake`,
  "Survey period",            "Full calendar year",        "April–October",
  "Design type",              "Access-point",              "Roving circuit",
  "Shift hours",              "8 hr / 6 hr (P1/P2)",       "6 hr / 6 hr (equal)",
  "Angler composition",       "~70% boat",                 "~88% bank",
  "Interview completion",     "~65% complete",             "~29% complete",
  "Primary species",          "White bass, walleye",       "Bluegill, catfish, crappie",
  "Season effort",            "~33,000 angler-hours",
    sprintf("~%s angler-hours", fmt_comma(eff_tot$estimates$estimate)),
  "CPUE",                     "1.51 fish/hr",
    sprintf("%s fish/hr", fmt_rate(cpue_all$estimates$estimate)),
  "Effort CV",                "~12%",                      fmt_pct(cv_eff),
  "Harvest CV",               "~27%",                      fmt_pct(cv_harv)
)

knitr::kable(compare_tbl, align = "lll",
             caption = "Design and estimate comparison between Harlan Reservoir and Cedar Lake.")
```

The main difference is the incomplete-trip rate. Harlan's access-point design mostly captured complete trips; Cedar Lake's roving protocol leaves `r fmt_pct(100 * n_incomp / n_int)` of records incomplete. Total catch depends on completed trips (`r n_comp`), while harvest, release, and CPUE use the fuller interview record. For roving surveys, total catch is the binding precision constraint.

The roving design also produces a lower absolute effort estimate because Cedar Lake is smaller, has fewer anglers, and runs only seven months. CPUE is actually higher here (`r fmt_rate(cpue_all$estimates$estimate)` vs 1.51 fish/hr), so raw lake-to-lake comparisons are misleading without normalization.

## Takeaway {#sec-cedar-takeaway}

Cedar Lake shows how the same tidycreel pipeline adapts to a different field situation from the Harlan access-point design. Three choices drove the analysis:

**Roving contact → MOR estimator.** Mid-trip interviews make the mean-of-ratios estimator the right choice for catch rate.

**High incomplete-trip rate → asymmetric precision.** At `r fmt_pct(100 * n_incomp / n_int)` incomplete contacts, total-catch precision is tighter than total-harvest precision because only completed trips contribute. Better total-catch intervals require more sampled days or an exit-point component.

**Bank-dominant fishery → stratify by day type.** The main heterogeneity at Cedar Lake is temporal, not access-related, so the day-type strata capture the relevant variance structure.

The season estimates — `r fmt_comma(eff_tot$estimates$estimate)` angler-hours, `r fmt_comma(harv_tot$estimates$estimate)` fish harvested, `r fmt_rate(cpue_all$estimates$estimate)` fish per angler-hour — all fall within management-grade precision (CV < 20%). For a seven-month survey with two periods per sampled day and one roving clerk, that is a cost-effective characterization of the urban fishery.

## Reproducibility note

The Cedar Lake estimates in this chapter reproduce exactly given the same seed, data files, and package version:

```{r}
#| label: cedar-session
cat("tidycreel:", as.character(packageVersion("tidycreel")), "\n")
cat("R:         ", R.version$version.string, "\n")
cat("Data seed: 2023 (set in data/cedar_*.rds generation)\n")
```

Version-control the four `.rds` files alongside the analysis script. Saving the design object avoids rerunning the pipeline for downstream reporting:

```{r}
#| label: cedar-save-design
#| eval: false
saveRDS(cedar_design, here("data/cedar_design_2023.rds"))
```

Loading the saved design restores the full stratum structure, counts, and interview data in one call.
