---
title: "Pre-Survey Planning: Sample Size and Schedule Design"
---

```{r}
#| label: setup-ps
#| include: false
library(tidycreel)
library(tidyverse)
library(here)

harlan_schedule   <- readRDS(here("data/harlan_schedule.rds"))
harlan_counts     <- readRDS(here("data/harlan_counts.rds"))
harlan_interviews <- readRDS(here("data/harlan_interviews.rds"))
harlan_catch      <- readRDS(here("data/harlan_catch.rds"))

theme_creel <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      axis.line        = element_line(colour = "grey70"),
      axis.ticks       = element_line(colour = "grey70")
    )
}
```

A creel survey that finishes the season with a 45% CV on the effort estimate has not
merely underperformed — it has wasted field resources and produced information that
cannot support management decisions. The time to prevent that outcome is before the
first count sheet is printed, not after the data have been collected. Pre-survey
planning translates management objectives into concrete resource commitments: how
many sampling days, how many angler interviews, and when to schedule them.

This chapter covers the prospective side of survey design. Where Chapter 15
asks "how precise was my estimate?", this chapter asks "how many sampling units do
I need to achieve a target precision?" The two questions are mirrors of each other,
and `tidycreel` provides a dedicated suite of functions for both.

## The planning inputs {#sec-ps-inputs}

Three quantities are needed before any sample size calculation can be run.

**A precision target.** Agency standards differ, but a coefficient of variation
(CV) of 20% or less is widely cited as the threshold for a management-grade creel
estimate [@mccormick2017; @pollock1994]. A target CV of 15% is more demanding
but achievable on productive fisheries with moderate sampling effort. The CV target
must be stated before the survey begins; setting it retrospectively invites
motivated reasoning.

**A stratification frame (N_h).** The total number of eligible sampling days per
stratum — the frame — bounds how much precision is achievable regardless of effort.
For a day-type-stratified survey, N_h comes from the season calendar: count the
weekdays and weekend days in the planned season window.

**Pilot variance estimates (ybar_h, s2_h).** Sample size formulas require
between-day variance within each stratum. For a new fishery with no prior data,
the literature suggests starting with a budget of approximately 30 sampling days
and iterating [@mccormick2017]. For an established program, historical data from
previous seasons provide reliable variance estimates.

The Harlan reservoir survey has been running long enough that a completed season
can serve as the pilot for planning the next one. The following chunk extracts the
three planning inputs from the 2022 Harlan data.

```{r}
#| label: ps-pilot-stats
#| echo: true

# Shift lengths: period 1 = AM (8 h), period 2 = PM (6 h)
shift_hours <- c("1" = 8L, "2" = 6L)

daily_effort <- harlan_counts |>
  mutate(
    period       = as.character(period),
    angler_hours = (boat_anglers + bank_anglers) * shift_hours[period]
  ) |>
  group_by(date, day_type) |>
  summarise(daily_effort = sum(angler_hours), .groups = "drop")

pilot_stats <- daily_effort |>
  group_by(day_type) |>
  summarise(
    n_sampled = n(),
    ybar      = mean(daily_effort),    # mean effort per sampled day
    s2        = var(daily_effort),     # between-day variance
    .groups   = "drop"
  )

pilot_stats
```

```{r}
#| label: ps-frame
#| echo: true

# Season frame: eligible days per stratum
N_h <- harlan_schedule |>
  group_by(day_type) |>
  summarise(N = n(), .groups = "drop")

N_h
```

The Harlan 2022 season ran across 211 calendar days: 150 weekdays and 61
weekend/holiday days. Weekday effort is on average more than twice the weekend
mean (375 vs 162 angler-hours per sampled day), and between-day variance is
correspondingly larger in the weekday stratum.

These six values — N_h, ybar_h, and s2_h for each of the two strata — are
everything `creel_n_effort()` needs.

## How many sampling days? {#sec-ps-effort-n}

`creel_n_effort()` implements the stratified sample size formula from
@mccormick2017, based on Cochran ([-@cochran1977]) eq. 5.25. It returns the total
number of sampling days needed to achieve a target CV on the effort estimate, along
with the proportional allocation of those days to each stratum.

```{r}
#| label: ps-n-effort
#| echo: true

N_hv    <- setNames(N_h$N,          N_h$day_type)
ybar_hv <- setNames(pilot_stats$ybar, pilot_stats$day_type)
s2_hv   <- setNames(pilot_stats$s2,  pilot_stats$day_type)

creel_n_effort(cv_target = 0.20, N_h = N_hv, ybar_h = ybar_hv, s2_h = s2_hv)
```

Twenty-eight sampling days are sufficient to hit a 20% CV on effort for the Harlan
2022 design. Of those, 20 fall in the weekday stratum and 9 in the weekend stratum,
reflecting proportional allocation to season length. Tightening the target to 15%
requires 45 days:

```{r}
#| label: ps-n-effort-grid
#| echo: true

# Range of CV targets
cv_targets <- c(0.10, 0.15, 0.20, 0.25)

purrr::map(
  cv_targets,
  ~ creel_n_effort(.x, N_hv, ybar_hv, s2_hv)
) |>
  purrr::set_names(paste0("CV = ", cv_targets)) |>
  purrr::map_dfr(~ as.list(.x), .id = "cv_target") |>
  select(cv_target, weekday, weekend, total)
```

The relationship between CV target and required days follows a familiar pattern:
precision improves rapidly at first, then diminishing returns set in. Cutting the
CV from 25% to 20% costs only 9 additional sampling days; cutting it from 15% to
10% costs 33 more. This trade-off is most clearly seen as a curve.

```{r}
#| label: fig-ps-cv-tradeoff
#| fig-cap: "Required sampling days versus target effort CV. Diminishing returns are steep: each incremental gain in precision demands disproportionately more sampling days. Reference lines mark commonly used agency thresholds."
#| fig-alt: "Line chart with target CV on the x-axis and required sampling days on the y-axis showing a steep curve that flattens as CV increases."

tradeoff <- tibble(cv_target = seq(0.08, 0.35, by = 0.01)) |>
  mutate(
    result  = purrr::map(cv_target, ~ creel_n_effort(.x, N_hv, ybar_hv, s2_hv)),
    n_total = purrr::map_int(result, ~ .x["total"])
  )

thresholds <- tibble(
  cv_target = c(0.15, 0.20),
  label     = c("CV = 15%", "CV = 20%")
)

ggplot(tradeoff, aes(x = cv_target * 100, y = n_total)) +
  geom_line(colour = "#2171b5", linewidth = 1) +
  geom_vline(
    data    = thresholds,
    aes(xintercept = cv_target * 100),
    linetype = "dashed",
    colour   = "#cb181d",
    linewidth = 0.6
  ) +
  geom_label(
    data  = thresholds,
    aes(x = cv_target * 100 + 0.4, y = 130, label = label),
    colour = "#cb181d",
    size   = 3.2,
    hjust  = 0,
    label.size = 0
  ) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    x = "Target CV (effort)",
    y = "Sampling days required"
  ) +
  theme_creel()
```

An important caveat: these sample sizes are minimums under the assumption that
observations are distributed according to the pilot variances. Real variances
fluctuate between years. A prudent practice is to add a 10–15% buffer to the
computed minimum or to re-run the calculation mid-season if conditions differ
substantially from the pilot year.

## How many interviews? {#sec-ps-cpue-n}

Effort precision is driven by the number of sampling days. CPUE precision is driven
by the number of angler interviews. The two sample sizes are independent because
they estimate different things: effort uses count data collected by surveyors, while
CPUE uses per-party catch and effort reported by anglers.

`creel_n_cpue()` computes the number of interviews required to achieve a target CV
on the CPUE ratio estimator [@cochran1977, Ch. 6]. It requires pilot CVs for catch
and effort per interview, plus the correlation between them.

```{r}
#| label: ps-cpue-pilot
#| echo: true

catch_by_interview <- harlan_catch |>
  group_by(interview_id) |>
  summarise(n_fish = sum(n_fish), .groups = "drop")

interviews_pilot <- harlan_interviews |>
  left_join(catch_by_interview, by = "interview_id") |>
  mutate(n_fish = if_else(is.na(n_fish), 0L, as.integer(n_fish)))

cv_catch  <- sd(interviews_pilot$n_fish)    / mean(interviews_pilot$n_fish)
cv_effort <- sd(interviews_pilot$hours_fished) / mean(interviews_pilot$hours_fished)
rho       <- cor(interviews_pilot$n_fish, interviews_pilot$hours_fished)

tibble(
  statistic = c("CV of catch per interview", "CV of effort per interview", "Correlation (rho)"),
  value     = round(c(cv_catch, cv_effort, rho), 3)
)
```

Catch is highly variable relative to its mean (CV ≈ 1.6), which is typical of
recreational fisheries: most parties catch nothing while a few catch many fish.
Effort (hours fished per party) is less dispersed (CV ≈ 0.7). The positive
correlation between catch and effort (ρ ≈ 0.35) means that parties who fish longer
also tend to catch more — a signal of behavioral variation across angler types.

```{r}
#| label: ps-n-cpue
#| echo: true

# Conservative estimate (rho = 0 over-estimates n when catch and effort covary)
creel_n_cpue(
  cv_catch  = cv_catch,
  cv_effort = cv_effort,
  rho       = 0,
  cv_target = 0.20
)

# Less conservative: use observed correlation
creel_n_cpue(
  cv_catch  = cv_catch,
  cv_effort = cv_effort,
  rho       = rho,
  cv_target = 0.20
)
```

Setting `rho = 0` is conservative: it over-estimates the required sample size
because it ignores the variance-reducing effect of positive catch–effort covariance.
The observed ρ = 0.35 reduces the 20%-CV interview requirement from 78 to 59. Using
the observed correlation is appropriate when pilot data are reliable; in the absence
of pilot data, the conservative `rho = 0` default is the safer choice.

A practical minimum of 100 interviews is also widely recommended. Below that
threshold, the ratio estimator's confidence interval coverage may fall short of its
nominal 95% even if the formula suggests fewer interviews are needed, because the
distribution of catch per angler is highly right-skewed [@pollock1994].

## Working backwards from a fixed budget {#sec-ps-cv-from-n}

Planning calculations usually run in the forward direction: "I need CV = 20%; how
many days?" But managers sometimes allocate a fixed number of sampling days before
the precision question is asked, and the analyst must then characterize what that
budget can deliver.

`cv_from_n()` inverts the sample size formula: given a fixed number of sampling
units, what CV should you expect? Set `type = "cpue"` for interviews or
`type = "effort"` for sampling days.

```{r}
#| label: ps-cv-from-n
#| echo: true

# Director approves 60 interviews: what CPUE CV is achievable?
cv_from_n(
  type      = "cpue",
  n         = 60L,
  cv_catch  = cv_catch,
  cv_effort = cv_effort,
  rho       = 0   # conservative
)

cv_from_n(
  type      = "cpue",
  n         = 100L,
  cv_catch  = cv_catch,
  cv_effort = cv_effort,
  rho       = 0
)
```

Sixty interviews delivers an expected CPUE CV of about 23%, which may exceed the
agency's 20% threshold. One hundred interviews brings the CV to 18%, comfortably
within standard precision targets. If neither budget is achievable, this output
makes the trade-off explicit and auditable — a more defensible position than an
after-the-fact rationalization of inadequate precision.

## Allocating days across strata {#sec-ps-allocation}

Once the total number of sampling days is determined, they must be distributed
across strata. Two allocation strategies dominate creel survey practice:
**proportional allocation** assigns days in proportion to each stratum's share of
the season calendar; **Neyman (optimal) allocation** directs more days to strata
with higher between-day variance, minimizing total variance for a fixed sample
budget [@cochran1977; @pollock1994].

`reallocate_strata()` computes Neyman-optimal allocation given a fixed day budget
and stratum-level pilot variances.

```{r}
#| label: ps-allocation
#| echo: true

n_budget <- 45L   # total sampling days available

# Proportional allocation
prop_alloc <- round(n_budget * N_hv / sum(N_hv)) |>
  as.integer() |>
  setNames(names(N_hv))

# Neyman-optimal allocation
neyman_alloc <- reallocate_strata(n_total = n_budget, N_h = N_hv, s2_h = s2_hv)

tibble(
  stratum  = names(N_hv),
  N_h      = as.integer(N_hv),
  prop     = prop_alloc,
  neyman   = neyman_alloc[names(N_hv)]
)
```

For the Harlan design, Neyman allocation adds 5 days to the weekday stratum and
removes 4 from the weekend stratum relative to proportional (32→37 weekday,
13→9 weekend). The weekday stratum has both more calendar days and substantially
higher between-day variance (161,734 vs 54,123), so directing additional sampling
effort there reduces the overall variance of the effort estimate.

```{r}
#| label: fig-ps-allocation
#| fig-cap: "Proportional versus Neyman allocation of 45 sampling days across day-type strata. Neyman allocation concentrates effort in the weekday stratum, which has both more eligible days and higher between-day variance."
#| fig-alt: "Grouped bar chart comparing proportional and Neyman allocation of sampling days for weekday and weekend strata, with Neyman bars taller for weekdays and shorter for weekends."

alloc_df <- tibble(
  stratum    = rep(c("Weekday", "Weekend"), 2),
  method     = rep(c("Proportional", "Neyman"), each = 2),
  n_days     = c(prop_alloc["weekday"], prop_alloc["weekend"],
                 neyman_alloc["weekday"], neyman_alloc["weekend"])
) |>
  mutate(method = factor(method, levels = c("Proportional", "Neyman")))

ggplot(alloc_df, aes(x = stratum, y = n_days, fill = method)) +
  geom_col(position = position_dodge(width = 0.65), width = 0.55) +
  geom_text(
    aes(label = n_days),
    position = position_dodge(width = 0.65),
    vjust    = -0.5,
    size     = 3.5,
    fontface = "bold"
  ) +
  scale_fill_manual(values = c("Proportional" = "#9ecae1", "Neyman" = "#2171b5")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Sampling days", fill = "Allocation") +
  theme_creel() +
  theme(legend.position = "top")
```

Neyman allocation is most beneficial when strata differ substantially in variance.
When strata have similar within-stratum variability, proportional allocation
performs nearly as well and has the practical advantage of simplicity: interviewers
can be scheduled without knowledge of historical variance estimates, and the
allocation is easier to explain to field crews and managers.

`optimal_n()` combines the two steps — total sample size and Neyman allocation —
into a single call. Rather than first computing `creel_n_effort()` and then calling
`reallocate_strata()`, `optimal_n()` takes the CV target directly and returns both
the required total days and the per-stratum allocation:

```{r}
#| label: ps-optimal-n
#| echo: true

optimal_alloc <- optimal_n(
  cv_target = 0.20,
  N_h       = N_hv,
  ybar_h    = ybar_hv,
  s2_h      = s2_hv
)
optimal_alloc
```

The result shows total days required under Neyman allocation to hit the target CV,
alongside the per-stratum breakdown. For the Harlan design, `optimal_n()` returns
the same total (28) as `creel_n_effort()` at this CV target, but allocates
differently: 23 weekday / 6 weekend (Neyman) versus 20 weekday / 9 weekend
(proportional). Neyman allocation concentrates days in the weekday stratum because
that stratum has both more calendar days and higher between-day variance. When
strata variances differ substantially across surveys, Neyman allocation can require
fewer total days than proportional for the same precision target; for the Harlan
design the totals happen to be equal, and the practical benefit is the more efficient
per-stratum split.

## Power to detect a change {#sec-ps-power}

CV targets address precision, not the capacity to detect change. A survey that
delivers a 20% CV on a seasonal CPUE estimate can still fail to detect a
biologically important regulation effect if individual catch rates are highly
variable. `creel_power()` calculates the probability of detecting a fractional
change in CPUE between two seasons given the number of interviews per season, the
historical CV of individual CPUE measurements, and a significance threshold.

The key input is `cv_historical`: the coefficient of variation of CPUE measured
from individual interviews (fish per hour per completed trip). This reflects how
heterogeneous anglers are — a fishery where most parties catch similar amounts has
a low CV; one where a few parties catch most of the fish has a high CV. For Harlan:

```{r}
#| label: ps-cv-historical
#| echo: true

# CV of CPUE per completed interview (fish per hour)
cpue_per_interview <- harlan_interviews |>
  left_join(
    harlan_catch |>
      group_by(interview_id) |>
      summarise(n_fish = sum(n_fish), .groups = "drop"),
    by = "interview_id"
  ) |>
  mutate(n_fish = if_else(is.na(n_fish), 0L, as.integer(n_fish))) |>
  filter(trip_status == "complete", hours_fished > 0) |>
  mutate(cpue = n_fish / hours_fished)

cv_historical <- sd(cpue_per_interview$cpue) / mean(cpue_per_interview$cpue)
round(cv_historical, 2)
```

Harlan's CV of `r round(cv_historical, 2)` is typical of multi-species recreational
fisheries where zero-catch parties are common: the distribution of individual catch
rates is strongly right-skewed, driving up the standard deviation relative to the
mean. High `cv_historical` means that detecting a moderate regulation effect
requires substantially more interviews than the precision target alone implies.

```{r}
#| label: ps-power-single
#| echo: true

# 390 interviews is the approximate Harlan annual sample
creel_power(
  n             = 390L,
  cv_historical = cv_historical,
  delta_pct     = 0.20,
  alpha         = 0.05,
  alternative   = "two.sided"
)
```

With 390 interviews per season — the approximate Harlan annual sample — power to
detect a 20% CPUE change is only about 18%. Even a 50% change is below 80% power
at that sample size:

```{r}
#| label: ps-power-large-effect
#| echo: true

creel_power(
  n             = 390L,
  cv_historical = cv_historical,
  delta_pct     = 0.50,
  alpha         = 0.05,
  alternative   = "two.sided"
)
```

This is not a failure of the Harlan survey design; it reflects the inherent
difficulty of detecting moderate changes in CPUE when individual catch rates are
heterogeneous. The power curve illustrates how detection threshold depends on both
sample size and individual-level variability.

```{r}
#| label: fig-ps-power-curve
#| fig-cap: "Statistical power to detect a fractional CPUE change as a function of interviews per season, for three effect sizes. Solid curves use the Harlan pilot CV (2.66); dashed curves show a hypothetical lower-variance fishery (CV = 0.80) as a contrast. Power is calculated using a two-sided normal approximation at α = 0.05."
#| fig-alt: "Power curves for two CV scenarios (high-variance and low-variance fisheries) across three effect sizes, showing that high individual catch-rate variability substantially limits power at feasible sample sizes."

power_grid <- expand.grid(
  n             = seq(50, 800, by = 25),
  delta_pct     = c(0.20, 0.30, 0.50),
  cv_hist       = c(cv_historical, 0.80)
) |>
  as_tibble() |>
  mutate(
    power = purrr::pmap_dbl(
      list(n, delta_pct, cv_hist),
      function(n_i, d_i, cv_i)
        creel_power(n_i, cv_historical = cv_i, delta_pct = d_i, alpha = 0.05)
    ),
    delta_label = paste0(delta_pct * 100, "% change"),
    fishery     = if_else(
      cv_hist > 1,
      paste0("High variability (CV = ", round(cv_hist, 2), ")"),
      paste0("Low variability (CV = ", round(cv_hist, 2), ")")
    )
  )

ltype_vals <- c("solid", "dashed") |>
  setNames(c(
    paste0("High variability (CV = ", round(cv_historical, 2), ")"),
    "Low variability (CV = 0.80)"
  ))

ggplot(power_grid, aes(
  x        = n,
  y        = power,
  colour   = delta_label,
  linetype = fishery
)) +
  geom_hline(yintercept = 0.80, colour = "grey60", linewidth = 0.5) +
  annotate("text", x = 810, y = 0.80, label = "80%", hjust = 0, size = 3, colour = "grey50") +
  geom_line(linewidth = 0.85) +
  scale_colour_manual(
    values = c("20% change" = "#cb181d", "30% change" = "#2171b5", "50% change" = "#238b45"),
    name   = "Effect size"
  ) +
  scale_linetype_manual(values = ltype_vals, name = "Fishery type") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_x_continuous(breaks = seq(0, 800, by = 200)) +
  labs(x = "Interviews per season", y = "Power") +
  theme_creel() +
  theme(legend.position = "right")
```

Two contrasts stand out. First, for the low-variance fishery (CV = 0.80, dashed),
80% power to detect a 30% change requires only about 90 interviews — a feasible
season target. For Harlan (CV = `r round(cv_historical, 2)`, solid), the same
target requires more than 1,200 interviews per season. Second, a 50% CPUE change
is detectable with 80% power at approximately 450 interviews for Harlan — a large
but not impossible sample if the survey priority justifies it.

The practical implication: power analysis for CPUE change detection should be
conducted early in the planning cycle, before interview targets are set, so that
resource decisions reflect the actual detection capacity of the survey. A survey
with adequate precision (CV ≤ 20%) may still have limited power to detect realistic
regulation effects on heterogeneous fisheries.

## Translating days to a calendar {#sec-ps-schedule}

Sample size outputs are unitless numbers: "sample 32 weekdays and 14 weekend days."
Turning those numbers into a field schedule requires randomly assigning specific
dates within each stratum across the season window. `generate_schedule()` does
this: it takes a date range and stratum-specific day counts and returns a
stratified random sample of survey dates.

```{r}
#| label: ps-generate-schedule
#| echo: true

planned_schedule <- generate_schedule(
  start_date = "2023-04-01",
  end_date   = "2023-10-31",
  n_periods  = 2,
  n_days     = c(weekday = 32L, weekend = 14L),
  include_all = FALSE,
  seed        = 2024
)

planned_schedule
```

The output is a `creel_schedule` object with one row per sampled day × period —
ready to pass directly to `creel_design()` when the survey begins. A quick
diagnostic check confirms that the selected days spread reasonably across the
season.

```{r}
#| label: fig-ps-schedule-dist
#| fig-cap: "Distribution of planned sampling days by month and day type from the generated 2023 Harlan schedule. Stratified random selection spreads days across the April–October season without heavy concentration in any single month."
#| fig-alt: "Stacked bar chart showing counts of sampled weekdays and weekend days in each month from April through October."

planned_schedule |>
  mutate(month = lubridate::month(date, label = TRUE, abbr = TRUE)) |>
  distinct(date, day_type, month) |>
  count(month, day_type) |>
  mutate(day_type = case_match(day_type, "weekday" ~ "Weekday", "weekend" ~ "Weekend")) |>
  ggplot(aes(x = month, y = n, fill = day_type)) +
  geom_col(position = "stack", width = 0.65) +
  scale_fill_manual(values = c("Weekday" = "#2171b5", "Weekend" = "#9ecae1")) +
  scale_y_continuous(breaks = 0:10) +
  labs(x = NULL, y = "Sampling days", fill = "Day type") +
  theme_creel() +
  theme(legend.position = "top")
```

The stratified random selection ensures that sampled days are spread across the
season rather than concentrated in any one month, which protects against
confounding seasonal variation with stratum effects.

If the season window shifts or some dates become unavailable (access closures,
equipment failure), `generate_schedule()` can be re-run with a new seed to
produce a replacement schedule. Each call produces a statistically valid design as
long as the selected dates remain a random sample within strata.

## A planning workflow for the 2023 Harlan season {#sec-ps-workflow}

The six functions form a decision sequence. Below is the complete planning
workflow for a hypothetical 2023 Harlan season, using 2022 as the pilot year.

**Step 1 — Set precision targets.**

```{r}
#| label: ps-workflow-targets
#| echo: true

targets <- tibble(
  estimand   = c("Effort",       "CPUE"),
  cv_target  = c(0.15,           0.20),
  rationale  = c("management-grade effort estimate", "ratio-estimator precision for CPUE")
)
targets
```

**Step 2 — Compute required sampling units.**

```{r}
#| label: ps-workflow-n
#| echo: true

n_effort <- creel_n_effort(
  cv_target = 0.15,
  N_h       = N_hv,
  ybar_h    = ybar_hv,
  s2_h      = s2_hv
)

n_cpue <- creel_n_cpue(
  cv_catch  = cv_catch,
  cv_effort = cv_effort,
  rho       = 0,
  cv_target = 0.20
)

tibble(
  resource          = c("Sampling days (effort)", "Angler interviews (CPUE)"),
  required          = c(n_effort["total"],         n_cpue),
  practical_minimum = c(30L,                       100L)
)
```

**Step 3 — Check against budget and revise if needed.**

The 45-day effort target and 80-interview CPUE target are both feasible for a
season-long Harlan survey. If the field budget supports only 30 sampling days,
`cv_from_n()` characterizes the precision achievable at that budget for both
effort and CPUE:

```{r}
#| label: ps-cv-from-n-effort
#| echo: true

# What effort CV is achievable with 30 sampling days?
cv_from_n(
  type   = "effort",
  n      = 30L,
  N_h    = N_hv,
  ybar_h = ybar_hv,
  s2_h   = s2_hv
)
```

Thirty days yields a higher effort CV than the 20% target, making the precision
trade-off explicit and auditable before the field season begins. The same call
with `type = "cpue"` and `n = 60` or `n = 100` was shown earlier in
@sec-ps-cv-from-n.

**Step 4 — Choose an allocation strategy.**

```{r}
#| label: ps-workflow-alloc
#| echo: true

final_alloc <- reallocate_strata(
  n_total = n_effort["total"],
  N_h     = N_hv,
  s2_h    = s2_hv
)

tibble(
  stratum = names(final_alloc),
  n_days  = as.integer(final_alloc)
)
```

**Step 5 — Generate the field calendar.**

```{r}
#| label: ps-workflow-schedule
#| echo: true

field_schedule <- generate_schedule(
  start_date = "2023-04-01",
  end_date   = "2023-10-31",
  n_periods  = 2,
  n_days     = as.integer(final_alloc) |> setNames(names(final_alloc)),
  include_all = FALSE,
  seed        = 99
)

# Summary counts by day_type and month
field_schedule |>
  mutate(month = lubridate::month(date, label = TRUE)) |>
  group_by(day_type, month) |>
  summarise(days_scheduled = n_distinct(date), .groups = "drop") |>
  tidyr::pivot_wider(names_from = day_type, values_from = days_scheduled, values_fill = 0L)
```

The calendar table shows how the 46 allocated sampling days distribute across strata and
months (Neyman integer rounding adds one day above the 45-day target). A roughly even spread across the April–October window is desirable; heavy
concentration in any one month would bias estimates toward conditions in that month.
If a gap appears — for instance, no weekend days scheduled in June — `generate_schedule()`
can be re-run with different seeds until a more balanced calendar is produced, or
`n_days` can be specified per month by using the `period_intensity` argument.

## What to do without pilot data {#sec-ps-no-pilot}

Not every new survey has a historical season to draw on. Three options exist in
order of preference:

1. **Conduct a short pilot study** — 10–15 sampling days spread across the season
   will produce rough variance estimates sufficient for planning purposes. The
   resulting sample sizes will be uncertain, but the direction of error is known:
   early-season variance often underestimates full-season variance in temperate
   fisheries.

2. **Borrow from a similar fishery** — If a nearby reservoir with comparable
   species composition and angler density has historical creel data, its
   within-stratum variance can serve as a proxy. Adjust upward if the target
   fishery is expected to be more heterogeneous.

3. **Use the literature defaults** — @mccormick2017 report that across 17 inland
   fisheries, a median of 16 sampling days achieved a 40% relative 95% CI for
   effort, and approximately 30 days is a reasonable starting point when variance
   is unknown. For CPUE, a minimum of 100 interviews provides adequate CI coverage
   even under highly skewed catch distributions [@pollock1994].

None of these options is as reliable as observed pilot data from the target
fishery. But planning under uncertainty is still better than no planning: a
30-day default budget forces an explicit precision conversation before the field
season begins, rather than a post-hoc reckoning with an inadequate estimate.

## `tidycreel` support {#sec-ps-tidycreel}

| Function | Purpose |
|---|---|
| `creel_n_effort()` | Sampling days for a target effort CV (proportional allocation) |
| `creel_n_cpue()` | Interviews for a target CPUE CV |
| `cv_from_n()` | Expected CV given a fixed number of sampling units; `type = "effort"` or `type = "cpue"` |
| `optimal_n()` | Sampling days and Neyman-optimal per-stratum allocation for a target effort CV |
| `reallocate_strata()` | Neyman-optimal allocation across strata given a fixed day budget |
| `creel_power()` | Power to detect a fractional CPUE change |
| `generate_schedule()` | Stratified random sampling calendar |

All functions that take stratum-level arguments expect named vectors. Names must
match across `N_h`, `ybar_h`, and `s2_h`; mismatched names will produce an error.
The output of `generate_schedule()` is a `creel_schedule` object that passes
directly to `creel_design()` without further transformation.

## Takeaway {#sec-ps-takeaway}

Pre-survey planning converts management objectives into field commitments.
`creel_n_effort()` answers the effort question; `creel_n_cpue()` answers the CPUE
question; `reallocate_strata()` distributes days optimally across strata;
`creel_power()` tests whether the planned effort is sufficient to detect
biologically important changes; and `generate_schedule()` turns the numerical
targets into a concrete field calendar.

The Harlan example illustrates a realistic planning sequence: 45 sampling days
under Neyman allocation (37 weekday, 9 weekend) will achieve approximately 15% CV
on effort, and 80–100 interviews will achieve approximately 20% CV on CPUE.
However, the power analysis reveals an important limitation: because individual
catch rates are highly variable (CV ≈ `r round(cv_historical, 2)`), detecting a
moderate CPUE change between seasons requires far more interviews than the precision
target implies. Precision and power are related but distinct objectives, and the
planning workflow must address both before the first field day is scheduled.
